# Introduction {#intro}

This book exists at the intersection between Geography and R (Figure \@ref(fig:venn)).
Suggested titles were *Geography with R* and *R for GIS*, each of which has advantages.
The former conveys the message that it's not just about spatial data: 
non-spatial attribute data are inevitably interwoven with geometry data, and Geography is about more than where something is on the map.
The latter option communicates that this is a book about using R as a *Geographic Information System* (GIS), to perform spatial operations on *geographic data* [@bivand_applied_2013].
However, the term GIS has some negative connotations <!-- I see what you mean by negative connotations but I had to think immediately of qualitative GIS and critical GIS, i.e., scientists from the humanities criticising GIS technologies. Think of the so-called "GIS crisis", positivism, make-believe objectivity, etc.--> (see Table \@ref(tab:gdsl)) and fails to communicate one of R's greatest strengths:
its unparalleled ability to seamlessly switch between geographic and non-geographic data processing, modeling and visualization tasks.

```{r venn, fig.cap="Venn diagram of the intersection between Geography and R.", echo=FALSE, message=FALSE}
library(sf) # load sf library
p = st_sfc(st_point(c(0, 1)), st_point(c(1, 1))) # create 2 points
b = st_buffer(p, dist = 1) # convert points to circles
i = st_intersection(b[1], b[2]) # find intersection between circles
plot(b) # plot circles
text(x = c(-0.5, 1.5), y = 1, labels = c("Geography", "R")) # add text
plot(i, col = "lightgrey", add = TRUE) # color intersecting area
```

The title was chosen because this book is about more than routine spatial data processing, something that is well captured by the term *geocomputation*.
What defines geocomputation as a development on previous work is the "creative and experimental use of GIS that it entails" [@longley_geocomputation:_1998].
With geocomputation geographic data can be used in new ways.

Another advantage of geocomputation is that it conveys the centrality of the concept of algorithms (which can be implemented in R functions) to advanced geographical research.
This is captured in another seminal book on the subject [@openshaw_geocomputation_2000]:

> GC is neither an equation-fest nor a computer gigaflop speed contest.
GeoComputation is about using the various different types of geodata and about
developing relevant geo-tools within the overall context of a 'scientific'
approach.

The computational approach to geographic data allows the creation of new tools from modular components.
R, being a functional language, makes the creation of the 'geo-tools' that Openshaw advocates relatively easy compared with the low level languages (e.g. FORTRAN and Java) that dominated the field at the time.

Geocomputation is a relatively
young field^[Geocomputation
has a ~30 year history dating back to the first [conference](http://www.geocomputation.org/) on the subject in 1996 (at the University of Leeds where the concept for this book was dreamt up) which was followed by a flurry of subsequent
publications.]
but methods published under the geocomputational banner have influenced the direction of geographical research, as we will see in subsequent chapters.
Algorithms are powerful tools that can become highly complex.
However, all algorithms are composed of smaller parts.
By teaching these foundations we aim to empower you to create your own solutions to geographic data problems. This can feel like breaking free from the metaphorical 'glass ceiling' imposed by GUI-based proprietary geographic information systems (see Table \@ref(tab:gdsl) for a definition of GUI).

<!-- todo: what contributions, which will we use, where in the book? -->

While embracing recent developments in the field, we also wanted to pay respects to the wider field of Geography, and its 2000 year history [@roller_eratosthenes_2010], of which geocomputation is a part.
Geography has played an important role in explaining and influencing humanity's relationship with the natural world and this book aims to be a part of the 'Geographic tradition'.
Geocomputation in this book therefore means more than simply analyzing geographic data on a computer.
It's about trying to make research which involves making geographic data more reproducible, scientific and socially beneficial.
This book is also part of the movement towards Geographic Data Science (GDS) which differs from GIS in several ways, some of which are outlined in Table \@ref(tab:gdsl).
Inspired by the emphasis of GDS research on reproducibility, this book aims to teach how to do geocomputation rather than just think about it.

```{r gdsl, echo=FALSE, message=FALSE}
d = readr::read_csv("extdata/gis-vs-gds-table.csv")
knitr::kable(x = d, caption = "Differences in emphasis between the fields of Geographic Information Systems (GIS) and Geographic Data Science (GDS).")
```

## Why Geocomputation with R?

In this book we treat R as a 'tool for the trade'.
Early geographers used a variety of tools including rulers, compasses and sextants to advance knowledge about the world. 
<!--nowadays part-->
It is important to remember that while R is a powerful tool, especially when interfaced with other software such as [GDAL](http://www.gdal.org/) and [PostGIS](http://postgis.net/), other tools may be better for certain tasks. <!-- What exactly do you mean? Which tasks? Is there a geocomputational task for which we do not have an interface? So maybe rephrase along the lines that R is not a GIS and therefore not suited for some geoprocessing tasks but that interfaces close this gap. -->

R is characterised by its flexibility, enabling geographical software developers to extend it in multiple ways.
A good example of this is support for generating interactive maps thanks to **leaflet** [@R-leaflet], which has been subsequently built-on by **tmap** and **mapview** [@R-mapview; @R-tmap].
These packages help overcome the criticism that R has "limited interactive [plotting] facilities" [@bivand_applied_2013].
This is no longer true, as illustrated by the code below which generates Figure \@ref(fig:interactive).

```{r, eval=FALSE, echo=FALSE}
a = osmdata::getbb("Hereford")
b = osmdata::getbb("Bialystok")
rowMeans(a)
rowMeans(b)
```

```{r interactive, fig.cap="World at night imagery from NASA overlaid by the authors' approximate home locations to illustrate interactive mapping with R."}
library(leaflet)
popup = c("Robin", "Jakub")
leaflet() %>%
  addProviderTiles("NASAGIBS.ViirsEarthAtNight2012") %>% 
  addAwesomeMarkers(lng = c(-3, 23), lat = c(52, 53), popup = popup)
```

It would be difficult to produce Figure \@ref(fig:interactive) with only 4 lines of code in another language, let alone embed the results in an interactive html page (the interactive version can be viewed at [robinlovelace.net/geocompr](http://robinlovelace.net/geocompr/intro.html)), illustrating R's flexibility. 

<!-- Other reasons for using R for geocompuation include: -->

R provides a stable language that should behave consistently across multiple computers, free of charge.
The use of R code therefore enables teaching geocomputation with reference to reproducible examples such as that provided in \@ref(fig:interactive) rather than abstract concepts.
But before cracking-on with the action, a few introductory remarks are needed to explain the approach taken here and provide context.

## R's spatial ecosystem

<!-- paragraphs (with references to chapters in the book): -->
<!-- 1. this book focus -> sf + raster/stars + leaflet/mapview (the recent state of spatial R); the history of R spatial is way longer -->
<!-- 2. R spatial story (from spatstat, by maptools, sp, raster, leaflet, to sf) -->
<!-- 3. short S story - an interface to Fortran and C -> nowodays R is an interface to many programming languages C, Fortran, C++, JavaScript, Go; R - an interface to libraries gdal (C/C++), proj (C), geos (C++) -->
<!-- 4. R also as an interface to GIS software GRASS GIS, SAGA, QGIS, ArcGIS (calls from R and from the software) + GIS javascript libraries leaflet; reference to the CLI/GUI integration chapter -->


<!-- 
- r, rstudio, gdal, proj4, geos, udunits 
- r packages sf, raster, etc.
- datasets 
-->

<!-- ## Introduction to GIS -->

<!-- 
- what's R
- what's GIS
- GIS data models (vector vs raster)
- coordinate reference system - CRS
- GIS data formats
- GDAL, GEOS, PROJ4
- GIS R package
- GIS beyond R 
-->

The most important recent evolution in R's spatial ecosystem has without doubt been support for simple features thanks to the **sf** package [@R-sf], described in Chapter \@ref(spatial-class).

