# Reading and writing spatial data {#read-write}

## Prerequisites {-}

- You must have run the code in Chapter \@ref(intro)

```{r, echo=FALSE, include=FALSE}
if(!exists("world"))
        source("code/01-introduction.R")
```

Spatial data comes in a wide variety of file formats, and **sf** is adept at handling them, via its interface to GDAL.
A major advantage of **sf** is that it is fast at geographical data I/O, as illustrated in the benchmark below:

```{r, results='hide'}
library(microbenchmark)
bench_read = microbenchmark(times = 5,
        st_read(f),
        rgdal::readOGR(f)
)
```

```{r}
bench_read$time[1] / bench_read$time[2]
```

The results demonstrate that **sf** can be much faster (*`r round(bench_read$time[1] / bench_read$time[2], 0)` times faster* in this case) than **rgdal** at reading-in the world countries shapefile.

The counterpart of `st_read()` is `st_write()`. This allows writing to a range of geographic vector file types, including the common formats `.geojson`, `.shp` and `.gpkg`. `st_read()` will decide which driver to use automatically, based on the file name, as illustrated in the benchmark below demonstrating write speeds for each format.

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

```{r, warning=FALSE}
system.time(st_write(world, "world.geojson", quiet = TRUE))
system.time(st_write(world, "world.shp", quiet = TRUE)) 
system.time(st_write(world, "world.gpkg", quiet = TRUE))
```

The full range of file-types supported by **sf** is reported by `st_drivers()`, the first 2 of which are shown below:

```{r}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```


<!-- ## Vector -->

<!-- 
- sf package 
- st_drivers
- st_read, read_sf
- st_write, write_sf
- text files 
- WKT, WKB, sp (st_as_sf)
-->

<!-- ## Raster -->

<!-- 
- raster package 
- raster
-->
