# Read, write, plot {#read-write}

## Prerequisites {-}

- You must have run the code in Chapter \@ref(intro)

```{r, echo=FALSE, include=FALSE}
if(!exists("world"))
        source("code/01-introduction.R")
```

## Introduction

Reading, writing and plotting are 3 of the most important initial stages in geocomputation.
If you cannot load your data, it's not even worth starting.
If you cannot write your processed data, it will make it hard to collaborate with others.
Furthermore, an often-overlooked issue is that if you cannot create visuals of your spatial data, it will be hard to understand.
For this reason this introductory chapter also has a section on basic map making.

Spatial data comes in a wide variety of file formats, and **sf** is adept at handling them, via its interface to GDAL.
A major advantage of **sf** is that it is fast at geographical data I/O, as illustrated in the benchmark below:

```{r, results='hide'}
library(microbenchmark)
bench_read = microbenchmark(times = 5,
        st_read(f),
        rgdal::readOGR(f)
)
```

```{r}
bench_read$time[1] / bench_read$time[2]
```

The results demonstrate that **sf** can be much faster (*`r round(bench_read$time[1] / bench_read$time[2], 0)` times faster* in this case) than **rgdal** at reading-in the world countries shapefile.

The counterpart of `st_read()` is `st_write()`. This allows writing to a range of geographic vector file types, including the common formats `.geojson`, `.shp` and `.gpkg`. `st_read()` will decide which driver to use automatically, based on the file name, as illustrated in the benchmark below demonstrating write speeds for each format.

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

```{r, warning=FALSE}
system.time(st_write(world, "world.geojson", quiet = TRUE))
system.time(st_write(world, "world.shp", quiet = TRUE)) 
system.time(st_write(world, "world.gpkg", quiet = TRUE))
```

The full range of file-types supported by **sf** is reported by `st_drivers()`, the first 2 of which are shown below:

```{r}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```

## Basic map making {#basic-map}

Basic maps in **sf** can be created quickly with the base `plot()` function. Unlike **sp**, however, **sf** by default creates a faceted plot, one sub-plot for each variable, as illustrated in the left-hand image in Figure \@ref(fig:sfplot). 

```{r sfplot, fig.cap="Plotting with sf, with multiple variables (left) and a single variable (right).", out.width="49%", fig.show='hold', warning=FALSE}
plot(world)
plot(world["population"])
```

As with **sp**, you can add layers to your maps created with `plot()`, with the argument `add = TRUE`^[In
fact, when you `plot()` an **sf** object, R is calling `sf:::plot.sf()` behind the scenes.
`plot()` is a generic method that behaves differently depending on the class of object being plotted.].
However, this only works if the initial plot has only 1 layer (result not shown):

```{r, fig.show='hide'}
plot(world["population"])
africa = world_continents[1, ]
plot(africa, add = TRUE, col = "red")
```

This can be very useful when quickly checking the geographic correspondence between two or more layers.
These plots work well for gaining a quick understanding of the data with few lines of code.
For more advanced map making we recommend using a dedicated visualisation package such as **tmap**, **ggplot2**, **mapview**, or **leaflet**.

<!-- 
- plot() function 
- map export 
-->

### Challenge

Using **sf**'s `plot()` command, create a map of Nigeria in context, like the one presented in figure \@ref(fig:nigeria). 

- Hint: this used the `lwd`, `main` and `col` arguments of `plot()`. 
- Bonus: make the country boundaries a dotted grey line.
- Hint: `border` is an additional argument of `plot()` for **sf** objects.

```{r nigeria, warning=FALSE, echo=FALSE, fig.cap="Map of Nigeria in context illustrating sf's plotting capabilities"}
nigeria = filter(world, name_long == "Nigeria")
bb_africa = st_bbox(africa)
plot(africa[2], col = "white", lwd = 3, main = "Nigeria in context", border = "lightgrey")
# plot(world, lty = 3, add = TRUE, border = "grey")
plot(world, add = TRUE, border = "grey")
plot(nigeria, col = "yellow", add = TRUE, border = "darkgrey")
ncentre = st_centroid(nigeria)
ncentre_num = st_coordinates(ncentre)
text(x = ncentre_num[1], y = ncentre_num[2], labels = "Nigeria")
```

### Further work

**sf** makes R data objects more closely alligned to the data model used in GDAL and GEOS, in theory making spatial data operations faster.
The work here provides a taster of the way that **sf** operates but there is much more to learn.
There is a wealth of information that is available in the package's vignettes: these are highly recommended.

As a final exercise, we'll see how to do a spatial overlay in **sf** by first converting the countries of the world into centroids and then subsetting those in Africa:

```{r, out.width="50%", fig.cap="Centroids in Africa"}
world_centroids = st_centroid(world)
plot(world_centroids[1])
africa_centroids = world_centroids[africa,]
plot(africa_centroids, add = TRUE, cex = 2)
```

Note: another way of acheiving the same result is with a GEOS function for identifying spatial overlay:

```{r}
sel_africa = st_covered_by(world_centroids, africa, sparse = FALSE)
summary(sel_africa)
```

This shows that there are 56 countries in Africa.
We can check if they are the same countries as follows:

```{r}
africa_centroids2 = world_centroids[sel_africa,]
identical(africa_centroids, africa_centroids2)
```

### Exercises

- Perform the same operations and map making for another continent of your choice.
- Bonus: Download some global geographic data and add attribute variables assigning them to the continents of the world.


<!-- ## Vector -->

<!-- 
- sf package 
- st_drivers
- st_read, read_sf
- st_write, write_sf
- text files 
- WKT, WKB, sp (st_as_sf)
-->

<!-- ## Raster -->

<!-- 
- raster package 
- raster
-->
