# Attribute data operations {#attr}

## Prerequisites {-}

- This chapter requires **tidyverse** and **sf**:

```{r, message=FALSE}
library(sf)
library(tidyverse)
```

- You must have loaded the `world` and `worldbank_df` data from the **spData** package:

```{r, results='hide'}
library(spData)
data("world")
data("worldbank_df")
```

## Introduction

Attribute data is non-geographic information associated with geographical data.
In the context of simple features, introduced in the previous chapter, this means a tabular data joined onto the `geometry` variables of `sf` objects.
This structure enables multiple columns to represent a range of attributes for thousands of features (one row per feature).

There is a strong overlap between geographical and non-geographical operations:
non-spatial subset, aggregate and join operations each have their geographical equivalents.
For this reason this chapter provides the foundation for next (chapter \@ref(spatial-data-operations)).
The two chapters share the same structure and input data; it is recommended that they are read together.

The non-spatial versions of these methods are common and easy to understand with R, so they are covered first.
The methods are largely cross-transferable to the trickier tasks of spatial data operations, so pay attention!

Simple features are represented as objects, such as `world`, with class `sf` in the **sf** package:

```{r}
class(world)
```

The output shows that `sf` objects have two classes (`sf` and `data.frame`), meaning that are essentially data frames.
Using data frames, the basic class used for data analysis in R, particularly within [the **tidyverse** package ecosystem](http://tidyverse.org/), has many advantages when it comes to attribute data operations.
It means that all the accumulated know-how in the R community for handling data frames to be applied to geographic data which contain attributes.

This 'world' dataset contains 10 non-geographical variables (and one geometry column) with data for almost 200 countries, as can be ascertained using base functions for working with tabular data:

```{r}
dim(world) # it is a 2 dimensional object, with rows and columns
nrow(world) # how many rows?
ncol(world) # how many columns?
```

Extracting the attribute data of an `sf` object is the same as removing the geometry column:

```{r}
world_df = world
st_geometry(world_df) = NULL
class(world_df)
```

This can be useful if the geometry column causes problem, e.g. by occupying large amounts of RAM.
However, for most cases there is no harm in keeping the geometry column, as data frame operations on `sf` will only act on the attribute data.
For this reason, being good at working with attribute data in geographic data is the same being proficient at handling data frames in R.
For many applications, the most effective and intuitive way to work with data frames is with the **dplyr** package.

## Base vs data.table vs dplyr

Simple feature objects of class `sf` behave exactly the same as `data.frame` objects for most base R operations.
Unlike objects of class `Spatial` defined by the **sp** package, `sf` objects are also compatible with **dplyr** and **data.table** packages.
This is an advantage because they provide fast functions for data manipulation.

Which method you use is largely a matter of preference.
In this chapter the focus is largely on **dplyr** because of it's intuitive function names and its ability to perform multiple chained operations using the pipe operator.
The important thing is that you select a data processing paradigm of choice, and master it.

## Attribute subsetting

```{r, eval=FALSE}
world[1:6, ] # subset rows
```

```{r, eval=FALSE}
world[, 1:3] # subset columns
```

After each operation, the geometry column is preserved.

**dplyr** makes working with data frames easier and is compatible with `sf` objects, after the package has been loaded:

The `select()` function, for example, can be used to both subset and renames columns in a single line, for example:

```{r}
world1 = select(world, name_long, continent, population = pop)
head(world1, n = 2)
```

This is more concise than the base R equivalent (which saves the result as an object called `world2` to avoid overriding the `world` dataset created previously):

```{r, eval=FALSE}
world2 = world[c("name_long", "continent", "pop")] # subset columns by name
names(world2)[3] = "population" # rename column manually
```

The *pipe* operator (` %>% `), which passes the output of one function into the first argument of the next function, is commonly used in **dplyr** data analysis workflows.
This works because the fundamental **dplyr** functions (or 'verbs', like `select()`) all take a data frame object in and spit a data frame object out.
Combining many functions together with pipes is called *chaining* or *piping*.
The advantage over base R for complex data processing operations is that this approach prevents nested functions and is easy to read because there is a clear order and modularity to the work (a piped command can be commented out, for example).

The example below shows yet another way of creating the renamed `world` dataset, using the pipe operator:

```{r}
world3 = world %>%
  select(name_long, continent)
```

Note that this can also be written without the pipe operator because, in the above code, the `world` object is simply 'piped' into the first argument of `select()`.
The equivalent **dplyr** code without the pipe operator is:

```{r}
world4 = select(world, name_long, continent)
```


The pipe operator can be used for many data processing tasks with attribute data:

<!-- todo - describe these: ==, !=, >, >=, <, <=, &, | -->


```{r}
# Filtering attribute data with dplyr
world %>%
  filter(pop > 1e9)
```

This is equivalent to the following base R code (not run to preserve the NAs):^[[Note](https://github.com/Robinlovelace/geocompr/issues/28) NAs do not work for subsetting by inequalities in base R, hence conversion of NAs to 0s in this version)]

```{r, eval=FALSE}
# subsetting simple feature rows by values
world$pop[is.na(world$pop)] = 0 # set NAs to 0
world_few_rows = world[world$pop > 1e9,]
```

### Exercises

## Attribute data aggregation 

```{r, results='hide'}
# data summary (not shown)
summary(world)

# data summary by groups (not shown)
world_continents = world %>% 
  group_by(continent) %>% 
  summarise(pop = sum(pop, na.rm = TRUE), country_n = n())
world_continents
```

```{r}
# sort variables
## by name
world_continents %>% 
  arrange(continent)
## by population (in descending order)
world_continents %>% 
  arrange(-pop)
```

### Exercises

## Attribute data joining 

<!-- https://github.com/dgrtwo/fuzzyjoin -->
<!-- http://r4ds.had.co.nz/relational-data.html -->
<!-- non-unique keys -->

Combining data from different sources is one of the most common task in data preparation. 
It could be done using joins - methods created to work with a pair of tables.
The **dplyr** package has a set of verbs to easily connect `data.frames` - `left_join()`, `right_join()`,  `inner_join()`, `full_join`, `semi_join()` and `anti_join()`. 
They are thoroughly explained in the Relational data chapter in the book R for Data Science [@grolemund_r_2016].

Working with spatial data, however, usually involves a connection between spatial data (`sf` objects) and tables (`data.frame` objects).
Fortunately, the **sf** package has all of the **dplyr** join functions adapted to work with `sf` objects.
The only important difference between combining two `data.frames` and combining `sf` with `data.frame` is a `geom` column.
Therefore, the result of data joins could be either an `sf` or `data.frame` object.

The easiest way to understand the concept of joins is to use a smaller datasets. 
We will use an `sf` object `north_america` with country codes (`iso_a2`), names and geometries, as well as `data.frame` object `wb_north_america` containing information about urban population and unemployment for three countries. It is important to add that the first object has data about Canada, Greenland and United States and the second one has data about Canada, Mexico and United States:

```{r}
north_america = world %>%
  filter(subregion == "Northern America") %>%
  select(iso_a2, name_long)
north_america
```

```{r}
plot(north_america[0])
```

```{r}
wb_north_america = worldbank_df %>% 
  filter(name %in% c("Canada", "Mexico", "United States")) %>%
  select(name, iso_a2, urban_pop, unemploy = unemployment)

wb_north_america
```

In this book, we focus on spatial data. 
All of the following examples will have a `sf` object as the first argument and a `data.frame` object as the second argument. 
A new `sf` object will be a result of these joins. 
However, the reverse order is also possible and will result in a `data.frame` object.
This is beyond the scope of this book, but we encourage you to try it.

### Left joins

Left join is the most often used type of joins.
The `left_join()` returns all observations from the left object (`north_america`) and the matched observations from the right object (`wb_north_america`). 
In cases, like `Greenland`, when we don't have a data in the right object, `NA` values will be introduced.

To connect two object we need to specify a key.
This is a variable (or variables) that uniquely identifies each observation (row). 
The argument `by` is used to state which variable is the key. 
In simple cases, a single, unique variable exist in both objects, for example `iso_a2` column:

```{r}
left_join1 = north_america %>% 
  left_join(wb_north_america, by = "iso_a2")
left_join1
```

It is also possible to join objects by different variables.
Both of the datasets have variables with names of countries, but they are named differently.
The `north_america` has a `name_long` column and the `wb_north_america` has a `name` column.
In these cases, we can use a named vector to specify the connection, e.g. `c("name_long" = "name")`:

```{r}
left_join2 = north_america %>% 
  left_join(wb_north_america, by = c("name_long" = "name"))
left_join2
```

The new object `left_join2`, however is still not perfectly connected as it has two duplicated variables - `iso_a2.x` and `iso_a2.y`.
To solve this problem we should specify all the keys:

```{r}
left_join3 = north_america %>% 
  left_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
left_join3
```

<!-- ```{r} -->
<!-- # error: keeps geom col -->
<!-- left_join4 = wb_north_america %>%  -->
<!--   left_join(north_america, by = c("iso_a2")) -->
<!-- left_join4 -->
<!-- ``` -->

### Right joins

The `right_join()` keeps all observations from the right object (`wb_north_america`), but in the same time preserves a `sf` class from the left object (`north_america`).

```{r}
right_join1 = north_america %>% 
  right_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
right_join1
```

You can see that the new object `right_join1` has information about Mexico, but drop information about Greenland.
What's more, our right object, as a `data.frame`, doesn't have a geometry representation of Mexico.
As a result, the `right_join1` object contains only non-spatial data of Mexico.
It could be easily illustrated using the `plot` function:

```{r}
plot(right_join1[0]) # Canada and United States only
```

<!-- ```{r} -->
<!-- # error: unwanted geom column added -->
<!-- right_join2 = wb_north_america %>%  -->
<!--   right_join(north_america, by = "iso_a2") #%>% plot() -->
<!-- right_join2 -->
<!-- ``` -->

### Inner joins

The `inner_join()` keeps only observations from the left object (`north_america`) where there are matching observations in the right object (`wb_north_america`). 
Additionally, all columns from the left and right object are kept:

```{r}
inner_join1 = north_america %>% 
  inner_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
inner_join1
```

<!-- ```{r} -->
<!-- # error: geom column added -->
<!-- inner_join2 =  wb_north_america %>%  -->
<!--   inner_join(north_america, by = "iso_a2")  -->
<!-- inner_join2 -->
<!-- ``` -->

### Semi joins

The `semi_join()` is very similar to the `inner_join()`. 
It also keeps only observations from the left object (`north_america`) where there are matching observations in the right object, but keeping just columns from the left one:
<!-- filtering? -->

```{r}
semi_join1 = north_america %>% 
  semi_join(wb_north_america, by = "iso_a2")
semi_join1
```

<!-- ```{r} -->
<!-- semi_join2 = wb_north_america %>% -->
<!--   semi_join(north_america, by = "iso_a2") -->
<!-- semi_join2 -->
<!-- ``` -->

### Anti joins

The `anti_join()` returns all rows from the left object that are not matching observations in the right object.
Only columns from the right object are kept:

```{r}
anti_join1 = north_america %>% 
  anti_join(wb_north_america, by = "iso_a2")
anti_join1
```

```{r}
plot(anti_join1[0])
```

<!-- ```{r} -->
<!-- anti_join2 = wb_north_america %>%  -->
<!--   anti_join(north_america, by = "iso_a2") -->
<!-- anti_join2 -->
<!-- ``` -->

### Full joins

The `full_join()` returns all rows and all columns from both the left and right object. 
It also puts `NA` in cases where there are not matching values and returns an empty geometry for cases that only exist in the right object:

```{r}
full_join1 = north_america %>% 
  full_join(wb_north_america, by = "iso_a2")
full_join1
```

<!-- ```{r} -->
<!-- # error: null geom -->
<!-- full_join2 = wb_north_america %>%  -->
<!--   full_join(north_america, by = "iso_a2") #%>% plot() -->
<!-- full_join2 -->
<!-- ``` -->

### Exercises

<!-- -->

## Attribute data creation


### Exercises

## Removing spatial information

Most of the function from **sf** package do not drop a `geometry` column. To extract a data frame `st_geometry()` or `st_set_geometry()` function can be used.

```{r}
world_st = world
st_geometry(world_st) = NULL
class(world_st)

# OR

world_st2 = world
world_st2 = world_st2 %>% st_set_geometry(NULL)
class(world_st2)
```


<!-- 
- dplyr, tidyr, and purrr packages
- lubridate??
- pipes
-->
