# Working with attribute table {#attr}

To be proficient at working with attribute data in **sf** means being proficient with handling data frames in R. This is a broad topic beyond the scope of this tutorial.
The subsequent examples demonstrate how attribute data in **sf** classes can be manipulated as data frames, and how to add/remove geometry data.

```{r, eval=FALSE}
# subsetting simple feature rows
w[1:3, ]
```

```{r}
# subsetting simple feature columns
selc = 1:3
w_few_cols = w[selc]
head(w_few_cols)
```

```{r}
library(dplyr)
```

```{r}
# subsetting simple feature columns by name
w_few_cols2 = w[c('name', 'continent')]

#OR

w_few_cols2 = w %>%
        select(., name, continent)

head(w_few_cols2)
```

```{r}
# ==, !=, >, >=, <, <=, &, |

# subsetting simple feature rows by values
w_few_rows = w[w$pop_est>1000000000, ]

#OR
w_few_rows = w %>% 
        filter(., pop_est>1000000000)

head(w_few_rows)
```

```{r}
# add a new column
w$area = raster::area(as(w, "Spatial")) / 1000000 #it there any function for area calculation on sf object?
w$pop_density = w$pop_est / w$area

# OR

w = w %>% 
        mutate(area=raster::area(as(., "Spatial")) / 1000000) %>% 
        mutate(pop_density=pop_est/area)
```

```{r}
# data summary
summary(w)

# data summary by groups
w_continents = w %>% 
        group_by(continent) %>% 
        summarise(continent_pop=sum(pop_est), country_n=n())
w_continents
```

```{r}
# sort variables
## by name
w_continents %>% 
        arrange(continent)
## by population (in descending order)
w_continents %>% 
        arrange(-continent_pop)
```

Most of the function from **sf** package do not drop a `geometry` column. `st_geometry()` or `st_set_geometry()` can be used to extract a data frame.

```{r}
w_st = w
st_geometry(w_st) = NULL
class(w_st)

# OR

w_st2 = w
w_st2 = w_st2 %>% st_set_geometry(., NULL)
class(w_st2)
```


<!-- 
- dplyr, tidyr, and purrr packages
- lubridate??
- pipes
-->

<!-- 
- view, add new rows/columns, subset, select, summarize 
-->
