# Attribute data operations {#attr}

## Prerequisites {-}

- This chapter requires **tidyverse** and **sf**:

```{r, message=FALSE}
library(sf)
library(tidyverse)
```

- You must have loaded the `world` and `worldbank_df` data from the **spData** package:

```{r, results='hide'}
library(spData)
data("world")
data("worldbank_df")
```

## Introduction

Attribute data is non-spatial information associated with geographic data.
In the context of simple features, introduced in the previous chapter, this means a data frame with a column for each variable and one row per geographic feature stored in the `geom` list-column of `sf` objects.
This structure enables multiple columns to represent a range of attributes for thousands of features (one row per feature).

There is a strong overlap between geographical and non-geographical operations:
non-spatial subset, aggregate and join each have their geographical equivalents.
This chapter therefore provides the foundation for next (chapter \@ref(spatial-data-operations)) in terms of structure and input data.
The non-spatial versions of these methods are widely used and (using functions such as `filter()` and `[`) cross-transferable to the trickier tasks of spatial data operations.

Simple features are represented with class `sf`, defined in the **sf** package.
This is shown in the result of the function `class()` applied to the `world` object:
Thus the object behaves in the same way as a `data.frame`, but it contains a special column called `geom`.

```{r}
class(world)
```

The output shows that `sf` objects actually have two classes, `sf` and `data.frame`. 
This duality is central to the concept of simple features:
most of the time `sf` can be treated as and behave like `data.frame`s.
Simple features are, in essence, data frames with a spatial extension.

The trusty `data.frame` (and extensions to it such as the `tibble` class used in the tidyverse) is a workhorse for data analysis in R.
Extending this system to work with spatial data has many advantages,
meaning that all the accumulated know-how in the R community for handling data frames to be applied to geographic data which contain attributes.

Before proceeding to perform various attribute operations of a dataset, it is worth taking time to think about its basic parameters.
In this case, the  `world` object contains 10 non-geographical columns (and one geometry list-column) with data for almost 200 countries.
This can be be checked using base R functions for working with tabular data such as `nrow()` and `ncol()`:

```{r}
dim(world) # it is a 2 dimensional object, with rows and columns
nrow(world) # how many rows?
ncol(world) # how many columns?
```

Extracting the attribute data of an `sf` object is the same as removing the geometry column:

```{r}
world_df = st_set_geometry(world, NULL)
class(world_df)
```

This can be useful if the geometry column causes problems, e.g. by occupying large amounts of RAM, or when you want to save the non-spatial data.
For most cases, however, there is no harm in keeping the geometry column because non-spatial data operations on `sf` objects act only on the attribute data.
For this reason, being good at working with attribute data in geographic data is the same being proficient at handling data frames in R.
For many applications, the most effective and intuitive way of working with data frames is with the **dplyr** package, as we will see in the next
section.^[
Unlike objects of class `Spatial` defined by the **sp** package, `sf` objects are also compatible with **dplyr** and **data.table** packages, which provide fast and powerful functions for data manipulation (see [Section 6.7](https://csgillespie.github.io/efficientR/data-carpentry.html#data-processing-with-data.table) of @gillespie_efficient_2016).
This chapter focusses on **dplyr** because of its intuitive function names and ability to perform multiple chained operations using the pipe operator.]

## Attribute subsetting

<!-- info about pull (dplyr 0.6): -->
<!-- https://github.com/tidyverse/dplyr/commit/0b9aabf6c06c9cd3b784b155044d497d4b93df3e -->
Because simple feature objects are also data frames, you can use a wide range of functions (from base R and packages) for subsetting them, based on attribute data.
Base R subsetting functions include `[`, `subset()` and  `$`.
**dplyr** subsetting functions include `select()`, `filter()`, and `pull()`.
Most of the functions preserve the geometry column.

You can use the `[` operator to subset rows and columns. 
Two arguments are required, one for rows (observations) and one for columns (variables), e.g. [rows, columns].
These arguments could be either numeric, which indicates a position, or character, which indicates a name of row or column.

```{r, eval=FALSE}
world[1:6, ] # subset rows by position
```

```{r, eval=FALSE}
world[, 1:3] # subset columns by position
```

```{r, eval=FALSE}
world[, c("name_long", "lifeExp")] # subset columns by name
```

You can also subset a `sf` object based on a given condition using (surprise, surprise...) the `subset()` function:

```{r, eval=FALSE}
small_countries = subset(world, area_km2 < 10000)
```

The `$` operator retrieves a variable by its name and returns a vector:

```{r, eval=FALSE}
world$name_long
```

<!-- , after the package has been loaded: [or - it is a part of tidyverse] -->
**dplyr** makes working with data frames easier and is compatible with `sf` objects.
The main **dplyr** functions that help with attribute subsetting are `select()`, `filter()` and `pull()`.

The `select()` function picks columns by its name.
For example, you could select only two columns - `name_long` and `pop`:

```{r}
world1 = select(world, name_long, pop)
head(world1, n = 2)
```

This function gives a possibility to select a range of columns with an `:` operator or exclude some columns with an `-` operator:

```{r, eval=FALSE}
# all columns between name_long and pop (inclusive)
world2 = select(world, name_long:pop)
head(world2, n = 2)
```

```{r, eval=FALSE}
# all columns except subregion and area_km2 (inclusive)
world3 = select(world, -subregion, -area_km2)
head(world3, n = 2)
```

The `select()` function can be also used to both subset and rename columns in a single line, for example:

```{r}
world4 = select(world, name_long, population = pop)
head(world4, n = 2)
```

This is more concise than the base R equivalent (which saves the result as an object called `world5` to avoid overriding the `world` dataset created previously):

```{r, eval=FALSE}
world5 = world[c("name_long", "pop")] # subset columns by name
names(world5)[3] = "population" # rename column manually
```

The `select()` function works with a number of special functions that help with more complicated selection, such as `contains()`, `starts_with()`, `num_range()`. 
More details could be find on the function help page - `?select`.

The `filter()` function is a **dplyr** alternative to the `subset()` function.
Its role is to keeps rows matching given criteria.

```{r, eval=FALSE}
# only countries with a life expectation larger than 82 years
world6 = filter(world, lifeExp > 82)
```

The standard set of comparison operators can be used in the `filter()` function: 

```{r, echo=FALSE}
operators = c("`==`", "`!=`", "`>`", "`>=`", "`<`", "`<=`", "`&`", "|", "`!`")
operators_exp = c("Equal to", "Not equal to", "Greater than", "Greater than or equal",
                  "Less than", "Less than or equal", "And", "Or", "Not")
knitr::kable(data_frame(Symbol = operators, Name = operators_exp))
```

<!-- describe these: ==, !=, >, >=, <, <=, &, | -->
<!-- add warning about = vs == -->
<!-- add info about combination of &, |, ! -->

The *pipe* operator (` %>% `), which passes the output of one function into the first argument of the next function, is commonly used in **dplyr** data analysis workflows.
This works because the fundamental **dplyr** functions (or 'verbs', like `select()`) all take a data frame object in and spit a data frame object out.
Combining many functions together with pipes is called *chaining* or *piping*.
The advantage over base R for complex data processing operations is that this approach prevents nested functions and is easy to read because there is a clear order and modularity to the work (a piped command can be commented out, for example).

The example below shows yet another way of creating the renamed `world` dataset, using the pipe operator:

```{r}
world7 = world %>%
  select(name_long, continent)
```

Note that this can also be written without the pipe operator because, in the above code, the `world` object is simply 'piped' into the first argument of `select()`.
The equivalent **dplyr** code without the pipe operator is:

```{r}
world8 = select(world, name_long, continent)
```

The `pull()`function retrieves a variable by its name or position and returns a vector:

```{r, eval=FALSE}
world %>% 
  pull(name_long)
```

The pipe operator can be used for many data processing tasks with attribute data.

```{r}
# 1,000,000,000 could be expressed as 1e9 in the scientific notation 
world %>%
  filter(pop > 1e9) 
```

This is equivalent to the following base R code (not run to preserve the NAs):^[[Note](https://github.com/Robinlovelace/geocompr/issues/28) NAs do not work for subsetting by inequalities in base R, hence conversion of NAs to 0s in this version)]

```{r, eval=FALSE}
# subsetting simple feature rows by values
world$pop[is.na(world$pop)] = 0 # set NAs to 0
world_few_rows = world[world$pop > 1e9,]
```

The ` %>% ` operator works the best for combining many operations.
For example, we want to (1) rename the `name_long` column into a `name` column, (2) picks only `name`, `subregion` and `gdpPercap` and (3) subset countries from "Eastern Asia" with gross domestic product per capita larger than 30,000$:

```{r}
world %>% 
  select(name = name_long, subregion, gdpPercap) %>% 
  filter(subregion == "Eastern Asia", gdpPercap > 30000)
```

### Exercises

<!-- 
1. Select columns A and B
2. Select columns with "region" -> use contains() 
3. Find all countries that: (few points)
4. Exercises about additional select and filter helper functions (one_of; between, is.na, near, etc.)
 -->

## Attribute data aggregation 

<!-- https://github.com/ropenscilabs/skimr ?? -->

The `summary()` function is used to obtain summary statistics of `sf` objects:

```{r}
# data summary 
summary(world)
```

This function role is to give a quick glimpse of data in an interactive mode.
The `summary()` function, however, is not very customizable and should not be used to create new objects.
`summarise()` from the `dplyr` package can be used as an alternative in these cases.
In this function, we need to define what kind of statistics and for which variables we want to calculate.
For example, we could calculate the world's population and number of countries in the world:

```{r}
# customized data summary
world_summary = world %>% 
  summarise(pop = sum(pop, na.rm = TRUE), country_n = n())
world_summary
```

<!-- ?? open a github issue-->
<!-- ```{r} -->
<!-- plot(world_summary) -->
<!-- ``` -->

The list of useful functions could be find in the helpfile of this function - `?summarise`.
The whole power of `summarise()` is revealed when it is combined with the `group_by()` function.
This allow to summary requested subsets of data.
For example, we want to get the total population and number of countries on every continent:

```{r}
# data summary by groups
world_continents = world %>% 
  group_by(continent) %>% 
  summarise(pop = sum(pop, na.rm = TRUE), country_n = n())
world_continents
```

<!-- ?? open a github issue -->
<!-- ```{r} -->
<!-- plot(world_continents) -->
<!-- ``` -->

<!-- should it stay or should it go (?) aka should we present the arrange function?: -->

<!-- ```{r} -->
<!-- # sort variables -->
<!-- ## by name -->
<!-- world_continents %>%  -->
<!--   arrange(continent) -->
<!-- ## by population (in descending order) -->
<!-- world_continents %>%  -->
<!--   arrange(-pop) -->
<!-- ``` -->

### Exercises

<!-- -->

## Attribute data joining 

<!-- https://github.com/dgrtwo/fuzzyjoin -->
<!-- http://r4ds.had.co.nz/relational-data.html -->
<!-- non-unique keys -->

Combining data from different sources is one of the most common task in data preparation. 
It could be done using joins - methods created to work with a pair of tables.
The **dplyr** package has a set of verbs to easily connect `data.frames` - `left_join()`, `right_join()`,  `inner_join()`, `full_join`, `semi_join()` and `anti_join()`. 
They are thoroughly explained in the Relational data chapter in the book R for Data Science [@grolemund_r_2016].

Working with spatial data, however, usually involves a connection between spatial data (`sf` objects) and tables (`data.frame` objects).
Fortunately, the **sf** package has all of the **dplyr** join functions adapted to work with `sf` objects.
The only important difference between combining two `data.frames` and combining `sf` with `data.frame` is a `geom` column.
Therefore, the result of data joins could be either an `sf` or `data.frame` object.

The easiest way to understand the concept of joins is to use a smaller datasets. 
We will use an `sf` object `north_america` with country codes (`iso_a2`), names and geometries, as well as `data.frame` object `wb_north_america` containing information about urban population and unemployment for three countries. It is important to add that the first object has data about Canada, Greenland and United States and the second one has data about Canada, Mexico and United States:

```{r}
north_america = world %>%
  filter(subregion == "Northern America") %>%
  select(iso_a2, name_long)
north_america
```

```{r}
plot(north_america[0])
```

```{r}
wb_north_america = worldbank_df %>% 
  filter(name %in% c("Canada", "Mexico", "United States")) %>%
  select(name, iso_a2, urban_pop, unemploy = unemployment)

wb_north_america
```

In this book, we focus on spatial data. 
All of the following examples will have a `sf` object as the first argument and a `data.frame` object as the second argument. 
A new `sf` object will be a result of these joins. 
However, the reverse order is also possible and will result in a `data.frame` object.
This is beyond the scope of this book, but we encourage you to try it.

### Left joins

Left join is the most often used type of joins.
The `left_join()` returns all observations from the left object (`north_america`) and the matched observations from the right object (`wb_north_america`). 
In cases, like `Greenland`, when we don't have a data in the right object, `NA` values will be introduced.

To connect two object we need to specify a key.
This is a variable (or variables) that uniquely identifies each observation (row). 
The argument `by` is used to state which variable is the key. 
In simple cases, a single, unique variable exist in both objects, for example `iso_a2` column:

```{r}
left_join1 = north_america %>% 
  left_join(wb_north_america, by = "iso_a2")
left_join1
```

It is also possible to join objects by different variables.
Both of the datasets have variables with names of countries, but they are named differently.
The `north_america` has a `name_long` column and the `wb_north_america` has a `name` column.
In these cases, we can use a named vector to specify the connection, e.g. `c("name_long" = "name")`:

```{r}
left_join2 = north_america %>% 
  left_join(wb_north_america, by = c("name_long" = "name"))
left_join2
```

The new object `left_join2`, however is still not perfectly connected as it has two duplicated variables - `iso_a2.x` and `iso_a2.y`.
To solve this problem we should specify all the keys:

```{r}
left_join3 = north_america %>% 
  left_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
left_join3
```

<!-- ```{r} -->
<!-- # error: keeps geom col -->
<!-- left_join4 = wb_north_america %>%  -->
<!--   left_join(north_america, by = c("iso_a2")) -->
<!-- left_join4 -->
<!-- ``` -->

### Right joins

The `right_join()` keeps all observations from the right object (`wb_north_america`), but in the same time preserves a `sf` class from the left object (`north_america`).

```{r}
right_join1 = north_america %>% 
  right_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
right_join1
```

You can see that the new object `right_join1` has information about Mexico, but drop information about Greenland.
What's more, our right object, as a `data.frame`, doesn't have a geometry representation of Mexico.
As a result, the `right_join1` object contains only non-spatial data of Mexico.
It could be easily illustrated using the `plot` function:

```{r}
plot(right_join1[0]) # Canada and United States only
```

<!-- ```{r} -->
<!-- # error: unwanted geom column added -->
<!-- right_join2 = wb_north_america %>%  -->
<!--   right_join(north_america, by = "iso_a2") #%>% plot() -->
<!-- right_join2 -->
<!-- ``` -->

### Inner joins

The `inner_join()` keeps only observations from the left object (`north_america`) where there are matching observations in the right object (`wb_north_america`). 
Additionally, all columns from the left and right object are kept:

```{r}
inner_join1 = north_america %>% 
  inner_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
inner_join1
```

<!-- ```{r} -->
<!-- # error: geom column added -->
<!-- inner_join2 =  wb_north_america %>%  -->
<!--   inner_join(north_america, by = "iso_a2")  -->
<!-- inner_join2 -->
<!-- ``` -->

### Semi joins

The `semi_join()` is very similar to the `inner_join()`. 
It also keeps only observations from the left object (`north_america`) where there are matching observations in the right object, but keeping just columns from the left one:
<!-- filtering? -->

```{r}
semi_join1 = north_america %>% 
  semi_join(wb_north_america, by = "iso_a2")
semi_join1
```

<!-- ```{r} -->
<!-- semi_join2 = wb_north_america %>% -->
<!--   semi_join(north_america, by = "iso_a2") -->
<!-- semi_join2 -->
<!-- ``` -->

### Anti joins

The `anti_join()` returns all rows from the left object that are not matching observations in the right object.
Only columns from the right object are kept:

```{r}
anti_join1 = north_america %>% 
  anti_join(wb_north_america, by = "iso_a2")
anti_join1
```

```{r}
plot(anti_join1[0])
```

<!-- ```{r} -->
<!-- anti_join2 = wb_north_america %>%  -->
<!--   anti_join(north_america, by = "iso_a2") -->
<!-- anti_join2 -->
<!-- ``` -->

### Full joins

The `full_join()` returns all rows and all columns from both the left and right object. 
It also puts `NA` in cases where there are not matching values and returns an empty geometry for cases that only exist in the right object:

```{r}
full_join1 = north_america %>% 
  full_join(wb_north_america, by = "iso_a2")
full_join1
```

<!-- ```{r} -->
<!-- # error: null geom -->
<!-- full_join2 = wb_north_america %>%  -->
<!--   full_join(north_america, by = "iso_a2") #%>% plot() -->
<!-- full_join2 -->
<!-- ``` -->

### Exercises

<!-- -->

## Attribute data creation
<!-- lubridate? -->

It is often the case when a new column needs to be created based on existing columns.
For example, we want to calculate population density for each country.
We need to divide a `pop` column (population) by a `area_km2` column (unit area in square km).
It could be done this way in base R:

```{r}
world_new <- world # not to overwrite our original data
world_new$pop_dens = world_new$pop / world_new$area_km2
```

Alternatively, we can use one of **dplyr** functions - `mutate()` or `transmute()`.
`mutate()` adds new columns at second-to-last position in the `sf` object (the last one is reserved for the geometry):

```{r, eval=FALSE}
world %>% 
  mutate(pop_dens = pop / area_km2)
```

The difference between `mutate()` and `transmute()` is that the latter do not preserve existing columns:

```{r, eval=FALSE}
world %>% 
  transmute(pop_dens = pop / area_km2)
```

<!-- unite ?? -->
<!-- https://github.com/edzer/sfr/issues/378 -->

<!-- ```{r} -->
<!-- world %>%  -->
<!--   unite(con_reg, continent:region_un, sep = ":") -->
<!-- ``` -->

<!-- separate -->

Two helper functions, `rename()` and `set_names` can be used to change columns names.
The first one, `rename()` replace an old name with a new one.
For example, we want to change a name of column from `name_long` to `name`:

```{r, eval=FALSE}
world %>% 
  rename(name = name_long)
```

`set_names` can be used to change names of many columns. 
In this function, we do not need to provide old names: 

```{r, eval=FALSE}
new_names = c("ISO_A2", "Name", "Continent", "Region", "Subregion", 
              "Country_type", "Area_in_km2", "Population", "Life_Expectancy",
              "GDP_per_capita", "geom")
world %>% 
  set_names(new_names)
```

### Exercises

<!-- calculate gdp -->

## Removing spatial information

Most of the function from **sf** package do not drop a `geometry` column. To extract a data frame `st_geometry()` or `st_set_geometry()` function can be used.

```{r}
world_st = world
st_geometry(world_st) = NULL
class(world_st)

# OR

world_st2 = world
world_st2 = world_st2 %>% st_set_geometry(NULL)
class(world_st2)
```
