# Basic map making {#basic-map}

```{r, echo=FALSE, include=FALSE}
if(!exists("world"))
        source("code/03-attr.R")
```

Like **sp**, basic maps in **sf** can be created quickly with the base `plot()` function. Unlike **sp**, however, **sf** by default creates a faceted plot, one sub-plot for each variable, as illustrated in the left-hand image in Figure \@ref(fig:sfplot). 

```{r sfplot, fig.cap="Plotting with sf, with multiple variables (left) and a single variable (right).", out.width="49%", fig.show='hold', warning=FALSE}
plot(world)
plot(world["population"])
```

As with **sp**, you can add layers to your maps created with `plot()`, with the argument `add = TRUE`.^[In
fact, when you `plot()` an **sf** object, R is calling `sf:::plot.sf()` behind the scenes.
`plot()` is a generic method that behaves differently depending on the class of object being plotted.]
However, this only works if the initial plot has only 1 layer (result not shown):

```{r, fig.show='hide'}
plot(world["population"])
africa = world_continents[1,]
plot(africa, add = TRUE, col = "red")
```

This can be very useful when quickly checking the geographic correspondence between two or more layers.
These plots work well for gaining a quick understanding of the data with few lines of code.
For more advanced map making we recommend using a dedicated visualisation package such as **tmap**, **ggplot2**, **mapview**, or **leaflet**.

<!-- 
- plot() function 
- map export 
-->

## Challenge

Using **sf**'s `plot()` command, create a map of Nigeria in context, like the one presented in figure \@ref(fig:nigeria). 

- Hint: this used the `lwd`, `main` and `col` arguments of `plot()`. 
- Bonus: make the country boundaries a dotted grey line.
- Hint: `border` is an additional argument of `plot()` for **sf** objects.

```{r nigeria, warning=FALSE, echo=FALSE, fig.cap="Map of Nigeria in context illustrating sf's plotting capabilities"}
nigeria = filter(world, name == "Nigeria")
bb_africa = st_bbox(africa)
plot(africa[2], col = "white", lwd = 3, main = "Nigeria in context", border = "lightgrey")
# plot(world, lty = 3, add = TRUE, border = "grey")
plot(world, add = TRUE, border = "grey")
plot(nigeria, col = "yellow", add = TRUE, border = "darkgrey")
ncentre = st_centroid(nigeria)
ncentre_num = st_coordinates(ncentre)
text(x = ncentre_num[1], y = ncentre_num[2], labels = "Nigeria")
```

# Further work

**sf** makes R data objects more closely alligned to the data model used in GDAL and GEOS, in theory making spatial data operations faster.
The work here provides a taster of the way that **sf** operates but there is much more to learn.
There is a wealth of information that is available in the package's vignettes: these are highly recommended.

As a final exercise, we'll see how to do a spatial overlay in **sf** by first converting the countries of the world into centroids and then subsetting those in Africa:

```{r, out.width="50%", fig.cap="Centroids in Africa"}
world_centroids = st_centroid(world)
plot(world_centroids[1])
africa_centroids = world_centroids[africa,]
plot(africa_centroids, add = TRUE, cex = 2)
```

Note: another way of acheiving the same result is with a GEOS function for identifying spatial overlay:

```{r}
sel_africa = st_covered_by(world_centroids, africa, sparse = FALSE)
summary(sel_africa)
```

This shows that there are 56 countries in Africa.
We can check if they are the same countries as follows:

```{r}
africa_centroids2 = world_centroids[sel_africa,]
identical(africa_centroids, africa_centroids2)
```

## Exercises

- Perform the same operations and map making for another continent of your choice.
- Bonus: Download some global geographic data and add attribute variables assigning them to the continents of the world.

# References
