# Geographic data I/O {#read-write}

The previous chapters provided an overview of spatial data classes in R, with a focus on simple features and rasters.
This chapter is about getting spatial data onto your computer and then, perhaps after processing it with techniques described in this book, back out to the world.
<!-- Not sure if this is the place for the following two sentences... Or if so, some kind of link is missing.-->
We include a section (\@ref(visual-outputs)) on visualization because outputting data in a human (not just computer) readable format enables non-programmers to benefit from your work.
If your aim is to use geocomputation to improve the world, e.g. by encouraging evidence-based policies, this final stage is vital.

I/O is short for "input/output" which means, in plain English, "reading and writing data".
We use the acronym instead of plain English not to confuse you or to make chapter names short, but because that's the term used in computer science and it is useful to think of data import and export from a computing perspective.^[
Concepts such as computational efficiency, hard disk space and 'idempotence' are useful when thinking about reading and writing geographic datasets, which can become large and difficult to handle.
Loading/saving data is yet another way of saying the same thing.
]

## Data Input (I)

To efficiently read data into R, it helps to have an understanding of what happens 'under the hood'.
Executing commands such as `sf::st_read()` (the main function we use for loading vector data), `raster::raster()` (the main function used for loading raster data),  or `readr::read_csv()` (which can we used to read spatial data from text files) silently sets off a chain of events that loads objects.
"Loading" in this context means loading the data into R or, more precisely, assigning objects to your workspace, stored in RAM accessible from the `.GlobalEnv`^[See http://adv-r.had.co.nz/Environments.html for more information on the environment] of your current R session.

### Vector data

Spatial vector data comes in a wide variety of file formats, and **sf** is able to handle most of them via its `st_read()` command.
Behind the scenes it uses GDAL, which supports the import of a very wide range of vector data formats[^1].
One of the major advantages of **sf** is that it is fast.
<!-- reference to the vignette -->
The first argument of `st_read()` is `dsn`, which should be a text string or an object containing a single text string.
The content of a text string could vary between different drivers.
In the most of cases, e.g. `ESRI Shapefile` or `GeoPackage`, it would be a file name.
The `st_read` function, based on the file extension, knows how to read a file:

```{r, message=FALSE, warning=FALSE, results='hide'}
library(sf)
vector_filepath = system.file("shapes/world.gpkg", package = "spData")
world = st_read(vector_filepath)
```

For some drivers, `dsn` could be provided as a folder name, access credentials for a database, or a GeoJSON string representation.
<!-- Do we have a place for the examples? -->

Some of vector driver formats could store many layers of data. 
`st_read` automatically read the first layer of the file specified in `dsn`, however it is also possible to select a different layer by its name using the `layer` argument.

Many GDAL drivers have different open options[^1]. 
For example, spatial data in text files could be stored in several ways.
Simple point dataset could just have two columns describing their x and y coordinates. 
To read these files, you need to specify names of those columns using the `options` argument:

```{r, message=FALSE, warning=FALSE, results='hide'}
cycle_hire_txt = system.file("misc/cycle_hire_xy.csv", package = "spData")
cycle_hire_xy = st_read(cycle_hire_txt, options = c("X_POSSIBLE_NAMES=X",
                                                 "Y_POSSIBLE_NAMES=Y"))
```

More complex geometries cannot be described using only two numbers, however they could be represented by the well-known text (WKT), well-known binary (WKB), and in the GeoJSON form, for example:

```{r, message=FALSE}
world_txt = system.file("misc/world_wkt.csv", package = "spData")
world_wkt = st_read(world_txt, options = "GEOM_POSSIBLE_NAMES=WKT")
```

```{block2 type='rmdnote'}
Not all of the supported vector file formats store information about theirs coordinate reference system.
In these situations, it is possible to add the missing information using the `st_set_crs()` function.
More on that in the section \@ref(crs-intro).
```

<!-- isn't that confusing for users? -->
**Tip**: `read_sf()` and `write_sf()` can be used as easy-to-remember alternatives to `st_read()` and `st_write()`.
Remember they hide information about the data source and overwrite existing data, though.

To find out which data formats **sf** supports, run `st_drivers()`. 
Here, we show only the first two drivers:

```{r, eval=FALSE}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```

```{r, echo=FALSE}
sf_drivers = st_drivers()
knitr::kable(head(sf_drivers, n = 2))
```

### Raster data

Raster data also could exist in many different file formats, with some of the formats supporting multilayer files[^2].
The **raster** package support reading a single layer by `raster()` and many layers using either `brick()` or `stack()`. 
More information about the raster representation in this package could be find in the section \@ref(raster-classes).

The `raster()` function makes it possible to read a simple single layer file: 

```{r, message=FALSE}
library(raster)
library(spDataLarge)
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
single_layer = raster(raster_filepath)
```

You can also read just one layer (band) from a multilayer file:

```{r}
raster_multilayer_filepath = system.file("raster/landsat.tif", package = "spDataLarge")
single_layer_band3 = raster(raster_multilayer_filepath, band = 3)
```

The whole multilayer raster file is read by `brick()`:

```{r}
multilayer_layer_brick = brick(raster_multilayer_filepath)
```

Finally, many single and multilayer rasters could be read together using the `stack()` function:

```{r}
raster_l1_filepath = system.file("raster/landsat_b2.tif", package = "spDataLarge")
raster_l2_filepath = system.file("raster/landsat_b3.tif", package = "spDataLarge")
 
multilayer_layer_stack = stack(raster_l1_filepath, raster_l2_filepath)
```

The `stack()` class is limited to datasets that have the same spatial extent and resolution.

## Data output (O)

<!-- raster + vector + text intro -->

### Vector data

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

The counterpart of `st_read()` is `st_write()`.
It allows to write **sf** objects to a wide range of geographic vector file formats, including the most common ones such as `.geojson`, `.shp` and `.gpkg`.
Based on the file name, `st_write()` decides automatically which driver to use. 
How fast the writing process is depends also on the driver.
<!-- ref to the vignette -->

```{r}
st_write(obj = world, dsn = "world.gpkg")
```

**Note**: if you try to write to the same data source again, the function will fail:

```{r, error=TRUE}
st_write(obj = world, dsn = "world.gpkg")
```

<!-- ##   GDAL Error 1: Layer world.gpkg already exists, CreateLayer failed. -->
<!-- ## Use the layer creation option OVERWRITE=YES to replace it. -->

The error message provides some information as to why the function failed.
The `GDAL Error 1` statement makes clear that the failure occurred at the GDAL level.
Additionally, the suggestion to use `OVERWRITE=YES` provides a clue how to fix the problem.
However, this is not a `st_write()` argument, it is a GDAL option.
Luckily, `st_write` provides a `layer_options` argument through which we can pass driver-dependent options:

```{r, results='hide'}
st_write(obj = world, dsn = "world.gpkg", layer_options = "OVERWRITE=YES")
```

Another solution is to use the `st_write()` argument `delete_layer`. Setting it to `TRUE` deletes already existing layers in the data source before the function attempts to write (note there is also a `delete_dsn` argument):

```{r, results='hide'}
st_write(obj = world, dsn = "world.gpkg", delete_layer = TRUE)
```

You can achieve the same with `write_sf()` since it is equivalent to (technically an *alias* for) `st_write()`, except that its defaults for `delete_layer` and `quiet` is `TRUE`.
<!-- This enables spatial data to be overwritten more concisely, and with less output going to screen: -->
<!-- What does this mean: overwritten more concisely? -->

```{r}
write_sf(obj = world, dsn = "world.gpkg")
```

The `layer_options` argument could be also used for many different purposes.
One of them is to write spatial data to a text file.
This can be done by specifing `GEOMETRY` inside of `layer_options`. 
It could be either `AS_XY` for simple point datasets or `AS_WKT` for more complex spatial data:

```{r, eval=FALSE}
library(spData)
st_write(cycle_hire_xy, "cycle_hire_xy.csv", layer_options = "GEOMETRY=AS_XY")
st_write(world_wkt, "world_wkt.csv", layer_options = "GEOMETRY=AS_WKT")
```

### Raster data

The `writeRaster()` function save the `Raster*` objects to files. 
There are several decision to be made while saving rasters, such as an output data type, file format, and specific GDAL options related to selected file format.
The complete list of an arguments to this function could be found by typing `?writeRaster()`.

<!-- datatypes -->
The **raster** package offers nine following datatypes: LOG1S, INT1S, INT1U, INT2S, INT2U, INT4S, INT4U, FLT4S, and FLT8S.
Datatype determines the type of numbers and the range of the raster values that would be written to disk (\@ref(tab:datatypes)).
Decision which datatype should be used depends on the expected range of the values in the output file.
For example, LOG1S is used for bitmap (binary) images, unsigned integers (INT1U, INT2U, INT4U) are mostly used for categorical data, while floats (FLT4S and FLTS8S) usually represent continuous data.
The default datatype used in `writeRaster()` is FLT4S.
While it works in most of the cases, the size of the output file would be large.
Therefore, a rule of thumb is to use the smallest representation that fits all the values (range of values could be checked using the `summary()` function).

```{r datatypes, echo=FALSE}
dT = tibble::tribble(
               ~Datatype,      ~`Minimum value`,        ~`Maximum value`,
               "LOG1S",             "FALSE (0)",              "TRUE (1)",
               "INT1S",                  "-127",                   "127",
               "INT1U",                     "0",                   "255",
               "INT2S",               "-32,767",                "32,767",
               "INT2U",                     "0",                "65,534",
               "INT4S",        "-2,147,483,647",         "2,147,483,647",
               "INT4U",                     "0",         "4,294,967,296",
               "FLT4S",              "-3.4e+38",               "3.4e+38",
               "FLT8S",             "-1.7e+308",              "1.7e+308"
  )
knitr::kable(dT, caption = "Datatypes supported by the raster package")
```

To save the `Raster*` object to a file, you need to specify the file extension, for example you need to add the `.tif` extension to create a GeoTIFF file:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U")
```

The `raster` file format (native to the `raster` package) is used when a file extension is missing or is invalid. 

Some raster file formats also accept additional creation options[^2].
For example, in GeoTIFF you could specify a compress method using the `COMPRESS` option^[Creation options for the GeoTIFF format are at http://www.gdal.org/frmt_gtiff.html]:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U",
            options = c("COMPRESS=DEFLATE"))
```

Full list of the supported file format for writing `Raster*` objects could be found using `writeFormats().`

## File formats

<!-- an intro -->
<!-- many different formats -->
<!-- main three groups (vector, raster, geodatabase) -->
<!-- format model (open/proprietary) (importance of interoperabillity) -->
<!-- way of storage (single file, multiple files, folders) -->
<!-- GDAL (it's great - you can read, convert, and very often (though not always) write) -->
<!-- GDAL info "it is possible to have smaller number of supported formats than there are on the GDAL webpage; you may need to recompile..." -->

```{r, echo=FALSE}
file_formats = tibble::tribble(~Name, ~Extension, ~Info, ~Type, ~Model, 
                         "ESRI Shapefile", ".shp (the main file)", "One of the most popular vector file format. Consists of at least three files. The main files size cannot exceed 2 GB. It lacks support for mixed type. Columns names are limited to 10 characters, and number of columns are limited at 255. It has poor support for Unicode standard. ", "Vector", "Partially open",
                         "GeoJSON", ".geojson", "Format designed for representation of simple features. It is often used for a web applications.", "Vector", "Open",
                         "KML", ".kml", "XML-based format for spatial visualization, developed for use with Google Earth. Zipped KML file from the KMZ format.", "Vector", "Open",
                         "GPX", ".gpx", "XML schema created for exchange of GPS data.", "Vector", "Open",
                         "GeoTIFF", ".tiff", "GeoTIFF is one of the most popular raster formats. Its structure is similar to the regular `.tif` format, however GeoTIFF also stores additional spatial metadata, such as coordinate reference system, spatial extent, `NoData` value, and the data resolution.", "Raster", "Open",
                         "Arc ASCII", ".asc", "Text format where the first six lines represent grid information and nodata value, followed by the values arranged in rows and columns.", "Raster", "?",
                         # "JPEG??"
                         "R-raster", ".gri, .grd", "Native format of the raster package. ", "Raster", "Open",
                         # "NCDF??"
                         # "HDF??"
                         "SQLite/SpatiaLite", ".sqlite", "SQLite is a standalone, relational database management system. It is used as a default database driver in GRASS GIS 7. SpatiaLite is an extension to SQLite providing a support for Simple Features.", "Vector and raster", "Open",
                         # "WKT/WKB??",
                         "ESRI FileGDB", ".gdb", "Collection of spatial and nonspatial objects created in the ArcGIS software.", "Vector and raster", "",
                         "GeoPackage", ".gpkg", "An extended SQLite database file designed to be lightweight.", "Vector and raster", "Open"
                         # "WKT/WKB"??
                         )
knitr::kable(file_formats, caption = "Selected spatial file formats")
```


<!-- 3. JPEG - (possibly mention SAGA's sdat, Erdas Imagine) -->
<!-- 1. SQLite/SpatialLite + mention GRASS (uses SQLite) -->
<!-- 3. WKT/WKB for transfering and storing geometry data on databases. PostGIS (has even its own raster WKT (https://trac.osgeo.org/postgis/wiki/WKTRasterTutorial01); WKT also supported by Spatiallite, Oracle, MySQL, etc. (https://en.wikipedia.org/wiki/Well-known_text#RDBMS_Engines_that_provide_support) -->
<!-- 4. ESRI geodatabase, Oracle spatial database (mention + gdal support?) -->

## Visual outputs

<!-- aka how to save a map to a file -->

## Exercises

1. Name three differences between `write_sf()` and the more well-known function `st_write()`.

1. What are the default arguments of `read_sf()` and `write_sf()` that enable two of these differences?


[^1]: A list of supported vector formats and theirs options could be found at http://www.gdal.org/ogr_formats.html.

[^2]: Full list of supported raster formats with theirs options could be found at http://www.gdal.org/formats_list.html
