# Geographical data I/O {#read-write}

The previous chapters introduced this book and provided an overview of spatial data classes in R, with a focus on simple features.
This chapter is about getting spatial data onto your computer and then, perhaps after processing it with techniques described in this book, back out to the world.
We include a section (\@ref(visual-outputs)) on visualization because outputting data in a human (not just computer) readable enables non-programmers to benefit from your work.
If your aim is to use geocomputation to improve the world, e.g. by encouraging evidence-based policies, this final stage is vital.

I/O is short for "input/output" which means, in plain English, "reading and writing data".
We use the acronym instead of plain English not to confuse you or to make chapter names short, but because that's the term used in computer science and it is useful to think of data import and export from a computing perspective.^[
Concepts such as computational efficiency, hard disk space and 'idempotence' are useful when thinking about reading and writing geographical datasets, which can become large and difficult to handle.
Loading/saving data is yet another way of saying the same thing.
]

<!-- todo: check we mention computational efficiency, disk-space and 'idempotence' -->

<!-- Old intro to this chapter - can we salvage anything from this? -->
<!-- Reading, writing and plotting are 3 of the most important initial stages in geocomputation. -->
<!-- If you cannot load your data, it's not even worth starting. -->
<!-- If you cannot write your processed data, it will make it hard to collaborate with others. -->
<!-- Furthermore, an often-overlooked issue is that if you cannot create visuals of your spatial data, it will be hard to understand. -->
<!-- For this reason this introductory chapter also has a section on basic map making. -->

## Data Input (I)

To efficiently read data into R, it helps to have an understanding of what happens 'under the hood'.
Executing commands such as `sf::st_read` (the main function we use for loading spatial data, from the **sf** package) or `readr::read_csv` silently sets-off a chain of events that loads spatial objects.
"Loading" in this context means loading the data into R or, more precisely, assigning objects to your workspace, stored in RAM accessible from the `.GlobalEnv` of your current R session.
<!-- coud add a footnote here mentioning `.GlobalEnv` -->

Spatial data comes in a wide variety of file formats, and **sf** is adept at handling them, via the command `st_read`.
This function (also called `read_sf`) uses the power of the GDAL C/C++ library behind the scenes, allowing **sf** to read a very wide range of spatial data formats.
The first arguement of `st_read` is `file`, which should be a text string or an object containing a single text string:

```{r}
library(sf)
f = system.file("shapes/wrld.shp", package = "spData")
world = st_read(f)
```


A major advantage of **sf** is that it is fast at geographical data I/O, as illustrated in the benchmark below:

```{r, results='hide'}
library(microbenchmark)
bench_read = microbenchmark(times = 5,
        st_read(f),
        rgdal::readOGR(f)
)
```

```{r}
bench_read$time[1] / bench_read$time[2]
```

The results demonstrate that **sf** can be much faster (*`r round(bench_read$time[1] / bench_read$time[2], 0)` times faster* in this case) than **rgdal** at reading-in the world countries shapefile.

The counterpart of `st_read()` is `st_write()`. This allows writing to a range of geographic vector file types, including the common formats `.geojson`, `.shp` and `.gpkg`. `st_read()` will decide which driver to use automatically, based on the file name, as illustrated in the benchmark below demonstrating write speeds for each format.

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

```{r, warning=FALSE}
system.time(st_write(world, "world.geojson", quiet = TRUE))
system.time(st_write(world, "world.shp", quiet = TRUE)) 
system.time(st_write(world, "world.gpkg", quiet = TRUE))
```

The full range of file-types supported by **sf** is reported by `st_drivers()`, the first 2 of which are shown below:

```{r}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```

## Data output (O)

## Visual outputs


<!-- ## Vector -->

<!-- 
- sf package 
- st_drivers
- st_read, read_sf
- st_write, write_sf
- text files 
- WKT, WKB, sp (st_as_sf)
-->

<!-- ## Raster -->

<!-- 
- raster package 
- raster
-->
