# Geographic data I/O {#read-write}

The previous chapters provided an overview of spatial data classes in R, with a focus on simple features and rasters.
This chapter is about getting spatial data onto your computer and then, perhaps after processing it with techniques described in this book, back out to the world.
Geocomputation usually involves reading existing data, and saving it after processing sections.
Therefore, it is important be aware of the most often used spatial file format and their strong and weak sides (see section \@ref(file-formats)).
For the same purposes knowing how to efficiently read and write spatial data is crucial (see sections \@ref(data-input) and \@ref(data-output)).
We also include a section (\@ref(visual-outputs)) on visualization because outputting data in a human (not just computer) readable format enables non-programmers to benefit from your work.
If your aim is to use geocomputation to improve the world, e.g. by encouraging evidence-based policies, this final stage is vital.

I/O is short for "input/output" which means, in plain English, "reading and writing data".
We use the acronym instead of plain English not to confuse you or to make chapter names short, but because that's the term used in computer science and it is useful to think of data import and export from a computing perspective.^[
Concepts such as computational efficiency, hard disk space and 'idempotence' are useful when thinking about reading and writing geographic datasets, which can become large and difficult to handle.
Loading/saving data is yet another way of saying the same thing.
]

## File formats

Spatial data is usually stored as files or geodatabases. 
File-based data formats could keep either vector or raster data, while geodatabases could store both.
Historically, GIS file formats were developed by mapping agencies and software companies. <!--examples-->
It caused many issues with exchange of data between different sources.

The big improvement came with the release of GDAL. <!--name and pronunciation-->
It has enabled for reading and writing many raster and vector data formats and in the next few years was incorporated into many open and proprietary GIS software, such as GRASS GIS, QGIS, ArcGIS, and ENVI.
<!-- GDAL (it's great - you can read, convert, and very often (though not always) write) -->
<!-- GDAL info "it is possible to have smaller number of supported formats than there are on the GDAL webpage; you may need to recompile..." -->

Another change for spatial data formats came with the foundation of Open Geospatial Consortium (OGC)^[http://www.opengeospatial.org].
This organization collaborate on a development and implementation of open standards for geospatial content including file formats.
Some of theirs products are the KML and GeoPackage formats, as well as the Simple Feature standard. 
Development and maintaining of spatial file formats in an open model provides several benefits over the proprietary formats and eases interoperability.

Nowadays, more than a hundred of spatial data formats exist, many of which could be used for the same purposes. 
<!-- In the same time, they could differ in many ways. -->
<!-- Spatial data could be stored as a single file (e.g. GeoPackage), multiple files (e.g. ESRI Shapefile), or folders (ESRI ArcInfo Coverages). -->
<!-- way of storage (single file, multiple files, folders) -->
Table \@ref(tab:formats) presents some basic information about selected, often used spatial file formats.


```{r formats, echo=FALSE}
file_formats = tibble::tribble(~Name, ~Extension, ~Info, ~Type, ~Model, 
                         "ESRI Shapefile", ".shp (the main file)", "One of the most popular vector file format. Consists of at least three files. The main files size cannot exceed 2 GB. It lacks support for mixed type. Columns names are limited to 10 characters, and number of columns are limited at 255. It has poor support for Unicode standard. ", "Vector", "Partially open",
                         "GeoJSON", ".geojson", "Format designed for representation of simple features. It is often used for a web applications.", "Vector", "Open",
                         "KML", ".kml", "XML-based format for spatial visualization, developed for use with Google Earth. Zipped KML file forms the KMZ format.", "Vector", "Open",
                         "GPX", ".gpx", "XML schema created for exchange of GPS data.", "Vector", "Open",
                         "GeoTIFF", ".tiff", "GeoTIFF is one of the most popular raster formats. Its structure is similar to the regular `.tif` format, however GeoTIFF also stores additional spatial metadata, such as coordinate reference system, spatial extent, `NoData` value, and the data resolution.", "Raster", "Open",
                         "Arc ASCII", ".asc", "Text format where the first six lines represent grid information and nodata value, followed by the values arranged in rows and columns.", "Raster", "Open",
                         # "JPEG??"
                         "R-raster", ".gri, .grd", "Native format of the raster package.", "Raster", "Open",
                         # "NCDF??"
                         # "HDF??"
                         "SQLite/SpatiaLite", ".sqlite", "SQLite is a standalone, relational database management system. It is used as a default database driver in GRASS GIS 7. SpatiaLite is an extension to SQLite providing a support for Simple Features.", "Vector and raster", "Open",
                         # "WKT/WKB??",
                         "ESRI FileGDB", ".gdb", "Collection of spatial and nonspatial objects created in the ArcGIS software. It allows to store multiple feature classes and enables use of topological definitions. Limited access to this format is provided by GDAL with the use of the OpenFileGDB and FileGDB drivers.", "Vector and raster", "Proprietary",
                         "GeoPackage", ".gpkg", "An extended SQLite database file designed to be lightweight. All elements, such as spatial data, metadata, attribute tables, coordinate reference system, are stored in a single file.", "Vector and raster", "Open"
                         # "WKT/WKB"??
                         )
knitr::kable(file_formats, caption = "Selected spatial file formats")
```

<!-- 3. JPEG - (possibly mention SAGA's sdat, Erdas Imagine) -->
<!-- 1. SQLite/SpatialLite + mention GRASS (uses SQLite) -->
<!-- 3. WKT/WKB for transfering and storing geometry data on databases. PostGIS (has even its own raster WKT (https://trac.osgeo.org/postgis/wiki/WKTRasterTutorial01); WKT also supported by Spatiallite, Oracle, MySQL, etc. (https://en.wikipedia.org/wiki/Well-known_text#RDBMS_Engines_that_provide_support) -->
<!-- 4. ESRI geodatabase, Oracle spatial database (mention + gdal support?) -->

## Data Input (I) {#data-input}

To efficiently read data into R, it helps to have an understanding of what happens 'under the hood'.
Executing commands such as `sf::st_read()` (the main function we use for loading vector data) or `raster::raster()` (the main function used for loading raster data) silently sets off a chain of events that reads data from files.
Moreover, there are many R packages containing wide range of spatial data or providing simple access to different data sources.
All of them load the data into R or, more precisely, assign objects to your workspace, stored in RAM accessible from the `.GlobalEnv`^[See http://adv-r.had.co.nz/Environments.html for more information on the environment] of your current R session.

### Vector data

Spatial vector data comes in a wide variety of file formats, most of which can be read-in via the **sf** function `st_read()`.
Behind the scenes this calls GDAL.
<!-- One of the major advantages of **sf** is that it is fast. -->
<!-- reference to the vignette -->
The first argument of `st_read()` is `dsn`, which should be a text string or an object containing a single text string.
The content of a text string could vary between different drivers.
In most cases, as with `.shp` file of `ESRI Shapefile` or `.gpkg` files of `GeoPackage` formats, the `dsn` would be a file name.
`st_read()` guesses the driver based on the file extension, as illustrated for a `.gpkg` file below:

```{r}
library(sf)
vector_filepath = system.file("shapes/world.gpkg", package = "spData")
world = st_read(vector_filepath)
```

For some drivers, `dsn` could be provided as a folder name, access credentials for a database, or a GeoJSON string representation.
<!-- Do we have a place for the examples? -->

Some of vector driver formats could store many layers of data. 
`st_read` automatically read the first layer of the file specified in `dsn`, however it is also possible to select a different layer by its name using the `layer` argument.

Many GDAL drivers have unique open options[^1]. 
For example, spatial data in text files could be stored in several ways.
Simple point dataset could just have two columns describing their x and y coordinates. 
To read these files, you need to specify names of those columns using the `options` argument:

```{r, results='hide'}
cycle_hire_txt = system.file("misc/cycle_hire_xy.csv", package = "spData")
cycle_hire_xy = st_read(cycle_hire_txt, options = c("X_POSSIBLE_NAMES=X",
                                                    "Y_POSSIBLE_NAMES=Y"))
```

More complex geometries cannot be described using only two numbers, however they could be represented by the well-known text (WKT), well-known binary (WKB), and in the GeoJSON form, for example:

```{r, results='hide'}
world_txt = system.file("misc/world_wkt.csv", package = "spData")
world_wkt = st_read(world_txt, options = "GEOM_POSSIBLE_NAMES=WKT")
```

```{block2 type='rmdnote'}
Not all of the supported vector file formats store information about theirs coordinate reference system.
In these situations, it is possible to add the missing information using the `st_set_crs()` function.
More on that in section \@ref(crs-intro).
```

<!-- isn't that confusing for users? -->
**Tip**: `read_sf()` and `write_sf()` can be used as easy-to-remember alternatives to `st_read()` and `st_write()`.
Remember they hide information about the data source and overwrite existing data, though.

To find out which data formats **sf** supports, run `st_drivers()`. 
Here, we show only the first two drivers (see Table \@ref(tab:drivers)):

```{r, eval=FALSE}
sf_drivers = st_drivers()
head(sf_drivers, n = 5)
```

```{r drivers, echo=FALSE}
sf_drivers = st_drivers() %>% dplyr::filter(name %in% c("ESRI Shapefile", "GeoJSON", "KML", "GPX", "GPKG"))
knitr::kable(head(sf_drivers, n = 5), caption = "Sample of available drivers for reading/writing vector data (it could vary between different GDAL versions).")
```

### Raster data

Raster data also could exist in many different file formats, with some of the formats supporting multilayer files.
The **raster** package allow reading a single layer by `raster()` and many layers using either `brick()` or `stack()`. 
More information about the raster representation in this package could be find in the section \@ref(raster-classes).

The `raster()` function makes it possible to read a single layer file.

```{r, message=FALSE}
library(raster)
library(spDataLarge)
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
single_layer = raster(raster_filepath)
```

You can also read just one layer (band) from a multilayer file by specifying the `band` argument.

```{r}
raster_multilayer_filepath = system.file("raster/landsat.tif", package = "spDataLarge")
single_layer_band3 = raster(raster_multilayer_filepath, band = 3)
```

The whole multilayer raster file is read by `brick()`.

```{r}
multilayer_layer_brick = brick(raster_multilayer_filepath)
```

Finally, many single and multilayer rasters could be read together using the `stack()` function:

```{r}
raster_l1_filepath = system.file("raster/landsat_b2.tif", package = "spDataLarge")
raster_l2_filepath = system.file("raster/landsat_b3.tif", package = "spDataLarge")
 
multilayer_layer_stack = stack(raster_l1_filepath, raster_l2_filepath)
```

The `stack()` class is limited to datasets that have the same spatial extent and resolution.

### Data packages

Nowadays, a vast amount of spatial data is available on the internet.
This data could be accessed using a graphical user interface (GUI) via web browser.
However, in this book we encourage you to create reproducible workflows.
Data download in many cases could be automated, which not only save a time, but also allow to repeat and update this process in the future. 

Data could be stored online in many ways. 
Traditionally, is has been distributed on servers as files.
<!-- that's probably not the best example - replace it with something better -->
<!-- btw naturalearth website has some problems today - the link will probably change in the future -->
You can easily access data files from R by `download.file()`, for example National Park Service units in the United States:

```{r, eval=FALSE}
download.file(url = "http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_parks_and_protected_lands.zip",
              destfile = "USA_parks.zip")
unzip(zipfile = "USA_parks.zip")
usa_parks = st_read("ne_10m_parks_and_protected_lands_area.shp")
```

Many open spatial datasets can be retrieved using R packages (Table \@ref(tab:datapackages)).
These packages either store small datasets or provide an access to dataset available on webservers.

```{r datapackages, echo=FALSE}
datapackages = tibble::tribble(~`Package name`, ~Description,
                               "osmdata", "An access to the OpenStreetMap data.",
                               "rnaturalearth", "Functions to download Natural Earth vector and raster data, including world country borders.",
                               "rnoaa", "An R interface to many National Oceanic and Atmospheric Administration (NOAA) data sources, such as climate data, sea ice data, and storm data.",
                               "raster", "The `get_data()` function provides for example elevation data from SRTM and interpolated climate data from WorldClim.", 
                               "rWBclimate", "An access to the World Bank climate data used in the World Bank climate knowledge portal."
)
knitr::kable(datapackages, caption = "Selected R packages for spatial data retrieval")
```

<!-- https://cdn.rawgit.com/Nowosad/Intro_to_spatial_analysis/05676e29/Intro_to_spatial_analysis.html#39 -->
<!-- Maybe add a section to Data I/O on where and how to retrieve data (with a focus on free data): osmdata (OpenStreetMap; maybe mention TomTom, HERE), Landsat (wrspathrow), Sentinel (mention Python API), AVHRR, RapidEye rgbif, letsR, etc. Of course, point to Transforming science through open data project (https://www.ropensci.org) -->
<!-- https://github.com/ropensci/GSODR -->
<!-- https://github.com/lbusett/MODIStsp -->
<!-- https://github.com/walkerke/tigris -->
<!-- https://github.com/ropensci/hddtools/ -->

For example, you can get borders of any country with the `ne_countries()` function from the **rnaturalearth** package:
```{r}
library(rnaturalearth)
usa = ne_countries(country = "United States of America") # United States borders
class(usa)
```

As a default, this and many other packages output data in the `Spatial*` class. 
You can easily convert it to the `sf` class with `st_as_sf()`:

```{r}
usa_sf = st_as_sf(usa)
```

It is also possible to get a raster data, for example using `getData()`.
The code below download a series of rasters that contain global monthly precipitation sums in a ten minutes spatial resolution.
As a result, the new object is a multilayer object of the `RasterStack` class.

```{r}
library(raster)
worldclim_prec = getData(name = "worldclim", var = "prec", res = 10)
class(worldclim_prec)
worldclim_prec[, 1]
```

Finally, many R packages contain example data as R objects or in a raw form.
Data in native R format could be accessed with the `data()` function.
For example, you can get a list of dataset in a package, `data(package = "spData")`, or load a dataset by specifying the name of a dataset, `data("cycle_hire", package = "spData")`.
To access a single raw file from the package, you need to specify the package name and the relative path to the dataset, for example:

```{r}
world_raw_filepath = system.file("shapes/world.gpkg", package = "spData")
world_raw = st_read(world_raw_filepath)
```

More information on getting data using R packages could be found in [section 5.5](https://csgillespie.github.io/efficientR/input-output.html#download) and [section 5.6](https://csgillespie.github.io/efficientR/input-output.html#accessing-data-stored-in-packages) of @gillespie_efficient_2016.

## Data output (O) {#data-output}

The ability of writing spatial data could be used for conversion between different formats and for saving newly created objects.
Depending on the data type (vector or raster), object class (e.g `multipoint` or `RasterLayer`), and type and amount of stored information (e.g. object size, range of values) - it is important to know how to store spatial files in the most efficient way.
The next two section will show how to do that.

<!-- should we add a note about recommended way to decide on a file name, for example "don't use spaces in the name", "create descriptive names" -->

### Vector data

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

The counterpart of `st_read()` is `st_write()`.
It allows to write **sf** objects to a wide range of geographic vector file formats, including the most common ones such as `.geojson`, `.shp` and `.gpkg`.
Based on the file name, `st_write()` decides automatically which driver to use. 
How fast the writing process is depends also on the driver.
<!-- ref to the vignette -->

```{r}
st_write(obj = world, dsn = "world.gpkg")
```

**Note**: if you try to write to the same data source again, the function will fail:

```{r, error=TRUE}
st_write(obj = world, dsn = "world.gpkg")
```

<!-- ##   GDAL Error 1: Layer world.gpkg already exists, CreateLayer failed. -->
<!-- ## Use the layer creation option OVERWRITE=YES to replace it. -->

The error message provides some information as to why the function failed.
The `GDAL Error 1` statement makes clear that the failure occurred at the GDAL level.
Additionally, the suggestion to use `OVERWRITE=YES` provides a clue how to fix the problem.
However, this is not a `st_write()` argument, it is a GDAL option.
Luckily, `st_write` provides a `layer_options` argument through which we can pass driver-dependent options:

```{r, results='hide'}
st_write(obj = world, dsn = "world.gpkg", layer_options = "OVERWRITE=YES")
```

Another solution is to use the `st_write()` argument `delete_layer`. Setting it to `TRUE` deletes already existing layers in the data source before the function attempts to write (note there is also a `delete_dsn` argument):

```{r, results='hide'}
st_write(obj = world, dsn = "world.gpkg", delete_layer = TRUE)
```

You can achieve the same with `write_sf()` since it is equivalent to (technically an *alias* for) `st_write()`, except that its defaults for `delete_layer` and `quiet` is `TRUE`.
<!-- This enables spatial data to be overwritten more concisely, and with less output going to screen: -->
<!-- What does this mean: overwritten more concisely? -->

```{r}
write_sf(obj = world, dsn = "world.gpkg")
```

<!-- how about saving multilayer gpkg? -->

The `layer_options` argument could be also used for many different purposes.
One of them is to write spatial data to a text file.
This can be done by specifing `GEOMETRY` inside of `layer_options`. 
It could be either `AS_XY` for simple point datasets (it creates two new columns for coordinates) or `AS_WKT` for more complex spatial data (one new column is created which contains the well-known-text representation of spatial objects).

```{r, eval=FALSE}
library(spData)
st_write(cycle_hire_xy, "cycle_hire_xy.csv", layer_options = "GEOMETRY=AS_XY")
st_write(world_wkt, "world_wkt.csv", layer_options = "GEOMETRY=AS_WKT")
```

### Raster data

The `writeRaster()` function save the `Raster*` objects to files. 
There are several decision to be made while saving rasters, such as an output data type, file format, and specific GDAL options related to selected file format.
The complete list of an arguments to this function could be found by typing `?writeRaster()`.

<!-- datatypes -->
The **raster** package offers nine following datatypes: LOG1S, INT1S, INT1U, INT2S, INT2U, INT4S, INT4U, FLT4S, and FLT8S^[Using INT4U is not recommended as R does not support 32-bit unsigned integers.<!--recheck this info-->].
Datatype determines the type of numbers and the range of the raster values that would be written to disk (\@ref(tab:datatypes)).
Decision which datatype should be used depends on the expected range of the values in the output file.
For example, LOG1S is used for bitmap (binary) images, unsigned integers (INT1U, INT2U, INT4U) are mostly used for categorical data, while floats (FLT4S and FLTS8S) usually represent continuous data.
The default datatype used in `writeRaster()` is FLT4S.
While it works in most of the cases, the size of the output file would be large.
Therefore, a rule of thumb is to use the smallest representation that fits all the values (range of values could be checked using the `summary()` function).

```{r datatypes, echo=FALSE}
dT = tibble::tribble(
               ~Datatype,      ~`Minimum value`,        ~`Maximum value`,
               "LOG1S",             "FALSE (0)",              "TRUE (1)",
               "INT1S",                  "-127",                   "127",
               "INT1U",                     "0",                   "255",
               "INT2S",               "-32,767",                "32,767",
               "INT2U",                     "0",                "65,534",
               "INT4S",        "-2,147,483,647",         "2,147,483,647",
               "INT4U",                     "0",         "4,294,967,296",
               "FLT4S",              "-3.4e+38",               "3.4e+38",
               "FLT8S",             "-1.7e+308",              "1.7e+308"
  )
knitr::kable(dT, caption = "Datatypes supported by the raster package")
```

To save the `Raster*` object to a file, the file extension needs to be specified. 
For example you need to add the `.tif` extension to create a GeoTIFF file:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U")
```

The `raster` file format (native to the `raster` package) is used when a file extension is missing or is invalid. 

Some raster file formats also accept additional creation options with the `options` argument[^2].
For example, in GeoTIFF you could specify a compress method using the `COMPRESS` option^[Creation options for the GeoTIFF format are at http://www.gdal.org/frmt_gtiff.html]:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U",
            options = c("COMPRESS=DEFLATE"))
```

Full list of the supported file format for writing `Raster*` objects could be found using `writeFormats().`

## Visual outputs

R support many different static and interactive graphics formats.

The most general method to save a static plot is to open a graphic device, create a plot, and close a graphic device, for example:

```{r, eval=FALSE}
png(filename = "lifeExp.png", width = 500, height = 350)
plot(world["lifeExp"])
dev.off()
```

Each of the available formats, `pdf()`, `bmp()`, `jpeg()`, `png()`, and `tiff()`, has its own function. 
You can specify several properties of the output plot, including width and height.

Additionally, several graphic packages provide its own function to save a graphical output.
For example, the **tmap** package has the `save_tmap()` function.
You can save a `tmap` object to different graphic formats by specifying the object name and a file path to a new graphic file.

```{r, eval=FALSE}
library(tmap)
tmap_obj = tm_shape(world) +
  tm_polygons(col = "lifeExp")
save_tmap(tm  = tmap_obj, filename = "lifeExp_tmap.png")
```

<!-- Note about that the `plot` function do not create an object -->
<!-- ```{r} -->
<!-- a = plot(world["lifeExp"]) -->
<!-- ``` -->

On the other hand, an interactive map could be usually saved as an HTML file. 
This can be done with the `saveWidget()` function from the `htmlwidget` package:

```{r, eval=FALSE}
library(mapview)
library(htmlwidgets)
mapview_obj = mapview(world, zcol = "lifeExp", legend = TRUE)
saveWidget(mapview_obj, file="my_interactive_map.html")
```

## Exercises

1. List and describe three types of vector, raster, and geodatabase formats.

1. Name at least differences between `read_sf()` and the more well-known function `st_read()`.

1. Read the `cycle_hire_xy.csv` file from the **spData** package (hint: it is in the `misc\` folder).
What is a geometry type of a new object? 

1. Download Germany borders data using **rnaturalearth** and create a new object called `germany_borders`.
Write this new object to a file of the GeoPackage format.

1. Download raster data of global monthly minimum temperature with five minutes resolution using the **raster** package.
Extract only values for June and save them to the `tmin_june.tif` file (hint: use `raster::subset()`).

1. Create a static map of Germany borders and save it to a PNG file.

1. Create an interactive map using data from the `cycle_hire_xy.csv` file. 
Export this map to the HTML file called `cycle_hire.html`.

[^1]: A list of supported vector formats and theirs options could be found at http://www.gdal.org/ogr_formats.html.

[^2]: Full list of supported raster formats with theirs options could be found at http://www.gdal.org/formats_list.html
