# Geographic data I/O {#read-write}

The previous chapters introduced this book and provided an overview of spatial data classes in R, with a focus on simple features.
This chapter is about getting spatial data onto your computer and then, perhaps after processing it with techniques described in this book, back out to the world.
We include a section (\@ref(visual-outputs)) on visualization because outputting data in a human (not just computer) readable enables non-programmers to benefit from your work.
If your aim is to use geocomputation to improve the world, e.g. by encouraging evidence-based policies, this final stage is vital.

I/O is short for "input/output" which means, in plain English, "reading and writing data".
We use the acronym instead of plain English not to confuse you or to make chapter names short, but because that's the term used in computer science and it is useful to think of data import and export from a computing perspective.^[
Concepts such as computational efficiency, hard disk space and 'idempotence' are useful when thinking about reading and writing geographic datasets, which can become large and difficult to handle.
Loading/saving data is yet another way of saying the same thing.
]

<!-- todo: check we mention computational efficiency, disk-space and 'idempotence' -->

<!-- Old intro to this chapter - can we salvage anything from this? -->
<!-- Reading, writing and plotting are 3 of the most important initial stages in geocomputation. -->
<!-- If you cannot load your data, it's not even worth starting. -->
<!-- If you cannot write your processed data, it will make it hard to collaborate with others. -->
<!-- Furthermore, an often-overlooked issue is that if you cannot create visuals of your spatial data, it will be hard to understand. -->
<!-- For this reason this introductory chapter also has a section on basic map making. -->

## Data Input (I)

To efficiently read data into R, it helps to have an understanding of what happens 'under the hood'.
Executing commands such as `sf::st_read` (the main function we use for loading spatial data, from the **sf** package) or `readr::read_csv` silently sets-off a chain of events that loads spatial objects.
"Loading" in this context means loading the data into R or, more precisely, assigning objects to your workspace, stored in RAM accessible from the `.GlobalEnv` of your current R session.
<!-- coud add a footnote here mentioning `.GlobalEnv` -->

Spatial data comes in a wide variety of file formats, and **sf** is adept at handling them, via the command `st_read`.
This function uses GDAL behind the scenes, enabling a very wide range of spatial data formats to be imported.
The first arguement of `st_read` is `file`, which should be a text string or an object containing a single text string:

```{r}
library(sf)
f = system.file("shapes/world.gpkg", package = "spData")
world = st_read(f)
```

**Tip**: `read_sf()` and `write_sf()` can be used as easy-to-remember alternatives to `st_read()` and `st_write()`. Remember they hide information about the data source and overwrite existing data, though..


A major advantage of **sf** is that it is fast.
To demonstrate this, we will use a function to compare `st_read` with it's **sp** equivalent, `rgdal::readOGR`:

```{r}
bench_read = function(file, n) {
  m = microbenchmark(times = n,
                     rgdal::readOGR(f),
                     st_read(f)
  )
  mean(m$time[1:n]) / mean(m$time[(n + 1):(n * 2)])
}
```

This function takes as arguments an input file (`file`) and a number of times to run each command (`n`) and returns how many times faster `st_read()` is than `readOGR()`.
Let's run the benchmark for the `world.gpkg` file represented by the object `f`:
We as illustrated in the benchmark below:

```{r, results='hide'}
library(microbenchmark)
read_world_gpkg = bench_read(file = f, n = 5)
```

```{r}
read_world_gpkg
```


The results demonstrate that **sf** was around `r round(read_world_gpkg)` times faster than **rgdal** at reading-in the world countries shapefile.
The relative performance of `st_read()` compared with other functions will vary depending on file format and the nature of the data.
To illustrate this point, we performed the same operation on a geojson file and found a greater speed saving:

```{r, results='hide'}
f = system.file("shapes/lnd.geojson", package = "spData")
read_lnd_geojson = bench_read(file = f, n = 5)
```

```{r}
read_lnd_geojson
```

In this case **sf** was around `r round(read_lnd_geojson)` times faster than **rgdal**.

The full range of file-types supported by **sf** is reported by `st_drivers()`, the first 2 of which are shown below:

```{r}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```

## Data output (O)

```{r}
st_write(obj = world, dsn = "world.gpkg")
```

**Note**: if you try write to the same data source again will fail.
This is demonstrated in the code below for a modified version of the world in which the population doubles in all countries (don't worry about the `dplyr` code for now, this is covered in Chapter \@ref(attr)):

```{r}
world_mod = dplyr::mutate(world, pop = pop * 2)
```

```{r, eval=FALSE}
st_write(obj = world, dsn = "world.gpkg")
##   GDAL Error 1: Layer world.gpkg already exists, CreateLayer failed.
## Use the layer creation option OVERWRITE=YES to replace it.
```

The error message provides some information about why it failed, some of which is provided in the comments.
It is an issue at the GDAL level.
This is clear from the statement `GDAL Error 1`.
A further clue is provided by suggestion to use `OVERWRITE=YES`: this is not an option in `st_write`.
To address this potentially common issue, `st_write()` has an argument, `delete_layer`, that deletes the previous layers in the data source before attempting to write (note there is also a `delete_dsn` argument).
Setting this argument to `TRUE` makes the rewrite operation work:

```{r}
st_write(obj = world_mod, dsn = "world.gpkg", delete_layer = TRUE)
```

This can also be done with the function `write_sf()`, which is equivalent to (technically an *alias* for) `st_write()`, except that it has `delete_layer = TRUE` and `quiet = TRUE` by default.
This enables spatial data to be overwritten more concisely and with less output going to screen:

```{r}
write_sf(obj = world_mod, dsn = "world.gpkg")
```

A blunter way to update file-based geographic data sources such as `.gpkg` files is to simply delete them.
This is not generally recommended, as it will not work for multi-file data sources such as `.shp` files:

```{r}
file.remove("world.gpkg")
```


The counterpart of `st_read()` is `st_write()`. This allows writing to a range of geographic vector file types, including the common formats `.geojson`, `.shp` and `.gpkg`. `st_read()` will decide which driver to use automatically, based on the file name, as illustrated in the benchmark below demonstrating write speeds for each format.

```{r, warning=FALSE}
system.time(st_write(world, "world.geojson", quiet = TRUE))
system.time(st_write(world, "world.shp", quiet = TRUE)) 
system.time(st_write(world, "world.gpkg", quiet = TRUE))
```

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```


## File formats

## Visual outputs

## Exercises


1. Name three differences between `write_sf()` and the more well-known function `st_write()`.

1. What are the default arguments of `read_sf()` and `write_sf()` that enable two of these differences?


<!-- ## Vector -->

<!-- 
- sf package 
- st_drivers
- st_read, read_sf
- st_write, write_sf
- text files 
- WKT, WKB, sp (st_as_sf)
-->

<!-- ## Raster -->

<!-- 
- raster package 
- raster
-->
