# Geographic data I/O {#read-write}

The previous chapters provided an overview of spatial data classes in R, with a focus on simple features and rasters.
This chapter is about getting spatial data onto your computer and then, perhaps after processing it with techniques described in this book, back out to the world.
<!-- Not sure if this is the place for the following two sentences... Or if so, some kind of link is missing.-->
We include a section (\@ref(visual-outputs)) on visualization because outputting data in a human (not just computer) readable format enables non-programmers to benefit from your work.
If your aim is to use geocomputation to improve the world, e.g. by encouraging evidence-based policies, this final stage is vital.

I/O is short for "input/output" which means, in plain English, "reading and writing data".
We use the acronym instead of plain English not to confuse you or to make chapter names short, but because that's the term used in computer science and it is useful to think of data import and export from a computing perspective.^[
Concepts such as computational efficiency, hard disk space and 'idempotence' are useful when thinking about reading and writing geographic datasets, which can become large and difficult to handle.
Loading/saving data is yet another way of saying the same thing.
]

## Data Input (I)

To efficiently read data into R, it helps to have an understanding of what happens 'under the hood'.
Executing commands such as `sf::st_read()` (the main function we use for loading vector data), `raster::raster()` (the main function used for loading raster data),  or `readr::read_csv()` (which can we used to read spatial data from text files) silently sets off a chain of events that loads objects.
"Loading" in this context means loading the data into R or, more precisely, assigning objects to your workspace, stored in RAM accessible from the `.GlobalEnv`^[See http://adv-r.had.co.nz/Environments.html for more information on the environment] of your current R session.

### Vector data

Spatial vector data comes in a wide variety of file formats, and **sf** is able to handle most of them via its `st_read()` command.
Behind the scenes it uses GDAL, which supports the import of a very wide range of vector data formats[^1].
The first argument of `st_read()` is `dsn`, which should be a text string or an object containing a single text string.
The content of a text string could vary between different drivers.
In the most of cases, e.g. `ESRI Shapefile` or `GeoPackage`, it would be a file name.
The `st_read` function, based on the file extension, knows how to read a file:

```{r, message=FALSE}
library(sf)
vector_filepath = system.file("shapes/world.gpkg", package = "spData")
world = st_read(vector_filepath)
```

For some drivers, `dsn` could be provided as a folder name, access credentials for a database, or a GeoJSON string representation.
<!-- Do we have a place for the examples? -->

Some of vector driver formats could store many layers of data. 
`st_read` automatically read the first layer of the file specified in `dsn`, however it is also possible to select a different layer by its name using the `layer` argument.

Many GDAL drivers have different open options[^1]. 
For example, spatial data in text files could be stored in several ways.
Simple point dataset could just have two columns describing their x and y coordinates. 
To read these files, you need to specify names of those columns using the `options` argument:

```{r}
cycle_hire_txt = system.file("misc/cycle_hire_xy.csv", package = "spData")
cycle_hire_xy = st_read(cycle_hire_txt, options = c("X_POSSIBLE_NAMES=X",
                                                 "Y_POSSIBLE_NAMES=Y"))
```



```{r}
world_txt = system.file("misc/world_wkt.csv", package = "spData")
world_wkt = st_read(world_txt, options = "GEOM_POSSIBLE_NAMES=WKT")
```

<!-- isn't that confusing? -->
**Tip**: `read_sf()` and `write_sf()` can be used as easy-to-remember alternatives to `st_read()` and `st_write()`.
Remember they hide information about the data source and overwrite existing data, though.

A major advantage of **sf** is that it is fast.
<!-- reference to the vignette -->

To find out which data formats **sf** supports, run `st_drivers()`. Here, we show only the first two drivers:

```{r}
sf_drivers = st_drivers()
head(sf_drivers, n = 2)
```

### Raster data

Raster data also could exist in many different file formats, with some of the formats supporting multilayer files[^2].

The **raster** package support reading a single layer by `raster()` and many layers using either `brick()` or `stack()`. 
More information about the raster representation could be find in the section \@ref(raster-classes).

The `raster()` function makes it possible to read a simple single layer file: 

```{r, message=FALSE}
library(raster)
library(spDataLarge)
raster_filepath = system.file("raster/srtm.tif", package = "spDataLarge")
single_layer = raster(raster_filepath)
```

Or to read just one layer (band) from a multilayer file:

```{r}
raster_multilayer_filepath = system.file("raster/landsat.tif", package = "spDataLarge")
single_layer_band3 = raster(raster_multilayer_filepath, band = 3)
```

The whole multilayer raster file is read by `brick()`:

```{r}
multilayer_layer_brick = brick(raster_multilayer_filepath)
```

Finally, many single and multilayer rasters could be read together with the `stack()` function:

```{r}
raster_l1_filepath = system.file("raster/landsat_b2.tif", package = "spDataLarge")
raster_l2_filepath = system.file("raster/landsat_b3.tif", package = "spDataLarge")
 
multilayer_layer_stack = stack(raster_l1_filepath, raster_l2_filepath)
```

The only limitation to `stack()` is that all datasets need to have the same spatial extent and resolution.

## Data output (O)

<!-- raster + vector + text intro -->

### Vector data

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world\\.")
file.remove(world_files)
```

The counterpart of `st_read()` is `st_write()`.
It allows to write **sf** objects to a wide range of geographic vector file formats, including the most common ones such as `.geojson`, `.shp` and `.gpkg`.
Based on the file name `st_write()` decides automatically which driver to use. 
How fast the writing process is depends also on the driver.
<!-- ref to the vignette -->

```{r}
st_write(obj = world, dsn = "world.gpkg")
```

**Note**: if you try to write to the same data source again, the function will fail.
<!-- Why are you creating a modified version? Could you not use the same object again to demonstrate that overwriting will fail? -->
This is demonstrated in the code below for a modified version of the world in which the population doubles in all countries (don't worry about the **dplyr** code for now, this is covered in Chapter \@ref(attr)):

```{r}
world_mod = dplyr::mutate(world, pop = pop * 2)
```

```{r, echo=FALSE}
if (file.exists("world.gpkg")) file.remove("world.gpkg")
```

```{r, eval=FALSE}
st_write(obj = world_mod, dsn = "world.gpkg")
##   GDAL Error 1: Layer world.gpkg already exists, CreateLayer failed.
## Use the layer creation option OVERWRITE=YES to replace it.
```

The error message (only partly reproduced above) provides some information as to why the function failed.
The `GDAL Error 1` statement makes clear that the failure occurred at the GDAL level.
Additionally, the suggestion to use `OVERWRITE=YES` provides a clue how to fix the problem.
However, this is a GDAL option, and not a `st_write()` argument.
Luckily, `st_write` provides a `layer_options` argument through which we can pass driver-dependent options:

```{r, results='hide'}
st_write(obj = world_mod, dsn = "world.gpkg", layer_options = "OVERWRITE=YES")
```

Another solution is to use the `st_write()` argument `delete_layer`. Setting it to `TRUE` deletes already existing layers in the data source before the function attempts to write (note there is also a `delete_dsn` argument):

```{r, results='hide'}
st_write(obj = world_mod, dsn = "world.gpkg", delete_layer = TRUE)
```

You can achieve the same with `write_sf()` since it is equivalent to (technically an *alias* for) `st_write()`, except that its defaults for `delete_layer` and `quiet` is `TRUE`.
This enables spatial data to be overwritten more concisely, and with less output going to screen:
<!-- What does this mean: overwritten more concisely? -->

```{r, echo=FALSE, results='hide'}
if (file.exists("world.gpkg")) file.remove("world.gpkg")
```

```{r}
write_sf(obj = world_mod, dsn = "world.gpkg")
```

<!-- Is the file.remove()-part important? -->
A blunter way to update file-based geographic data sources such as `.gpkg` files is to simply delete them.
This is not generally recommended, as it will not work for multi-file data sources such as `.shp` files:

```{r, results='hide'}
file.remove("world.gpkg")
```

<!-- temporal text subsection -->

<!-- saving sf POINT object as csv -->
```{r, eval=FALSE}
library(spData)
st_write(cycle_hire, "cycle_hire_xy.csv", layer_options = "GEOMETRY=AS_XY")
st_write(world, "world_wkt.csv", layer_options = "GEOMETRY=AS_WKT")
```

### Raster data

The `writeRaster()` function save the `Raster*` objects to files. 
There are several decision to be made while saving rasters, such as an output data type, file format, and specific GDAL options related to selected file format.
The complete list of an arguments to this function could be found by typing `?writeRaster()`.

<!-- datatypes -->
The **raster** package offers nine following datatypes: LOG1S, INT1S, INT1U, INT2S, INT2U, INT4S, INT4U, FLT4S, and FLT8S.
Datatype determines the type of numbers and the range of the raster values that would be written to disk.
Decision which datatype should be used depends on the expected range of the values in the output file.
For example, LOG1S is used for bitmap (binary) images, unsigned integers (INT1U, INT2U, INT4U) are mostly used for categorical data, while floats (FLT4S and FLTS8S) usually represent continuous data.
The datatype used in `writeRaster()` is FLT4S.
While it works in most of the cases, the size of the output file would be large.
Therefore, a rule of thumb is to use the smallest representation that fits all the values (range of values could be checked using the `summary()` function).
<!-- table? -->

<!-- file formats -->

<!-- options (geotiff) -->
GeoTIFF is one of the most popular raster formats. 
Its structure is similar to the regular `.tif` format, however GeoTIFF also stores additional spatial metadata, such as coordinate reference system, spatial extent, `NoData` value, and the data resolution.
You need to add the `.tif` file extention to save the raster object to a GeoTIFF file, for example:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U")
```

Some raster file formats also accept additional creation options^[Creation options for the GeoTIFF format are at http://www.gdal.org/frmt_gtiff.html].
For example, you could specify a compress method using the `COMPRESS` option:

```{r, eval=FALSE}
writeRaster(x = single_layer,
            filename = "my_raster.tif",
            datatype = "INT2U",
            options = c("COMPRESS=DEFLATE"))
```

Other raster file formats include `.grd`, `.nc`, `.asc`, and `.img`.
Full list of the supported file format for writing `Raster*` objects could be found using `writeFormats().`

<!-- temporal text subsection -->
<!-- saving raster object as csv ?? - should we include that?-->

## File formats

### Vector formats

```{r, eval=FALSE, echo=FALSE}
vector_formats = tribble(~Name, ~Extension, ~Info, ~Model,
                         "ESRI Shapefile", ".shp (the main file)", "", "",
                         "GeoPackage", ".gpkg", "", "",
                         # "WKT/WKB"??
                         "GeoJSON", ".geojson", "", "",
                         "GPX", ".gpx", "", "",
                         "KML", ".kml", "", ""
                         )
```

<!-- 1. shp - (short history, was more or less an accident, a long-living makeshift arrangement) - don't use shp! -->
<!-- 2. gpkg - (maybe just mention here, gpkg since it is an SQLite 3 extension supporting both vector and raster formats) - reference to the "geodatabases" section -->
<!-- 3. wkt and wkb -->
<!-- 4. geojson - (web formats with a focus on geojson and here we can also mention kml, gml) -->
<!-- 5. maybe gpx - (maybe mention GPS formats such as GPX) -->

### Raster formats

```{r, eval=FALSE, echo=FALSE}
raster_formats = tribble(~Name, ~Extension, ~Info, ~Model,
                         "GeoTIFF", ".tiff", "", "",
                         "Arc ASCII", ".asc", "", "",
                         # "JPEG??"
                         "R-raster", ".gri, .grd", "", "",
                         # "NCDF??"
                         # "HDF??"
                         "GeoPackage", ".gpkg", "", ""
                         )
```


<!-- 1. geotiff -->
<!-- 2. asc - (popular formats: asc (-> exported from ESRI)) -->
<!-- 3. JPEG - (possibly mention SAGA's sdat, Erdas Imagine) -->
<!-- 4. rasterfile - (raster's rasterfile (vignette("rasterfile"))) -->
<!-- 5. gpkg - (maybe just mention here, gpkg since it is an SQLite 3 extension supporting both vector and raster formats) - reference to the "geodatabases" section -->
<!-- 6. ncdf ? -->

### Geodatabases 
<!--(can store vector and raster data)-->

```{r, eval=FALSE, echo=FALSE}
geodatabases = tribble(~Name, ~Extension, ~Info, ~Model,
                         "SQLite", ".sqlite", "", "",
                         # "WKT/WKB??",
                         "ESRI FileGDB", ".gdb", "", "",
                         "GeoPackage", ".gpkg", "", ""
                         )
```


<!-- 1. SQLite/SpatialLite + mention GRASS (uses SQLite) -->
<!-- 2. gpkg - explanation connected to the previous sections and to SQLite -->
<!-- 3. WKT/WKB for transfering and storing geometry data on databases. PostGIS (has even its own raster WKT (https://trac.osgeo.org/postgis/wiki/WKTRasterTutorial01); WKT also supported by Spatiallite, Oracle, MySQL, etc. (https://en.wikipedia.org/wiki/Well-known_text#RDBMS_Engines_that_provide_support) -->
<!-- 4. ESRI geodatabase, Oracle spatial database (mention + gdal support?) -->

## Visual outputs

<!-- aka how to save a map to a file -->

## Exercises

1. Name three differences between `write_sf()` and the more well-known function `st_write()`.

1. What are the default arguments of `read_sf()` and `write_sf()` that enable two of these differences?


[^1]: A list of supported vector formats and theirs options could be found at http://www.gdal.org/ogr_formats.html.

[^2]: Full list of supported raster formats with theirs creation options could be found at http://www.gdal.org/formats_list.html