# Geometric operations {#transform}

## Prerequisites {-}

- This chapter requires the following packages:^[**lwgeom** is also needed for a couple of more advanced transformations.]

```{r, message=FALSE}
library(raster)
library(sf)
library(tidyverse)
```

- It also relies on **spData** and **spDataLarge**, which load `cycle_hire_osm`, `nz`, `seine`, `us_states` and `world` datasets and provide external files:

```{r, message=FALSE}
library(spData)
library(spDataLarge)
```

## Introduction

The previous three chapters have demonstrated how geographic datasets are structured in R (Chapter \@ref(spatial-class)) and how to manipulate them based on their non-geographic attributes (\@ref(attr)) and spatial properties (\@ref(spatial-operations)).
This chapter goes a step further, by showing how to modify the *geometry* underlying spatial datasets.

Section \@ref(geo-vec) covers transforming vector geometries.
This includes simplication, buffers and centroids and even shifting/scaling/rotating single geometries using 'affine transformations' (see sections \@ref(simplification) to \@ref(affine-transformations)).
Vector geometries can also be modified by another spatial object with 'clipping' operations, demonstrated in section \@ref(clipping).
Geometry unions, which underlie spatial data aggregation, are covered in section \@ref(geometry-unions).
Advanced transformations of vector geometries include type transformations (from few multipolygons to many polygons, for example) and 'rasterization', operations that are demonstrated in sections \@ref(type-transformations) and \@ref(rasterization) respectively.

Section \@ref(geo-ras) covers geometric transformations on raster objects.
This involves changing the size and number of the underlying pixels, and assigning them new values.
It teaches how to change the resolution (also called raster aggregation and disaggregation), the extent and the origin of a raster.
These operations are especially useful if one would like to align raster datasets from diverse sources (see section \@ref(raster-alignment)).
Only making sure that these rasters share the same header information allows the jointly usage of map algebra operations (see section \@ref(map-algebra)). 

A vital type of geometry transformation is *reprojecting* from one coordinate reference system (CRS) to another.
Because of the importance of reprojection, introduced in Chapter \@ref(spatial-class), and the fact that it applies to raster and vector geometries alike, it is the topic of the first section in this chapter.

## Reprojecting geographic data {#reproj-geo-data}

<!-- So far this intro mainly shows examples how to reproject vector data, and a few exemplarily pitfalls.
And though the pitfall examples are good, I wonder if we can keep them a bit more general (so far only related to vector data) or if at times a coding example is necessary, maybe only a figure would be sufficient.
We should especially point out that since we project from 3D to 2D we can only keep two out of three spherical (real-world) properties: angle, length, area (-> could be also part of which CRS to use or even better we refer to this explanation in "which CRS to use").
So a specific use case determines which projection to use. 
In sea navigation we need angles and lengths, real estate is especially interested in correct areas, etc.

Hence, maybe we should move parts from c2 to c5 since the stuff in c2 is already rather specific, and not a short intro to CRS anymore. 
Another solution would be to leave the intro crs stuff in c2 (extended by when to use which CRS) and in c5 we don't have a joint vector-raster projection section but have a repojection section in the vector and raster section.
(In the case of raster this would make perfectly sense since for reprojecting one needs all the geometric operations in section raster alignment (warp, shift, scale, aggregate, etc.).
Maybe (not sure) a few figures would also help (conic, azimuthal projection; ellipsoids, etc.)
-->

Section \@ref(crs-intro) introduced coordinate reference systems (CRSs) and demonstrated their importance for geocomputation.
This section goes further, by demonstrating some problems that can arise when using an inappropriate CRS and how to *transform* geometries from one CRS to another.

Many spatial operations assume that you are using a *projected* CRS.
The GEOS engine underlying most spatial operations in **sf**, for example, assumes your data is in a projected CRS.
For this reason **sf** contains a function for checking if geometries have a geographic or projected CRS.
This is illustrated below using the example of London introduced in section \@ref(vector-data), which is created by *coercing* a `data.frame` into an `sf` object (the `coords` argument specifies the coordinates):

```{r}
london = data.frame(lon = -0.1, lat = 51.5) %>% 
  st_as_sf(coords = c("lon", "lat"))
st_is_longlat(london)
```

The results show that when geographic data is created from scratch, or is loaded from a source that has no CRS metadata, the CRS is unspecified by default.
The CRS can be set with `st_set_crs()`:^[The CRS can also be added when creating `sf` objects with the `crs` argument (e.g. `st_sf(geometry = st_sfc(st_point(c(-0.1, 51.5))), crs = 4326)`).
The same argument can also be used to set the CRS when creating raster datasets (e.g. `raster(crs = "+proj=longlat")`).]

```{r}
london = st_set_crs(london, 4326)
st_is_longlat(london)
```

Spatial operations on objects without a CRS run on the implicit assumption that they are projected, even when in reality they are not.
This can lead to problems, as illustrated by the following code chunk, which creates a buffer of one degree around `london`:

```{r}
london_buff = st_buffer(london, dist = 1)
```

Note the warning that informs us that the result has limited use because distances in geographic CRSs are in degrees, rather than meters or some other suitable measure of distance.
The consequences of a failure to work on projected data are illustrated in Figure \@ref(fig:crs-buf) (left panel):
note how the buffer is elongated in the north-south direction.
This because lines of longitude are converge towards the Earth's poles making them closer together (lines of latitude by contrast have constant distance from each other).  

Do not interpret the warning about the geographic (`longitude/latitude`) CRS as "the CRS should not be set": it almost always should be!
It is better understood as a suggestion to *reproject* the data onto a projected CRS.
This suggestion does not always need to be heeded: performing spatial and geometric operations makes little or no difference in some cases (e.g., spatial subsetting).
But for operations involving distances such as buffering, the only way to ensure a good result is to create a projected copy of the data and run the operation on that.
This is done in the code chunk below:

```{r}
london_proj = data.frame(x = 530000, y = 180000) %>% 
  st_as_sf(coords = 1:2, crs = 27700)
```

The result is a new object that is identical to `london`, but reprojected onto a suitable CRS (the British National Grid, which has an EPSG code of 27700 in this case) that has units of meters. 
We can verify that the CRS has changed using `st_crs()` as follows (some of the output has been replace by `...`):

```{r, eval=FALSE}
st_crs(london_proj)
#> Coordinate Reference System:
#>   EPSG: 27700 
#>   proj4string: "+proj=tmerc +lat_0=49 +lon_0=-2 ... +units=m +no_defs"
```

Notable components of this CRS description include the EPSG code (`EPSG: 27700`), the origin (`+lat_0=49 +lon_0=-2`) and units (`+units=m`).
The fact that the units of the CRS are meters (rather than degrees) tells us that this is a projected CRS: geometry operations on `london_proj` will work without a warning, meaning buffers can be produced from it using proper units of distance.
Moving 1 degree means moving 100+ km (111,320 meters) at the equator.
This is used as the new buffer distance:

```{r}
london_proj_buff = st_buffer(london_proj, 111320)
```

The result in Figure \@ref(fig:crs-buf) (right panel) shows that buffers based on a projected CRS are not distorted:
every part of the buffer's border is equidistant to London.

```{r crs-buf, fig.cap="Buffer on vector represenations of London with a geographic (left) and projected (right) CRS. The circular point represents London and the grey outline represents the outline of the UK.", fig.asp=1, fig.show='hold', out.width="45%", echo=FALSE}
uk = rnaturalearth::ne_countries(scale = 50) %>% 
  st_as_sf() %>% 
  filter(grepl(pattern = "United Kingdom|Ire", x = name_long))
plot(london_buff, graticule = st_crs(4326), axes = TRUE)
plot(london, add = TRUE)
plot(uk$geometry, add = TRUE, border = "grey")
uk_proj = uk %>%
  st_transform(27700)
plot(london_proj_buff, graticule = st_crs(27700), axes = TRUE)
plot(london_proj, add = TRUE)
plot(uk_proj$geometry, add = TRUE, border = "grey")
```

The importance of CRSs (primarily whether they are projected or geographic) has been demonstrated using the example of London.
The subsequent sections go into more depth, exploring which CRS to use and the details of reprojecting vector and raster objects.

### Which CRS to use?

While CRSs can be set manually --- as illustrated in the previous section with `st_set_crs(london, 4326)` --- it is more common in real world applications for CRSs to be set automatically when data is read-in.
The main task involving CRSs is often to *transform* objects provided in one CRS into another.
But when should data be transformed? And into which CRS?
There are no clear-cut answers to these questions and CRS selection always involves tradeoffs [@maling_coordinate_1992].
However there are some general principles, provided in this section, that can help decide. 

The question of *when to transform* is easier to answer.
Transformation to a projected CRS may be vital, for instance when planned geometric operations --- those involving distance measurements or area calculations, for example --- require meaningful units of distance by the object has be provided with a geographic CRS.
Conversely if the outputs of a project are to be published in an on-line map, it may be necessary to convert them to a geographic CRS.
If the visualization phase of a project involves publishing results using [leaflet](https://github.com/Leaflet/Leaflet) via the common format [GeoJSON](http://geojson.org/) (a common scenario) projected data should probably be transformed to WGS84. 
Another case is when two objects with different CRSs must be compared or combined: performing a geometric operation on two objects with different CRSs results in an error.
This is demonstrated in the code chunk below, which attempts to find the distance between the projected and unprojected versions of `london`:

```{r, eval=FALSE}
st_distance(london, london_proj)
# > Error: st_crs(x) == st_crs(y) is not TRUE
```

To make the `london` and `london_proj` objects geographically comparable one of them must be transformed into the CRS of the other.
But which CRS to use?
The answer is usually 'to the projected CRS', which in this case is the British National Grid (BNG, EPSG:27700):

```{r}
london2 = st_transform(london, 27700)
```

Now that a transformed version of `london` has been created, using the **sf** function `st_transform()`, the distance between the two representations of London can be found.
It may come as a surprise that `london` and `london2` are just over 2 km apart!^[
The difference in location between the two points is not due to imperfections in the transforming operation (which is in fact very accurate) but the low precision of the manually-created coordinates that created `london` and `london_proj`.
Also surprising may be that the result is provided in a matrix with units of meters.
This is because `st_distance()` can provide distances between many features and because the CRS has units of meters.
Use `as.numeric()` to coerce the result into a regular number.]

```{r}
st_distance(london2, london_proj)
```

The question of *which CRS* is tricky, and there is often no 'right' answer:
"There exist no all-purpose projections, all involve distortion when far from the centre of the specified frame" [@bivand_applied_2013].
For geographic CRSs the answer is often [WGS84](https://en.wikipedia.org/wiki/World_Geodetic_System#A_new_World_Geodetic_System:_WGS_84), not only for web mapping (covered in the previous paragraph) but also because GPS datasets and thousands of raster and vector datasets are provided in this CRS by default.
WGS84 is the most common CRS in the world so it is worth knowing it's EPSG code: 4326.
This 'magic number' can be used to convert objects with unusual projected CRSs something that is widely understood.

What about when a projected CRS is required?
In some cases it is not something that we are free to decide:
"often the choice of projection is made by a public mapping agency" [@bivand_applied_2013].
This means that when working with local data sources, it is likely preferable to work with the CRS in which the data was provided, to ensure compatibility, even if the 'official' CRS is not the most accurate.
The example of London was easy to answer because a) the CRS 'BNG' (with its associated EPSG code 27700) is well-known and b) the original dataset (`london`) already had that CRS.

What about when a projected CRS is needed but the study region lacks a well-known CRS?
Again, although there is no universal answer there is a sensible default CRS: Universal Transverse Mercator ([UTM](https://en.wikipedia.org/wiki/Universal_Transverse_Mercator_coordinate_system)).
UTM is not actually a single CRS but a system of CRSs that covers the entire world, and breaks it into 60 segments, each containing 6 lines of longitude.
All UTM projections have the same datum (WGS84) and their EPSG codes run sequentially from 32601 to 32660.
This makes it possible to create a function (we'll call it `lonlat2UTM`) to calculate the EPSG code associated with any point on the planet as follows:^[
Thanks to Josh O'Brien who provided the basis for this function in an answer to a question on [stackoverflow](https://stackoverflow.com/questions/9186496).
] 

<!-- Idea: create full function with message and flexibility in later chapter (RL) -->
```{r}
lonlat2UTM = function(lonlat) {
  utm = (floor((lonlat[1] + 180) / 6) %% 60) + 1
  utm + 32600
}
```

Let's use this function to identify the UTM zone and associated EPSG code for London:

```{r}
epsg_utm = lonlat2UTM(st_coordinates(london))
st_crs(epsg_utm)
```

As expected by viewing a map of UTM zones (such as that provided by [dmap.co.uk](http://www.dmap.co.uk/utmworld.htm)), the EPSG code returned refers to UTM zone 30, which would represent a good projected CRS for England if the BNG did not exist.
<!-- London can be transformed into this CRS as follows (result not shown): -->

<!-- ```{r} -->
<!-- lnd_utm = st_transform(london, crs = epsg_utm) -->
<!-- ``` -->

Another approach to automatically selecting a projected CRS specific to a local dataset is to create an azimuthal equidistant ([AEQD](https://en.wikipedia.org/wiki/Azimuthal_equidistant_projection)) projection for the centre-point of the study area.
This involves creating a custom CRS (with no EPSG code) with units of meters based on the centrepoint of a dataset.
This approach should be used with caution: no other datasets will be compatible with the custom CRS created and results may not be accurate when used on extensive datasets covering hundreds of kilomtres.

Although we used vector datasets to illustrate the points outlined in this section, the principles apply equally to raster datasets.
The subsequent sections explain features of CRS transformation that are unique to each geographic data model, continuing with vector data in section \@ref(reprojecting-vector-geometries) and moving-on to explain how raster transformation is different, in section \@ref(reprojecting-raster-geometries).

<!-- This approach is used in the **stplanr** function `geo_select_crs()` which returns a CRS object that can be used in other functions (see `?stplanr::geo_select_aeq` for further details): -->

<!-- ```{r} -->
<!-- stplanr::geo_select_aeq(london) -->
<!-- ``` -->

<!-- Another **stplanr** function, `geo_buffer()`, uses this behind the scenes to enable buffers to be created around objects with geographic CRSs with units of metres, and returns the result in the original CRS, as illustrated in the code chunk below: -->

<!-- ```{r} -->
<!-- london_proj_buff2 = stplanr::geo_buffer(london, dist = 111320) -->
<!-- ``` -->

<!-- ```{r, eval=FALSE, echo=FALSE} -->
<!-- library(tmap) -->
<!-- tmap_mode("view") -->
<!-- qtm(st_transform(london_proj_buff, 4326)) + -->
<!--   qtm(london_proj_buff2, "red") + -->
<!--   qtm(london_buff) -->
<!-- ``` -->


### Reprojecting vector geometries

Chapter \@ref(spatial-class) demonstrated how vector geometries are made-up of points, which combine to create more complex objects such as lines and polygons.
Reprojecting vectors thus consists of transforming the coordinates of these points.

<!-- Depending on projections used, reprojection could be either lossy or lossless. -->
<!-- I don't understand the following sentence -->
<!-- For example, loss of spatial information could occur when the new CRS is only adequate for smaller area than input vector. -->
<!-- Do you have an example for the next sentence? -->
<!-- The precision could be also lost when transforming coordinate systems with different datums - in those situations approximations are used. -->
<!-- However, in most cases CRS vector transformation is lossless. -->

To illustrate this point we'll use the dataset `cycle_hire_osm` from **spData**, representing cycle hire locations across London.
Its CRS can be queried as follows:

```{r}
st_crs(cycle_hire_osm)
```

CRSs in the **sf** package be using their `epsg` code or a `proj4string` definition (see section \@ref(crs-in-r)).

Let's create a new version of `cycle_hire_osm` in a projected CRS, using the `epsg` number of 27700:

```{r}
cycle_hire_osm_projected = st_transform(cycle_hire_osm, 27700)
st_crs(cycle_hire_osm_projected)
```

Note that the result shows that the `epsg` has been updated and that `proj4string` element of the CRS now contains, among other things `+proj=tmerc` (meaning it is a projected CRS using the [tranverse Mercator](https://en.wikipedia.org/wiki/Transverse_Mercator_projection) projection) and `+units=m` (meaning the units of the coordinates are meters).
Another function, from the **rgdal** library, provides a note containing the name of the CRS:

```{r}
crs_codes = rgdal::make_EPSG()[1:2]
dplyr::filter(crs_codes, code == 27700)
```

The result shows that the EPSG code 27700 represents the British National Grid, a result that could have been found by searching online for "[CRS 27700](https://www.google.com/search?q=CRS+27700)".
The formula that converts a geographic point into a point on the surface of the Earth is provided by the `proj4string` element of the `crs` (see [proj4.org](http://proj4.org/) for further details):

```{r}
st_crs(27700)$proj4string
```

```{block2 type='rmdnote'}
The EPSG code can be found inside the `crs` attribute of the object's geometry.
It is hidden from view for most of the time except when the object is printed but can be can identified and set using the `st_crs` function, for example `st_crs(cycle_hire_osm)$epsg`.
```

Existing CRS are well suited for most purposes.
<!-- examples -->
In the same time, `proj4string` definitions are highly modifiable and allow for CRS customization.
<!-- as we mentioned in section \@ref(crs-in-r). -->
We can present that using selected world projections.
The Mollweide projection is recommended when it is important to preserve areas [@jenny_guide_2017] (Figure \@ref(fig:mollproj)).
To use this projection, we need to specify it using the `proj4string` element, `"+proj=moll"`, in the `st_transform` function:

```{r}
world_mollweide = st_transform(world, crs = "+proj=moll")
```
<!-- plot(world_mollweide$geom) -->
<!-- plot(world_mollweide$geom, graticule = TRUE) -->

```{r mollproj, echo=FALSE, fig.cap="Mollweide projection of the world", warning=FALSE}
par_old = par()
par(mar = c(0, 0, 1, 0))
plot(world_mollweide$geom, graticule = TRUE, main = "the Mollweide projection")
par(par_old)
```

On the other hand, the goal for many visualization purposes is to have a map with minimized area, direction, and distance distortions.
One of the most popular projection to achieve that is the Winkel tripel projection (Figure \@ref(fig:wintriproj)).^[This projection is used, among others, by the National Geographic Society.]
`st_transform_proj()` from the **lwgeom** package allows for coordinate transformations to a wide range of CRSs, inluding the Winkel tripel projection: 

```{r}
world_wintri = lwgeom::st_transform_proj(world, crs = "+proj=wintri")
```
<!-- plot(world_wintri$geom) -->
<!-- plot(world_wintri$geom, graticule = TRUE) -->

```{r wintriproj, echo=FALSE, fig.cap="Winkel tripel projection of the world", warning=FALSE}
world_wintri_gr = st_graticule(lat = c(-89.9, seq(-80, 80, 20), 89.9)) %>% 
  lwgeom::st_transform_proj(crs = "+proj=wintri")
par_old = par()
par(mar = c(0, 0, 1, 0))
plot(world_wintri_gr$geometry, main = "the Winkel tripel projection", col = "grey")
plot(world_wintri$geom, add = TRUE)
par(par_old)
```

```{block2 type='rmdnote'}
Two main functions for transformation of simple features coordinates are `sf::st_transform()` and `sf::sf_project()`. 
The `st_transform` function uses the GDAL interface to PROJ.4, while `sf_project()` (which works with two-column numeric matrices, representing points) and `lwgeom::st_transform_proj()` use the PROJ.4 API directly.
The first one is appropriate in most situations, and provides a set of the most often used parameters and well defined transformations.
The second one allows for a greater customization of a projection, which includes cases when some of the PROJ.4 parameters (e.g. `+over`) or projection (`+proj=wintri`) is not available in `st_transform()`.
```

```{r, eval=FALSE, echo=FALSE}
# demo of sf_project
mat_lonlat = as.matrix(data.frame(x = 0:20, y = 50:70))
plot(mat_lonlat)
mat_projected = sf_project(from = st_crs(4326)$proj4string, to = st_crs(27700)$proj4string, pts = mat_lonlat)
plot(mat_projected)
```

Moreover, PROJ.4 parameters can be modified in most CRS definitions.
The below code transforms the coordinates to the Lambert azimuthal equal-area projection centered on longitude and latitude of `0` (Figure \@ref(fig:laeaproj1)).

```{r}
world_laea1 = st_transform(world, crs = "+proj=laea +x_0=0 +y_0=0 +lon_0=0 +lat_0=0")
```
<!-- plot(world_laea1$geom) -->
<!-- plot(world_laea1$geom, graticule = TRUE) -->

```{r laeaproj1, echo=FALSE, fig.cap="Lambert azimuthal equal-area projection of the world centered on longitude and latitude of 0", warning=FALSE}
par_old = par()
par(mar = c(0, 0, 1, 0))
plot(world_laea1$geom, graticule = TRUE, main = "the Lambert azimuthal equal-area projection")
par(par_old)
```

We can change the PROJ.4 parameters, for example the center of the projection using the `+lon_0` and `+lat_0` parameters. 
The code below gives the map centered on New York City (Figure \@ref(fig:laeaproj2)).

```{r}
world_laea2 = st_transform(world, crs = "+proj=laea +x_0=0 +y_0=0 +lon_0=-74 +lat_0=40")
```
<!-- plot(world_laea2$geom) -->
<!-- plot(world_laea2$geom, graticule = TRUE) -->

```{r laeaproj2, echo=FALSE, fig.cap="Lambert azimuthal equal-area projection of the world centered on New York City", warning=FALSE}
par_old = par()
par(mar = c(0, 0, 1, 0))
plot(world_laea2$geom, graticule = TRUE, main = "the Lambert azimuthal equal-area projection")
par(par_old)
```

More information on CRS modifications can be found in the [Using PROJ.4](http://proj4.org/usage/index.html) documentation.

<!-- https://github.com/r-spatial/lwgeom/issues/6 -->
<!-- ```{r} -->
<!-- # devtools::install_github("r-spatial/lwgeom") -->
<!-- library(lwgeom) -->
<!-- world_3 = lwgeom::st_transform_proj(world, crs = "+proj=wintri") -->
<!-- plot(world_3$geom) -->
<!-- ``` -->
<!-- http://bl.ocks.org/vlandham/raw/9216751/ -->

### Reprojecting raster geometries

The projection concepts described in the previous section apply equally to rasters.
However, there are important differences in reprojection of vectors and rasters:
transforming a vector object involves changing the coordinates of every vertex but this does not apply to raster data.
Rasters are composed of rectangular cells of the same size (expressed by map units, such as degrees or meters), so it is impossible to transform coordinates of pixels separately.

Raster reprojection involves creating a new raster object, often with a different number of columns and rows than the original.
The attributes must subsequently be re-estimated, allowing the new pixels to be 'filled' with appropriate values.
This two-stage process is done with `projectRaster()` from the **raster** package.
Like the `st_transform()` function demonstrated in the previous section, `projectRaster()` takes a geographic object (a raster dataset in this case) and a `crs` argument.
However, `projectRaster()` only accepts the lengthy `proj4string` definitions of a CRS rather than concise EPSG codes.

```{block2 type='rmdnote'}
It is possible to use a EPSG code in a `proj4string` definition with `"+init=epsg:MY_NUMBER"`.
For example, one can use the `"+init=epsg:4326"` definition to set CRS to WGS84 (EPSG code of 4326).
The PROJ.4 library automatically adds the rest of parameters and converts it into `"+init=epsg:4326 +proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"`,
```

Let's take a look at two examples of raster transformation - using categorical and continuous data.
Land cover data are usually represented by categorical maps.
The `nlcd2011.tif` file provides information for a small area in Utah, USA obtained from [National Land Cover Database 2011](https://www.mrlc.gov/nlcd2011.php) in the NAD83 / UTM zone 12N CRS.

```{r}
cat_raster = raster(system.file("raster/nlcd2011.tif", package = "spDataLarge"))
cat_raster
```

In this region, 14 land cover classes were distinguished^[Full list of NLCD2011 land cover classes can be found at https://www.mrlc.gov/nlcd11_leg.php]:

```{r}
unique(cat_raster)
```

When reprojecting categorical raster, we need to ensure that our new estimated values would still have values of our original classes.
This could be done using the nearest neighbor method (`ngb`).
In this method, value of the output cell is calculated based on the nearest cell center of the input raster.

For example, we want to change the CRS to WGS 84.  
It can be desired when we want to visualize a raster data on top of a web basemaps, such as the Google or OpenStreetMap map tiles.
The first step is to obtain the proj4 definition of this CRS, which can be done using the [http://spatialreference.org](http://spatialreference.org/ref/epsg/wgs-84/) webpage. 
The second and last step is to define the reprojection method in the `projectRaster()` function, which in case of categorical data is the nearest neighbor method (`ngb`):

```{r}
wgs84 = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
cat_raster_wgs84 = projectRaster(cat_raster, crs = wgs84, method = "ngb")
cat_raster_wgs84
```

Many properties of the new object differs from the previous one, which include the number of columns and rows (and therefore number of cells), resolution (transformed from meters into degrees), and extent.
In the same time, it keeps the same land cover classes - `unique(cat_raster_wgs84)`.
<!-- freq(cat_raster_wgs84) -->
<!-- freq(cat_raster) -->

This process of reprojection is almost identical for continuous data.
The `srtm.tif` file contains digital elevation model for the same area in Utah from [the Shuttle Radar Topography Mission (SRTM)](https://www2.jpl.nasa.gov/srtm/).
Each value in this raster represents elevation measured in meters.

```{r}
con_raster = raster(system.file("raster/srtm.tif", package = "spDataLarge"))
con_raster
```

This dataset has a geographic CRS and we want to transform it into a projected CRS.
The nearest neighbor method should not be used for continuous raster data, as we want to preserve gradual changes in values.
Instead we will use the bilinear method which computes the output cell value based on the four nearest cells in the original raster. 
<!-- Quadric and cubic polynomials are also popular nterpolation functions for resampling with more complexity and improved accuracy. @liu_essential_2009: 111-->
The new value is the distance-weighted average of the values from these four cells, i.e., the closer the input cell is to the center of the output cell, the stronger is its weight.

<!-- I am not really sure what the following note wants to tell the reader. -->
```{block2 type='rmdnote'}
All the grid cells in equal-area projections have the same size.
Therefore, these projections are recommended when performing many raster operations, such as distance calculations.
```

<!-- unclear what is meant -->
In the fist step we need to obtain the proj4 definition of the existing projected CRS appropriate for this area or create a new one using the [Projection Wizard](http://projectionwizard.org/) online tool [@savric_projection_2016].
For this example, we used the Oblique Lambert azimuthal equal-area projection.
The second step is to define the `bilinear` reprojection method:

```{r}
equalarea = "+proj=laea +lat_0=37.32 +lon_0=-113.04"
con_raster_ea = projectRaster(con_raster, crs = equalarea, method = "bilinear")
con_raster_ea
```

<!-- we have said so before, so maybe we can delete? -->
Reprojecting (continuous) rasters also changes spatial properties, such as the number of cells, resolution, and extent.
Moreover, it slightly modifies values in the new raster, which can be seen by comparing the outputs of the `summary()` function between `con_raster` and `con_raster_ea`.

```{r, eval=FALSE}
summary(con_raster)
summary(con_raster_ea)
```

<!-- why new na? -->

<!-- res option in projectRaster? -->
<!-- note1: in most of the cases reproject vector, not raster-->
<!-- note2: equal area projections are the best for raster calculations -->
<!-- q: should we mentioned gdal_transform? -->

## Geometric operations on vector data {#geo-vec}

This section is about operations that in some way change the geometry of vector (`sf`) objects.
It is more advanced than the spatial data operations presented in the previous Chapter (in section \@ref(spatial-vec)) because here we drill down into the geometry:
the functions discussed in this section work on objects of class `sfc` (simple feature geometry collections) in addition to objects of class `sf`.

### Simplification

Simplification is a process for generalization of vector objects (lines and polygons) usually for its use in smaller scale maps.
Additional reason for simplification is reduction of the object size and therefore the size of a saved file.
Therefore, data is often simplified before its use in interactive maps. 
The **sf** package provides the `st_simplify()` function^[It uses the simplify algorithm from GEOS.]
Level of generalization is controlled by the `dTolerance` argument, which expects a value in map units.
As the result, all vertices in the simplified geometry will be within this value from the original ones.

We can use data of the Seine, Marne and Yonne rivers (on the left in Figure \@ref(fig:seine-simp)) as an example of line vector data simplification:

```{r}
seine_simp = st_simplify(seine, dTolerance = 2000) #2000 m
```

```{r seine-simp, echo=FALSE, fig.cap="Comparison of the original data of the Seine, Marne and Yonne rivers and its simplified version using `st_simplify`.", warning=FALSE}
par_old = par()
par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
plot(seine$geometry, col = 1, main = "Original data")
plot(seine_simp$geometry, col = 1, main = "st_simplify")
par(par_old)
```

The new object, `seine_simp`, is not only visually generalized (on the right in Figure \@ref(fig:seine-simp)), but also has twice smaller size than the original object:

```{r}
object.size(seine)
object.size(seine_simp)
```

Let's take a look at the second example of simplification.
It will use, `us_states`, polygons representing the contiguous United States.
As we showed in section \@ref(reproj-geo-data), GEOS assumes that the data is in a projected CRS and this could lead to unexpected results when using a geographic CRS.
Therefore, the first step is to project the data into some adequate projected CRS, such as US National Atlas Equal Area (epsg = 2163) (on the left in Figure \@ref(fig:us-simp)):

```{r}
us_states2163 = st_transform(us_states, 2163)
```

The `st_simplify` function uses the standard Douglas-Peucker algorithm, which is realatively fast.

```{r}
us_states_simp1 = st_simplify(us_states2163, dTolerance = 100000) #100 km
```

On the other hand, this algorithm simplify objects on a per-geometry basis, and therefore does not preserve topology.
This could be seen on the middle map in Figure \@ref(fig:us-simp), where polygons of single states are splitted creating holes or overlapping each other.
An alternative simplification method can be used to solve this issue, such as the Visvalingam algorithm implemented in the **rmapshaper** package's function `ms_simplify()`.
<!-- https://bost.ocks.org/mike/simplify/ -->
In the below example, we would simplified the `us_states2163` object using only 0.1% of original data vertices (argument `keep`), and also assure that all of the original objects (states) would be be kept (argument `keep_shapes`).

```{r}
# proportion of points to retain (0-1; default 0.05)
us_states_simp2 = rmapshaper::ms_simplify(us_states2163, keep = 0.005,
                                          keep_shapes = TRUE)
```

Finally, the visual comparison of the original dataset and two simplified versions shows differences between the Douglas-Peucker (`st_simplify`) and Visvalingam (`ms_simplify`) algorithms outputs (Figure \@ref(fig:us-simp)):

```{r us-simp, echo=FALSE, fig.cap="Comparison of the original data of the contiguous United States and two simplified versions using `st_simplify` and `ms_simplify`.", warning=FALSE}
par_old = par()
par(mfrow = c(1, 3), mar = c(0, 0, 1, 0))
plot(us_states2163$geometry, col = 8, main = "Original data")
plot(us_states_simp1$geometry, col = 8, main = "st_simplify")
plot(us_states_simp2$geometry, col = 8, main = "ms_simplify")
par(par_old)
```

### Centroids and buffers

<!-- centroids intro -->
There are two main functions that create single point representations of more complex vector objects - `st_centroid()` and `st_point_on_surface()`.

The `st_centroid()` function calculates the geometric center of a geometry.
We can create centroids for polygons, lines (see black points on Figure \@ref(fig:centr)) and multipoints:

```{r}
nz_centroid = st_centroid(nz)
seine_centroid = st_centroid(seine)
```

Centroids could be useful to represent more complex objects - lines and polygons, for example to calculate distances between centers of polygons.
They are also often used as places where polygons or lines labels are put. 
However, it is important to know that centroids could be located outside of the given object, e.g. in cases of irregular shaped polygons or lines.
Examples of this could be seen on the right plot on Figure \@ref(fig:centr).

Alternatively, the `st_point_on_surface()` can be used.

```{r}
nz_pos = st_point_on_surface(nz)
seine_pos = st_point_on_surface(seine)
```

It ensures that the created point lies on the given object (see red points on Figure \@ref(fig:centr)).

```{r centr, warning=FALSE, echo=FALSE, fig.cap="Comparison between the outputs of `st_centroid()` (black points) and `st_point_on_surface()` (red points) on New Zeleand's regions (left) and the Seine, Marne and Yonne rivers (right)."}
par_old = par()
par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
plot(nz$geometry)
plot(nz_centroid$geometry, add = TRUE)
plot(nz_pos$geometry, add = TRUE, col = "red")
plot(seine$geometry)
plot(seine_centroid$geometry, add = TRUE)
plot(seine_pos$geometry, add = TRUE, col = "red")
par(par_old)
```

### Affine transformations

### Clipping 

Spatial clipping is a form of spatial subsetting that involves changes to the `geometry` columns of at least some of the affected features.

Clipping can only apply to features more complex than points: 
lines, polygons and their 'multi' equivalents.
To illustrate the concept we will start with a simple example:
two overlapping circles with a center point one unit away from each other and radius of one:

```{r points, fig.cap="Overlapping circles."}
b = st_sfc(st_point(c(0, 1)), st_point(c(1, 1))) # create 2 points
b = st_buffer(b, dist = 1) # convert points to circles
l = c("x", "y")
plot(b)
text(x = c(-0.5, 1.5), y = 1, labels = l) # add text
```

Imagine you want to select not one circle or the other, but the space covered by both `x` *and* `y`.
This can be done using the function `st_intersection()`, illustrated using objects named `x` and `y` which represent the left and right-hand circles:

```{r}
x = b[1]
y = b[2]
x_and_y = st_intersection(x, y)
plot(b)
plot(x_and_y, col = "lightgrey", add = TRUE) # color intersecting area
```

The subsequent code chunk demonstrate how this works for all combinations of the 'Venn' diagram representing `x` and `y`, inspired by [Figure 5.1](http://r4ds.had.co.nz/transform.html#logical-operators) of the book R for Data Science [@grolemund_r_2016].
<!-- Todo: reference r4ds -->

```{r venn-clip, echo=FALSE, fig.cap="Spatial equivalents of logical operators.", warning=FALSE}
old_par = par()
par(mfrow = c(3, 3), mai = c(0.1, 0.1, 0.1, 0.1))
plot(b)
y_not_x = st_difference(y, x)
plot(y_not_x, col = "grey", add = TRUE)
text(x = 0.5, y = 1, "st_difference(y, x)")
plot(b)
plot(x, add = TRUE, col = "grey")
text(x = 0.5, y = 1, "x")
plot(b, add = TRUE)
x_or_y = st_union(x, y)
plot(x_or_y, col = "grey")
text(x = 0.5, y = 1, "st_union(x, y)")
x_and_y = st_intersection(x, y)
plot(b)
plot(x_and_y, col = "grey", add = TRUE) 
text(x = 0.5, y = 1, "st_intersection(x, y)")
# x_xor_y = st_difference(x_xor_y, x_and_y) # failing
x_not_y = st_difference(x, y)
x_xor_y = st_sym_difference(x, y)
plot(x_xor_y, col = "grey")
text(x = 0.5, y = 1, "st_sym_difference(x, y)")
plot.new()
plot(b)
plot(x_not_y, col = "grey", add = TRUE)
text(x = 0.5, y = 1, "st_difference(x, y)")
plot(b)
plot(y, col = "grey", add = TRUE)
plot(b, add = TRUE)
text(x = 0.5, y = 1, "y")
par(old_par)
```

To illustrate the relationship between subsetting and clipping spatial data, we will subset points that cover the bounding box of the circles `x` and `y` in Figure \@ref(fig:venn-clip).
Some points will be inside just one circle, some will be inside both and some will be inside neither.

There are two different ways to subset points that fit into combinations of the circles: via clipping and logical operators.
But first we must generate some points.
We will use the *simple random* sampling strategy to sample from a box representing the extent of `x` and `y`.
To generate this points will use a function not yet covered in this book, `st_sample()`.
Next we will generate the situation plotted in Figure \@ref(fig:venn-subset):

```{r venn-subset, fig.cap="Randomly distributed points within the bounding box enclosing circles x and y."}
bb = st_bbox(st_union(x, y))
pmat = matrix(c(bb[c(1, 2, 3, 2, 3, 4, 1, 4, 1, 2)]), ncol = 2, byrow = TRUE)
box = st_polygon(list(pmat))
set.seed(2017)
p = st_sample(x = box, size = 10)
plot(box)
plot(x, add = TRUE)
plot(y, add = TRUE)
plot(p, add = TRUE)
text(x = c(-0.5, 1.5), y = 1, labels = l)
```

```{r, echo=FALSE}
# An alternative way to sample from the bb
bb = st_bbox(st_union(x, y))
pmulti = st_multipoint(pmat)
box = st_convex_hull(pmulti)
```

### Geometry unions

Spatial aggregation can also be done in the **tidyverse**, using **dplyr** functions as follows:

```{r, eval = FALSE}
group_by(us_states, REGION) %>%
  summarize(sum(pop = total_pop_15, na.rm = TRUE))
```

For attribute data aggregation the grouping variable is another variable, typically one with few unique values relative to the number of rows (see section \@ref(vector-attribute-aggregation)).
What we did not cover in that section was that attribute data aggregation dissolves the geometries of touching polygons.
The `REGION` variable in the `us_states` dataset is a good example:
there are 49 states (excluding Hawaii and Alaska) which can be aggregated into four regions.
This is demonstrated in the code chunk below, the results of which are illustrated in Figure \@ref(fig:us-regions):

```{r}
regions = aggregate(x = us_states[, "total_pop_15"], by = list(us_states$REGION),
                    FUN = sum, na.rm = TRUE)
```
<!--
show also tidyverse way, so what you are doing is basically a spatial join and a subsequent aggregation without a grouping variable. Didactically, it might be better to present a grouping variable.
-->

```{r, echo=FALSE}
# st_join(buff, africa[, "pop"]) %>%
#   summarize(pop = sum(pop, na.rm = TRUE))
# summarize(africa[buff, "pop"], pop = sum(pop, na.rm = TRUE))
```

```{r us-regions, fig.cap="Spatial aggregation on contiguous polygons, illustrated by aggregating the population of US states into regions, with population represented by color. Note the operation automatically dissolves boundaries between states.", echo=FALSE, warning=FALSE, fig.asp=0.2, out.width="100%"}
# # with base R ----
# old_par = par()
# breaks = c(1e5, 5e6, 1e7, 7e7, 2e8)
# rdf = st_set_geometry(regions, NULL)
# us_states$region_pop = inner_join(dplyr::select(us_states, REGION),
#                                   dplyr::select(rdf, Group.1, total_pop_15),
#                                   by = c("REGION" = "Group.1")) %>% 
#   pull(total_pop_15)
# 
# par(mfrow = c(1, 2))
# plot(us_states[, "total_pop_15"], main = "US states", breaks = breaks, key.pos = NULL)
# plot(regions[, "total_pop_15"], main = "US regions", breaks = breaks, key.pos = NULL)
# par(old_par)
# with tmap ----
library(tmap)
us_states_facet = dplyr::select(us_states, REGION, total_pop_15) %>%
  mutate(Level = "State")
regions_facet = rename(regions, REGION = Group.1) %>%
  mutate(Level = "Region")
us_facet = rbind(us_states_facet, regions_facet) %>% 
  mutate(Level = factor(Level, levels = c("State", "Region")))

qtm(us_facet, "total_pop_15") +
  tm_facets(by = "Level", nrow = 1, drop.units = TRUE)
```

The equivalent result can be achieved using **tidyverse** functions as follows (result not shown):

```{r, eval=FALSE}
regions2 = us_states %>% 
  group_by(REGION) %>%
  summarize(sum(pop = total_pop_15, na.rm = TRUE))
```

### Type transformations

Geometry casting is powerful operation which enable transformation of the geometry type.
It is implemented in the `st_cast` function from the `sf` package.
Importantly, `st_cast` behaves differently on single simple feature geometry (`sfg`) objects, and simple feature geometry column (`sfc`) and simple features objects.

Let's create a multipoint to illustrate how geometry casting works on simple feature geometry (`sfg`) objects:

```{r}
multipoint = st_multipoint(matrix(c(1, 3, 5, 1, 3, 1), ncol = 2))
```

In this case, `st_cast` can be useful to transform the new object into linestring or polygon (Figure \@ref(fig:single-cast)):

<!-- a/ points -> lines -> polygons  -->
```{r}
linestring = st_cast(multipoint, "LINESTRING")
polyg = st_cast(multipoint, "POLYGON")
```

```{r single-cast, echo = FALSE, fig.cap="Examples of linestring and polygon created based on multipoint using the `st_cast` function", warning=FALSE}
par_old = par()
par(mfrow = c(1, 3), mar = c(0, 0, 1, 0))
plot(multipoint, main = "MULTIPOINT")
plot(linestring, main = "LINESTRING")
plot(polyg, col = "grey", main = "POLYGON")
par(par_old)
```

This process can be also reversed using `st_cast`:

```{r}
multipoint_2 = st_cast(linestring, "MULTIPOINT")
multipoint_3 = st_cast(polyg, "MULTIPOINT")
all.equal(multipoint, multipoint_2, multipoint_3)
```

```{block2 type='rmdnote'}
For single simple feature geometries (`sfg`), `st_cast` provides also geometry casting from non-multi to multi types (e.g. `POINT` to `MULTIPOINT`) and from multi types to non-multi types.
However, only the first element of the old object would remain in the second group of cases.
<!-- note: beware of information lost (you will get a warning) -->
```

```{r, include=FALSE}
cast_all <- function(xg) {
  lapply(c("MULTIPOLYGON", "MULTILINESTRING", "MULTIPOINT", "POLYGON", "LINESTRING", "POINT"), 
         function(x) st_cast(xg, x))
}
t = cast_all(multipoint)
t2 = cast_all(polyg)
```

Geometry casting of simple feature geometry column (`sfc`) and simple features objects works the same as for single geometries in most of the cases. 
One imporant difference is conversion between multi to non-multi types.
As a result of this process, multi-objects are split into many non-multi objects.

We would use a new object, `multilinestring_sf`, as an example (on the left in Figure \@ref(fig:line-cast)):

```{r}
multilinestring_list = list(matrix(c(1, 4, 5, 3), ncol = 2), 
                            matrix(c(4, 4, 4, 1), ncol = 2),
                            matrix(c(2, 4, 2, 2), ncol = 2))
multilinestring = st_multilinestring((multilinestring_list))
multilinestring_sf = st_sf(geom = st_sfc(multilinestring))
multilinestring_sf
```

You can imagine it as a road or river network. 
The new object has only one row that define all the lines.
This restrict number of operation that could be done, for example it prevent adding names to each line segment or calculating lengths of single lines.
The `st_cast` function can be used in this situation, as it separates one mutlilinestring into three linestrings:

```{r}
linestring_sf2 = st_cast(multilinestring_sf, "LINESTRING")
linestring_sf2
```

```{r line-cast, echo=FALSE, fig.cap="Examples of type casting between MULTILINESTRING (left) and LINESTRING (right).", warning=FALSE}
par_old = par()
par(mfrow = c(1, 2), mar = c(0, 0, 1, 0))
plot(multilinestring_sf$geom, col = "black", main = "MULTILINESTRING")
plot(linestring_sf2$geom, col = c("red", "green", "blue"), main = "LINESTRING")
par(par_old)
```

The newely created object allows for attributes creation (see more in section \@ref(vec-attr-creation)) and length measurement:

```{r}
linestring_sf2$name = c("Riddle Rd", "Marshall Ave", "Foulke St")
linestring_sf2$length = st_length(linestring_sf2)
linestring_sf2
```

<!-- ### Class conversion -->
<!-- placeholder for: -->
<!-- sf -> sp -->
<!-- sp -> sf -->
<!-- stars; https://github.com/r-spatial/stars/blob/master/vignettes/blog1.Rmd -->

### Rasterization
<!-- - vector to raster -->
<!-- why? map tilling, simple analysis of terrain? -->
<!-- rasterize() -->
<!-- https://github.com/ecohealthalliance/fasterize -->

## Geometric operations on raster data {#geo-ras}

<!-- two useful citations for into on geometric operations on raster:

> Geometric operations include the shift, rotation and warping of images to a given shape or framework.
In remote sensing applications, geometric operations are mainly used for the co-registration of images of the same scene acquired by different sensor systems or at different times or from different positions, and for rectifying an image to fit a particular coordinate system (geocoding).
Image mosaic is a geometric operation that was commonly used in the early days of remote sensing image processing when computer power was inadequate for the massive demands of the geocoding process, but this is no longer the case.
Once a set of adjacent images is accurately rectified to a map projection system, such as a UTM coordinate system (see Chapter 13 in Part Two for details) the images, though separate, are effectively in a mosaic.

@liu_essential_2009: 105

> Vector data store their coordinate information implicitly with each node position in real-world coordinates and these are used directly to plot positions and to re-project from one coordinate system to another.
Raster data on the other hand have a regular local row and column number system for each pixel so that, internally, the geometry and position of a pixel’s location are implicit.
Externally, however, a world geographic reference must be explicitly stated; this requires the geographic location of the image origin and the ground distance represented by each pixel.
A transformation is then performed, which converts local image coordinates to realworld coordinates for each pixel location using the geometric operations described in Chapter 9. This transformation information is also stored explicitly.

@liu_essential_2009: 173-74

-->

### Raster alignment

```{r, echo=FALSE}
source("code/create-rasters.R")
```

When merging or performing map algebra on rasters, their resolution, projection, origin and/or extent has to match.
Otherwise, how should we add the values of one raster with a resolution of 0.2 decimal degrees to a second with a resolution of 1 decimal degree?
The same problem arises when we would like to merge satellite imagery from different sensors with different projections and resolutions.
We can deal with such mismatches by aligning the rasters.

This section uses the `elev` object from \@ref(manipulating-raster-objects).
The `projectRaster()` function reprojects one raster to a desired projection, say from UTM to WGS84.
Equally, map algebra operations require the same extent.
Following code adds one row and two columns to each side of the raster while setting all new values to an elevation of 1000 meters (\@ref(fig:extend-example)).

```{r extend-example, fig.cap = "Original raster extended by 1 one row on each side (top, bottom) and two columns on each side (right, left)."}
elev_2 = extend(elev, c(1, 2), value = 1000)
plot(elev_2)
```

Performing an algebraic operation on two objects with differing extents in R, the **raster** package returns the result for the intersection, and says so in a warning.

```{r}
elev_3 = elev + elev_2
```

However, we can also align the extent of two rasters with the `extend()` command.
Here, we extend the `elev` object to the extend of `elev_2`. 
The newly added rows and column receive the default value of the `value` parameter, i.e., `NA`.

```{r}
elev_4 = extend(elev, elev_2)
```

The `aggregate()` and `disaggregate()` functions help to change the cell size resolution of a raster.
For instance, let us aggregate `elev` from a resolution of 0.5 to a resolution of 1, that means we aggregate by a factor of 2 (Fig. \@ref(fig:aggregate-example)).
Additionally, the output cell value should correspond to the mean of the input cells (but  one could use other functions as well such as `median()`, `sum()`, etc.):

```{r aggregate-example, fig.cap = "Original raster (left). Aggregated raster (right)."}
elev_agg = aggregate(elev, fact = 2, fun = mean)
par(mfrow = c(1, 2))
plot(elev)
plot(elev_agg)
```

Note that the origin of `elev_agg` has changed, too.

```{r}
origin(elev)
origin(elev_agg)
```

The origin is the point closest to (0, 0) if you moved towards it in steps of x and y resolution.
If two rasters have different origins, their cells do not overlap completely which would make map algebra impossible.
To change the origin , use `origin()`.^[If the origins of two raster datasets are just marginally apart, it sometimes is sufficient to simply increase the `tolerance` argument  of `raster::rasterOptions()`.]
Looking at figure \@ref(fig:origin-example) reveals the effect of changing the origin.

```{r origin-example, fig.cap = "Plotting rasters with the same values but different origins."}
# plot the aggregated raster
plot(elev_agg)
# change the origin
origin(elev_agg) = c(0, 0)
# plot it again
plot(elev_agg, add = TRUE)
```

The `resample()` command lets you align several raster properties in one go, namely origin, extent and resolution.
Let us resample an extended `elev_agg` to the properties of `elev` again.

```{r}
# add 2 rows and columns, i.e. change the extent
elev_agg = extend(elev_agg, 2)
elev_disagg = resample(elev_agg, elev)
```

Though our disaggregated `elev_disagg` retrieved back its original resolution, cell size and extent, its values differ. 
However, this is to be expected, disaggregating **cannot** predict values at a finer resolution, it simply uses an interpolation algorithm.
It is important to keep in mind that disaggregating results in a finer resolution, the corresponding values, however, are only as accurate as their lower resolution source.

Finally, if you want to align many (possibly hundreds or thousands of) images stored on disk, you might want to checkout the `gdalUtils::align_rasters()` function.
Nevertheless, you may also use **raster** with very large datasets. 
This is because **raster**:

1. lets you work with raster datasets that are too large to fit into the main memory (RAM) by only processing chunks of it.
2. tries to facilitate parallel processing.
For more information have a look at the help pages of `beginCluster()` and `clusteR()`.
Additionally, check out the *Multi-core functions* section in `vignette("functions", package = "raster")`.

### Aggregation {#ras-agg}

### Vectorization

<!-- ref to rasterization -->
Vectorization is a process of converting rasters into vectors.^[This term should not be confused with a code vectorization.]
The simplest form of vectorization is a conversion from a raster to points.
The `rasterToPoints()` function creates point representations of every non-NA raster grid cell centroids and it is usually used for continuous data, such as elevation (Figure \@ref(fig:raster-vectorization1)).
<!-- Spatial* class mentioned -->
<!-- why? -->

```{r}
elev_point = rasterToPoints(elev, spatial = TRUE) %>% 
  st_as_sf()
```

```{r raster-vectorization1, echo=FALSE, fig.cap="Raster and point representation of the elev dataset.", warning=FALSE}
library(tmap)
library(grid)
colfunc = colorRampPalette(c("lightyellow", "rosybrown"))

p1 = tm_shape(elev) + 
  tm_raster(legend.show = FALSE, palette = colfunc(36), n = 36) +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4),
            title = "raster")

p2 = tm_shape(elev_point) + 
  tm_bubbles(col = "layer", legend.col.show = FALSE, palette = colfunc(36), n = 36) +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0.05, 4),
            title = "points")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
print(p1, vp = viewport(layout.pos.col = 1))
print(p2, vp = viewport(layout.pos.col = 2))
```

On the other hand, categorical rasters (such as scanned maps or satellite images) represent discrete features that could occupy areas larger than only one cell.
The `rasterToPolygons()` function could be used to extract these features into a `SpatialPolygonsDataFrame` object.
We can also convert the result into an object of class sf.^[The `spex::polygonize()` function could be used as a faster alternative. It returns an sf object a default.]
<!-- it could be also used when we want to add attributes -->

```{r}
grain_poly = rasterToPolygons(grain) %>% 
  st_as_sf()
```

The output consists of many polygons representing grid cells from the original raster object.
They could be aggregated into irregular polygons having the same value (Figure \@ref(fig:raster-vectorization1)). ^[see section \@ref(spatial-aggr)]

```{r}
grain_poly2 = grain_poly %>% 
  group_by(layer) %>%
  summarize()
```

```{r raster-vectorization2, echo=FALSE, fig.cap=".", warning=FALSE}
library(tmap)
library(grid)
cols = c("sand" = "rosybrown", "clay" = "brown", "silt" = "sandybrown")

p1p = tm_shape(grain) + 
  tm_raster(legend.show = FALSE, palette = cols) +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4),
            title = "raster")

p2p = tm_shape(grain_poly) + 
  tm_polygons("layer", legend.show = FALSE, palette = cols, lwd = 3) +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4),
            title = "polygons")

p3p = tm_shape(grain_poly2) + 
  tm_polygons("layer", legend.show = FALSE, palette = cols, lwd = 3) +
  tm_layout(outer.margins = rep(0.01, 4), 
            inner.margins = rep(0, 4),
            title = "aggregated polygons")

grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 3)))
print(p1p, vp=viewport(layout.pos.col = 1))
print(p2p, vp=viewport(layout.pos.col = 2))
print(p3p, vp=viewport(layout.pos.col = 3))
```

<!-- rasterToContour() -->
<!-- e.g. landcover map to polygonss -->

<!-- ```{r} -->
<!-- nlcd2011 = raster(system.file("raster/nlcd2011.tif", package = "spDataLarge")) -->
<!-- plot(nlcd2011) -->
<!-- ``` -->

<!-- it's slow - maybe it worth to use https://www.rdocumentation.org/packages/spex/versions/0.4.0 -->
<!-- ```{r} -->
<!-- nlcd2011_poly = rasterToPolygons(nlcd2011) -->
<!-- plot(nlcd2011_poly) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- system.time({nlcd2011_poly2 = polygonize(nlcd2011)}) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- nlcd2011_cont = rasterToContour(nlcd2011) %>%  -->
<!--   st_as_sf() -->
<!-- plot(nlcd2011_cont) -->
<!-- ``` -->

## Exercises

<!-- CRS CONVERSION -->
<!-- 1. vector reprojection exercise (e.g. modification of proj4) -->
1. Transform the `world` dataset to the transverse Mercator projection (`"+proj=tmerc"`) and plot the result.
What has changed and why?
Try to transform it back into WGS 84 and plot the new object.
Why the new object differs from the original one?
```{r, echo=FALSE, eval=FALSE}
# see https://github.com/r-spatial/sf/issues/509
world_tmerc = st_transform(world, "+proj=tmerc")
plot(world_tmerc$geom)
 world_4326 = st_transform(world_tmerc, 4326)
plot(world_4326$geom)
```
1. Write code that uses functions `aggregate()` and `st_buffer()` to answers the following question: What proportion of the world's population lives in countries that intersect a circle with a 10 degree radius of the intersection between the equator and the [9^th^ meridian](https://en.wikipedia.org/wiki/9th_meridian_east)?
```{r, echo=FALSE, eval=FALSE}
center9 = st_sf(st_sfc(st_point(c(-9, 0)), crs = 4326))
buff9 = st_buffer(center9, dist = 10)
agg9 = aggregate(world["pop"], buff9, FUN = sum)
agg9$pop / sum(world$pop, na.rm = TRUE) # 1%
```
1. Assuming that people are evenly distributed across countries, estimate the population living *within* the circle created to answer the previous question.
    ```{r, echo=FALSE, eval=FALSE}
    interp9 = st_interpolate_aw(x = world["pop"], to = buff9, extensive = TRUE)
    ```
1. Warning messages should have been produced during the working to find the answer to the previous questions. What do these warnings mean and how could they be stopped? 
    - Bonus: rewrite code that generated the answer to the previous question using a projected CRS (suggestion: UTM).
<!-- AFFINE TRANSFORMATION -->
<!-- CLIPPING -->
1. Write code that subsets points that are contained within `x` *and* `y` (illustrated by the plot in the 2^nd^ row and the 1^st^ column in Figure \@ref(fig:venn-clip)).
    - Create a randomly located point with the command `st_point()` (refer back to section \@ref(sfg) to see how to create spatial data 'from scratch').
1. Try to transform the categorical raster (`cat_raster`) into WGS 84 using the bi-linear interpolation method. 
What has changed?
How it influences the results?
```{r, echo=FALSE, eval=FALSE}
wgs84 = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
cat_raster_wgs84 = projectRaster(cat_raster, crs = wgs84, method = "bilinear")
cat_raster_wgs84
```
1. Transform the continuous raster (`cat_raster`) into WGS 84 using the nearest neighbor interpolation method. 
What has changed?
How it influences the results?
1. Advanced challenge: find the point with the highest number of people within a 10 degree radius.
```{r, echo=FALSE, eval=FALSE}
con_raster = raster(system.file("raster/srtm.tif", package="spDataLarge"))
con_raster_wgs84 = projectRaster(con_raster, crs = wgs84, method = "ngb")
con_raster_wgs84
```
<!-- GEOMETRY TRANSFORMATION -->
<!-- VECTOR -->
<!-- SIMPLIFICATION -->
<!-- CENTROIDS AND BUFFERS-->

<!-- UNIONS -->
<!-- TYPE TRANS. -->
<!-- RASTERIZATION -->
<!-- RASTER -->
<!-- ALIGNMENT -->
<!-- AGGREGATION -->
<!-- VECTORIZATION -->
