# Coordinate reference systems/reprojecting {#coord}

<!-- - crs - a heart of spatial data -->
<!-- - short history  -->
<!-- - types of crs (geographic vs cartesian; local vs regional vs global) -->
<!-- - objectives - 1/ to combine different datasets, 2/ area calculations, 3/ distance mesasurement, 4/ navigation, 5/ spatial data representations -->
<!-- - proj -->
<!-- - proj4 + epsg -->
<!-- - the most popular epsg -->

## Prerequisites {-}

- This chapter requires **tidyverse**, **sf**, and **spData** packages:

```{r, message=FALSE}
library(sf)
library(tidyverse)
```

- You must have loaded the `world` data from the spData package:

```{r, results='hide'}
library(spData)
```

## Introduction

<!--in most of the cases reproject vector, not raster-->

## Vector data

<!-- CRS assign -->

- In case when a coordinate reference system (CRS) is missing or the wrong CRS is set, `st_crs()` or `st_set_crs` function can be used:

```{r, warning=FALSE}
world_set3410 = st_set_crs(world, 3410)
st_crs(world_set3410)
```


```{r coord_compare0, echo=FALSE}
par(mfrow = c(1, 2), mar=c(0,0,0,0))
plot(world_set3410[0]);plot(world[0])
```

<!-- Reprojection -->

- The `st_transform()` can be used to transform coordinates

```{r}
world_3410 = st_transform(world, 3410)
st_crs(world_3410)
```

```{r coord_compare, echo=FALSE}
par(mfrow = c(1, 2), mar=c(0,0,0,0))
plot(world_3410[0]);plot(world[0])
```

<!--
- st_as_sf(x, coords = c("x","y"))
- st_crs(x)
- st_transform(x, crs)
- ==
- !st_is_longlat(x)
- st_set_crs(x, crs)
- st_proj_info
- st_bbox
-->

## Raster data

<!--
- projectRaster
-->
