# Reprojections and Transformations {#coord}

<!-- - crs - a heart of spatial data -->
<!-- - short history  -->
<!-- - types of crs (geographic vs cartesian; local vs regional vs global) -->
<!-- - objectives - 1/ to combine different datasets, 2/ area calculations, 3/ distance mesasurement, 4/ navigation, 5/ spatial data representations -->
<!-- - proj -->
<!-- - proj4 + epsg -->
<!-- - the most popular epsg -->

## Prerequisites {-}

- This chapter requires **tidyverse**, **sf**, and **spData** packages:

```{r, message=FALSE}
library(sf)
library(tidyverse)
```

- You must have loaded the `world` data from the spData package:

```{r, results='hide'}
library(spData)
```

## Introduction

<!--in most of the cases reproject vector, not raster-->

## Setting and transforming the CRS


Note that the result shows that the `epsg` has been updated and that `proj4string` element of the CRS has been updated to contain, among other things `+proj=tmerc` (meaning it is a projected CRS using the [tranverse Mercator](https://en.wikipedia.org/wiki/Transverse_Mercator_projection) projection) and `+units=m` (meaning the units of the coordinates are meters).
Another function, from the **rgdal** library, provides a note about the CRS (its name):

```{r}
crs_codes = rgdal::make_EPSG()[1:2]
dplyr::filter(crs_codes, code == 27700)
```

The result shows that the EPSG code 27700 represents the British National Grid, a result that could have been found by searching online for "[CRS 27700](https://www.google.com/search?q=CRS+27700)".
This projection is clearly inappropriate for the data: the coordinates represent degrees of longitude and latitude, and this can also be seen by plotting it over a basemap, e.g. with the **mapview** package: `mapview::mapview(sf_points)`.

The formula that converts a geographic point into a point on the surface of the Earth is provided by the `proj4string` element of the `crs` (see [proj4.org](http://proj4.org/) for further details):

```{r}
st_crs(27700)$proj4string
```

The CRS of `sf_points` can be restored to its orginial, correct value as follows (which demonstrates another way of setting CRSs):

```{r}
st_crs(sf_points) = old_crs
```

### Pro tip {-}

The EPSG code can be found inside the `crs` attribute of the object's geometry.
It is hidden from view for most of the time except when the object is printed but can be can identified and set using the functions `attribute()` and `attr()`, as illustrated below:

```{r}
attr(sf_points$geometry, which = "crs")$epsg
```

### Vector data

<!-- CRS assign -->

- In case when a coordinate reference system (CRS) is missing or the wrong CRS is set, `st_crs()` or `st_set_crs` function can be used:

```{r, warning=FALSE}
world_set3410 = st_set_crs(world, 3410)
st_crs(world_set3410)
```


```{r coord_compare0, echo=FALSE}
par(mfrow = c(1, 2), mar=c(0,0,0,0))
plot(world_set3410[0]);plot(world[0])
```

<!-- Reprojection -->

- The `st_transform()` can be used to transform coordinates

```{r}
world_3410 = st_transform(world, 3410)
st_crs(world_3410)
```

```{r coord_compare, echo=FALSE}
par(mfrow = c(1, 2), mar=c(0,0,0,0))
plot(world_3410[0]);plot(world[0])
```

<!--
- st_as_sf(x, coords = c("x","y"))
- st_crs(x)
- st_transform(x, crs)
- ==
- !st_is_longlat(x)
- st_set_crs(x, crs)
- st_proj_info
- st_bbox
-->

### Raster data

<!--
- projectRaster
-->

## Affine transformations

### Translating

### Scaling

### Rotating

### Reflecting

### Shearing
