# Advanced map making

## Prerequisites {-}

- ...

```{r, include=FALSE}
pkgs_c11 = c(
  "cowplot",
  "wbstats",
  "rnaturalearth",
  "viridis",
  "sf",
  "tidyverse",
  "tmap",
  "units"
)
pacman::p_load(char = pkgs_c11)
if(!exists("world")) {
  source("code/01-introduction.R")
}
```

# Advanced map making {#adv-map}

<!-- 
- tmap
- ggplot2/ggmap (geom_sf, coord_sf)
- leaflet/mapview
- rasterVis ?
- shiny ?
- animations
- inset maps
- color palettes (viridis, etc.)
-->

## Animations

Animated maps can be useful for communicating how spatial phenomena shift over time.
An advantage of facetted plots are that they can be printed, but the approach has disadvantages:
faceted maps can become very small with more than ~9 maps in one figure, and it can be hard to see the spatial relationships between each facet when each map is on a different part of the page!

```{r ani_data_prep}
options(scipen = 999)

wb_data_create = function(indicator, our_name, year, ...){
  df = wb(indicator = indicator, startdate = year, enddate = year, ...) %>%
    as_data_frame() %>%
    select(iso_a2=iso2c, value) %>%
    mutate(indicator = our_name) %>%
    spread(indicator, value)
  return(df)
}

data_lifeExp = seq(1963, 2013, by=5) %>%
  set_names(.) %>%
  map_df(~wb_data_create(.x, indicator = "SP.DYN.LE00.IN",
                   our_name = "lifeExp",
                   country = "countries_only"), .id='year') %>%
  spread(year, lifeExp)

world_sf_temporal = ne_countries(returnclass = 'sf') %>%
  left_join(., data_lifeExp, by = c('iso_a2')) %>%
  mutate(area_km2 = set_units(st_area(.), km^2)) %>%
  select(iso_a2, name_long, continent, region_un, subregion, type, area_km2, `1963`:`2013`) %>%
  gather(year, lifeExp, `1963`:`2013`)
```

```{r animation1, eval=FALSE}
m1 = tm_shape(world_sf_temporal) + 
  tm_polygons("lifeExp") +
  tm_facets(by = "year", nrow = 1, ncol = 1, drop.units = TRUE)

animation_tmap(m1, filename = "figures/11-lifeExp_animation.gif", width = 2000, height = 600, delay = 40)
```

```{r animation2, eval=FALSE}
world_sf_temporal2 = filter(world_sf_temporal, continent == "South America")
m2 = tm_shape(world_sf_temporal2) +
  tm_polygons("lifeExp", n = 12) +
  tm_facets(by = "name_long", along = "year", drop.units = TRUE, free.coords = TRUE)
animation_tmap(m2, filename = "figures/11-lifeExp_sa_animation.gif", width = 1600, height = 1000, delay = 40)
```
