# Bridges to GIS software {#gis}

An important feature of R is that users must use the command-line interface (CLI) to interact with the computer:
you type commands and hit `Enter` to execute them interactively.
The most popular open-source GIS software packages, by contrast, feature a prominent graphical user interface (GUI):
you *can* interact with QGIS, SAGA and gvSIG by typing into (dockable) command-lines, but most users interact with such programs by 'pointing and clicking their way through life' as Gary Sherman puts it below [@sherman_desktop_2008]:^[
<!-- yes, we should shorten the footnote or put it somewhere into the text. I just rewrote it to make clearer what was meant. At least this was what I gathered. -->
The mature GRASS GIS software package and PostGIS are quite popular in academia and industry, and could be seen as products which buck this trend as they are built around the command-line. 
In [2008](http://gama.fsv.cvut.cz/~landa/publications/2008/gis-ostrava-08/paper/landa-grass-gui-wxpython.pdf) GRASS developers added a sophisticated GUI, shifting the emphasis slightly away from its CLI.
However, GRASS lacks a sophisticated and feature-rich IDE such as RStudio that supports 'CLI newbies'. 
On the other hand, PostGIS is the spatial extension of the popular PostgreSQL open source database, and therefore not really a dedicated GIS. 
This is also highlighted by the fact that PostGIS lacks any geovisualization capabilities and its description of 'legacy GIS' on its [website](http://workshops.boundlessgeo.com/postgis-intro/introduction.html).
Similar to GRASS, PostgreSQGL provides a (partial) GUI called [pgAdmin](https://www.pgadmin.org/) to facilitate the editing and administration of the database. 
To make it clear, though PostGIS provides spatial functions, its main purpose is the handling of spatial objects in a relational database management system.
Therefore, frequently users store their spatial data in PostGIS, and interact with it through a dedicated GIS software such as QGIS. Of course, you can also use R to access data from PostGIS (**sf**, **rgdal**, **rpostgis**).
In summary,  a typical workflow would be: perform a large spatial query with PostGIS, then load the result into a further application (QGIS, R) for further geoprocessing.   
]

> With the advent of 'modern' GIS software, most people want to point and
click their way through life. That’s good, but there is a tremendous amount
of flexibility and power waiting for you with the command line. Many times
you can do something on the command line in a fraction of the time you
can do it with a GUI.

Gary Sherman is well-qualified to comment on this matter as the creator of QGIS, the world's premier open source GUI-based GIS!

The 'CLI vs GUI' debate is often adversarial and polarized but it does not have to be: both options are great if well chosen in accordance with your needs and tasks.
The advantages of a good CLI such as that provided by R are numerous. Among others, a CLI:

- facilitates the automation of repetitive tasks. We strongly dislike the manual execution of iterations, and would recommend to always think about how to solve such tasks programmatically.
This way, you most likely discover new programming features and concepts, i.e., you learn and enhance your skill set. 
By contrast, what are you learning from executing the same tasks a 1000 times?
As a nice side-effect, automation is surely more effective and less error-prone.
- ensures transparency and reproducibility (which also is the backbone of good scientific practice).
In short, it is easier to share your R script than explain a sequence of 10+ 'points and clicks'.
- encourages extending existing software by making it easy to modify and extend functions.
If you are missing an algorithm you need, the CLI gives you the freedom to write your own!
<!-- add link to most sexiest job in the 21st century -->
- is the way to (geo-)data science.
Professional and advanced technical skills will certainly enhance your career prospects, and are in dire need across a wide range of disciplines.
- is fun, but admittedly that is a subjective argument.
<!-- any other points that we have missed? -->
- ... 

On the other hand, GUI-based GIS systems (particularly QGIS) are also advantageous. For instance, think of:

- the GUI. The really user-friendly graphical interface spares the user from programming.
Though you probably wouldn't read the book if this were your main objective.
- digitizing and all related tools (trace, snap, topological rules, etc.). Note that there is also the new **mapedit** package but its intention is to allow the quick editing of a few spatial features, and not professional, large-scale cartographic digitizing.
- georeferencing
- stereoscopic mapping (e.g., LiDAR and structure from motion)
- the built-in geodatabase management system often integrated in Desktop GIS (ArcGIS, GRASS GIS) and all related advantages such as object relational modeling, topology, fast (spatial) querying, etc.
- map production, in case you only want to create a beautiful map once. If you have to produce it over and over again, then maybe the CLI approach is better.
- zooming and dragging on WMS (though this is also possible with **mapview** and **leaflet**)
<!-- any other points that we have missed so far? -->
- ...

This general overview already points out the differences between R's CLI and desktop GIS. However, there is more: dedicated GIS software provides hundreds of geoalgorithms that are simply missing in R.
The good news is that 'GIS bridges' enable the access to these with the comfort of the R command line.^[
The term 'bridge' was probably first used in the R-spatial world for the coupling of R with GRASS [@neteler_open_2008].
Roger Bivand elaborates on this in his talk "Bridges between GIS and R", delivered at the 2016 GEOSTAT summer school.
The resulting slides can be found on Roger's personal website at [http://spatial.nhh.no/misc](http://spatial.nhh.no/misc/?C=M;O=D) in the file
`geostat_talk16.zip`.
]

The R language was originally designed as an interface to and extension of other languages, especially C and FORTRAN, to enable access to statistical algorithms in a user-friendly and intuitive read-evaluate-print loop (REPL) [@chambers_extending_2016].
R was not originally intended to be a GIS.
This makes the breadth of R's geospatial capabilities astonishing to many who are unaware of its ability to replicate established GIS software for many operations on spatial data.
There are some domains where R can now outperform desktop GIS including spatial statistical modeling, online interactive visualization and the generation of animated or faceted maps.

Instead of implementing existing GIS algorithms in R, it makes sense to avoid 'reinventing the wheel' by taking advantage of R's ability to interface with other languages (especially C++, which is used for much low-level and high-performance GIS work).
Using compiled code for new geoprocessing functionality (particularly with the help of the excellent **Rcpp** package) could form the basis of new R packages, building on the success of **sf**.
However, there is already a wide range of algorithms that can be accessed via R's interfaces to dedicated GIS software.
It makes sense to understand these before moving to develop your own optimized algorithms.
For this reason this chapter focuses on 'bridges' to the mature GIS products [QGIS](http://qgis.org/) (via the package **RQGIS**), [SAGA](http://www.saga-gis.org/) (**RSAGA**) and [GRASS](https://grass.osgeo.org/) (**rgrass7**) from within R.
Obviously, we here focus on open-source software solutions, however, there is also a bridge to the commercial GIS leader [ArcGIS](https://www.arcgis.com) through the **RPyGeo** package.
And the so-called [R-ArcGIS bridge](https://github.com/R-ArcGIS/r-bridge) allows to use R from within ArcGIS.
As a final note, we would like to point out that aside from interfaces to desktop GIS there are also interfaces to geospatial libraries such as [GDAL](www.gdal.org) (**gdalUtils**, **rgdal**, **sf**) and [GEOS](https://trac.osgeo.org/geos/) (**rgeos**, **sf**). 
By the end of the chapter you should have a working knowledge of the functionality such packages open up, and a more nuanced understanding of the  'CLI vs GUI' debate.
As mentioned in chapter \@ref(intro), doing GIS at the command-line makes it more reproducible, in-line with the principles of Geographic Data Science.

## (R)QGIS
QGIS is one of the most popular open-source GIS [\@ref(tab:gis_comp); @graser_processing:_2015]. Its main advantage lies in the fact that it provides a unified interface to several other open-source GIS.
First and foremost, this means that you have access to GDAL/OGR, GRASS and SAGA through QGIS but also to other third-party providers such as [TauDEM](http://hydrology.usu.edu/taudem/taudem5/index.html), [Orfeo Toolbox](https://www.orfeo-toolbox.org/) and [Lastools](https://rapidlasso.com/lastools/) (tools for LiDAR data) [@graser_processing:_2015]. 
To run all these geoalgorithms (frequently more than 1000 depending on your set up) outside of the QGIS GUI, QGIS provides a Python API.
**RQGIS** establishes a tunnel to this Python API through the **reticulate** (cite reticulate) package. 
Basically, functions `set_env` and `open_app` are doing this. 
Note that it is optional to run `set_env` and `open_app` since all functions depending on their output will run them automatically if needed.
Before running **RQGIS** you have to make sure to have installed QGIS and all its (third-party) dependencies such as SAGA and GRASS.
To help you with the installation process, please follow the steps as detailed in `vignette("install_guide", package = "RQGIS")` for several platforms (Windows, Linux, MacOS).

```{r qgis_setup}
library("RQGIS")
set_env()
open_app()
```

Leaving the `path`-argument of `set_env` unspecified will search the computer for a QGIS installation.
Hence, it is faster to specify explicitly the path to your QGIS installation.
Subsequently, `open_app` sets all paths necessary to run QGIS from within R, and finally creates a so-called QGIS custom application [http://docs.qgis.org/testing/en/docs/pyqgis_developer_cookbook/intro.html#using-pyqgis-in-custom-applications](http://docs.qgis.org/testing/en/docs/pyqgis_developer_cookbook/intro.html#using-pyqgis-in-custom-applications).
We are now ready for some QGIS geoprocessing from within R! First of all, we load some data from the **spData**-package, namely the boroughs of London (`lnd`) and cycle hire points in London (`cycle_hire`).

```{r}
library("spData")
data("cycle_hire")
data("lnd")
```
<!-- silly example, other suggestions welcome also with regard to RSAGA and rgrass7-->
In chapter @ref(spatial-class), we already learned how to do a spatial overlay using the **sf**-package.
Of course, any GIS is also able to perform spatial overlays. Here, we would like to know how many cycle points we can find per borough.
First of all, we need to come up with the name of the function in QGIS. `find_algorithms` lets you search all QGIS geoalgorithms with the help of regular expressions.
Here, we assume that the short description of the function contains first the word "point" and secondly somewhere later also the word "poly".
If you have no clue at all what to look for you can leave the `search_term`-argument empty which will return a list of all available QGIS geoalgorithms.
If you also want to have a short description for each geoalgorithm, set the `name_only`-parameter to FALSE.

```{r}
find_algorithms("points.*poly", name_only = TRUE)
```

Now that we know the name of the function ("qgis:countpointsinpolygon"), we wonder how we can use it.
`get_usage` returns all function parameters and default values. 
`open_help` lets you access the corresponding online help.

```{r}
alg = "qgis:countpointsinpolygon"
get_usage(alg)
```

```{r, eval = FALSE}
open_help(alg)
```

Finally, we can let QGIS do the work.
Note that the workhorse function `run_qgis` accepts R named arguments, i.e., you can specify the parameter names as returned by `get_usage` as you would do in any other regular R function.
Note also that `run_qgis` accepts spatial objects residing in R's global environment as input (here: `lnd` and `cycle_hire`). 
But of course, you could also specify paths to shapefiles stored on disk.
Setting the `load_output` to `TRUE` automatically loads the QGIS output into R.
Since we only indicated the name of the output ("cycle.shp"), `run_qgis` saves the output to a temporary folder as returned by `tempdir()`, and loads it into R as an **sf**-object.

```{r}
bike_points = run_qgis(alg, POLYGONS = lnd, POINTS = cycle_hire, FIELD = "no_bikes", 
                       OUTPUT = "cycle.shp", load_output = TRUE)
summary(bike_points$no_bikes)
sum(bike_points$no_bikes > 0)
```

In case you leave some parameters of a geoalgorithm unspecified, `run_qgis` will automatically use the default values as arguments if available.
To find out about the default values, run `get_args_man`.  

```{r}
get_args_man(alg)
```

In this case the output tells us, had we left the `FIELDS`-parameter unspecified, our output (attribute) field would have been named "NUMPOINTS" (instead of "no_bikes").

<!--
"grass7:v.vect.stats" would achieve the same but is unavailable in QGIS
-->

Other notes:
- Leaving the output parameter(s) unspecified, saves the resulting QGIS output to a temporary folder created by QGIS.
`run_qgis` prints these paths to the console after successfully running the QGIS engine.
- If the output consists of multiple files and you have set `load_output` to `TRUE`, `run_qgis` will return a list with each element corresponding to one output file.

To learn more about **RQGIS** please refer to the (hopefully) forthcoming paper (cite). 

## (R)SAGA
Similar to QGIS, the System for Automated Geoscientific Analyses (SAGA;  \@ref(tab:gis_comp)) provides the possibility to run SAGA modules from Python (SAGA Python API).
In addition, there is also a command line interface (saga_cmd.exe) to execute SAGA modules (see also [https://sourceforge.net/p/saga-gis/wiki/Executing%20Modules%20with%20SAGA%20CMD/](https://sourceforge.net/p/saga-gis/wiki/Executing%20Modules%20with%20SAGA%20CMD/)).
**RSAGA** uses the latter to run SAGA from within R.

Though SAGA is a hybrid GIS, its main focus has been on raster processing, and here particularly on digital elevation models (soil properties, terrain attributes, climate parameters). 
Hence, SAGA is especially good at the fast processing of large (high-resolution) rasters data sets [@conrad_system_2015]. 
Therefore, we will introduce **RSAGA** with a raster and use case from @muenchow_geomorphic_2012.
Specifically, we would like to compute the SAGA wetness index from a digital elevation model.
First of all, we need to make sure that **RSAGA** will find SAGA on the computer when called.
For this, all **RSAGA** functions using SAGA in the background make use of `rsaga.env()`. 
Usually, `rsaga.env()` will detect SAGA automatically by searching several likely directories (see its help for more information). However, we have 'hidden' SAGA in the OSGEO4W-installation, a location `rsaga.env()` does not search automatically. `linkSAGA` searches your computer for a valid SAGA installation. If it finds one, it adds it to the PATH environment variable thereby making sure that `rsaga.env` runs successfully.

```{r, warning = FALSE, message = FALSE}
library("RSAGA")
library("link2GI")
linkSAGA()
rsaga.env()
```

Secondly, we need to write the digital elevation model to a SAGA-format. 
Note that calling `data("landslides")` attaches two object to the global environment - `dem`, a digital elevation model in the form of a `list`, and `landslides`, a `data.frame` containing observations representing the presence or absence of a landslide:

```{r}
data("landslides")
write.sgrd(data = dem, file = file.path(tempdir(), "dem"), header = dem$header)
```
 
The organization of SAGA is modular.
Libraries contain so-called modules, i.e., geoalgorithms.
To find out which libraries are available, run:

```{r}
tail(rsaga.get.libraries(), 10)
```

Instead of presenting all available libraries we have shown only the last ten. 
We choose the library 'ta_hydrology' (ta is the abbreviation for terrain analysis).
Subsequently, we can access the available modules of a specific library (here: 'ta_hydrology') as follows:

```{r}
rsaga.get.modules(libs = "ta_hydrology")
```

Similarly to `RQGIS::get_usage()`, `rsaga.get.usage()` prints the function parameters of a specific geoalgorithm, e.g., the `SAGA Wetness Index`, to the console.

```{r, eval = FALSE}
rsaga.get.usage(lib = "ta_hydrology", module = "SAGA Wetness Index")
```

Finally, you can run SAGA from within R using **RSAGA**'s geoprocessing workhorse function `rsaga.geoprocessor()`. 
The function expects a parameter-argument list in which you have specified all necessary parameters.

```{r}
params = list(DEM = file.path(tempdir(), "dem.sgrd"),
              TWI = file.path(tempdir(), "twi.sdat"))
rsaga.geoprocessor(lib = "ta_hydrology", module = "SAGA Wetness Index", 
                   param = params)
```

To facilitate the access to the SAGA interface, RSAGA frequently provides user-friendly wrapper-functions with meaningful default values (see RSAGA documentation for examples, e.g., `?rsaga.wetness.index`).
So the function call for calculating the 'SAGA Wetness Index' becomes as simple as:

```{r}
rsaga.wetness.index(in.dem = file.path(tempdir(), "dem"), 
                    out.wetness.index = file.path(tempdir(), "twi"))
```

Of course, we would like to inspect our result visually. To load and plot the SAGA output file, we use the **raster** package. 

```{r}
library("raster")
twi = raster::raster(file.path(tempdir(), "twi.sdat"))
plot(twi, col = RColorBrewer::brewer.pal(n = 9, name = "Blues"))
# or using mapview
# proj4string(twi) = paste0("+proj=utm +zone=17 +south +ellps=WGS84 +towgs84=", 
#                           "0,0,0,0,0,0,0 +units=m +no_defs")
# mapview(twi, col.regions = RColorBrewer::brewer.pal(n = 9, "Blues"), 
#         at = seq(cellStats(twi, "min") - 0.01, cellStats(twi, "max") + 0.01, 
#                  length.out = 9))
```

You can find a much more extended version of the here presented example in the RSAGA vignette `vignette("RSAGA-landslides")`.
This example includes statistical geocomputing, i.e., it uses a GIS to derive terrain attributes as predictors for a non-linear Generalized Linear Model (GAM) to predict spatially landslide susceptibility [@muenchow_geomorphic_2012].
The term statistical geocomputation emphasizes the strength of combining R's data science power with the geoprocessing power of a GIS which is at the very heart of building a bridge from R to GIS.

## GRASS through **rgrass7**

Geographical Resources Analysis Support System (GRASS)   \@ref(tab:gis_comp) [@neteler_open_2008].
Inexperienced GIS users might find GRASS a bit intimidating in the beginning.
Before starting the GUI you have to create a map-set, a location (basically a geodatabase), define the projection and boundaries of your GRASS project. 
You have to do exactly the same from within R:


```{r, eval = FALSE}
library("rgrass7")
initGRASS(gisBase = "C:/Program Files (x86)/GRASS GIS 7.4.2")
```

**link2GI** facilitates this tedious task.
So instead of defining all the stuff yourself, you call:
 
 ```{r}
 library("link2GI")
  ```
use **link2GI** for an easy set-up of the GRASS environment (though it is beneficial to GRASS users to have a working knowledge of how GRASS works internally)

example ideas: 
- r.sim.water (requires various processing steps)
- traveling salesman problem; nice example, finding the shortest paths between 25 randomly select cycle points.


```{r}
library(sf)
# Download osm data
dir_main <- tempdir()
dir_main <- "D:/test"
download.file(file.path("http://download.geofabrik.de/europe/great-britain/", 
                        "england/greater-london-latest-free.shp.zip"),
              destfile = file.path(dir_main, "london_osm.zip"))
# unzip the downloaded file
unzip(zipfile = file.path(dir_main, "london_osm.zip"), exdir = dir_main)
# load streets into R environment
streets = sf::st_read(file.path(dir_main, "gis.osm_roads_free_1.shp"))
st_crs(streets) = 4326
data(cycle_hire, package = "spData")

b_box = st_bbox(cycle_hire)
coords = expand.grid(c(b_box["xmin"], b_box["xmax"]), 
                     c(b_box["ymin"], b_box["ymax"]))[c(1, 3, 4, 2, 1), ]
poly <- st_polygon(list(as.matrix(coords)))
poly <- st_sfc(poly)
st_crs(poly) = 4326
int = st_intersects(streets, poly)
streets_int = streets[sapply(int, length) > 0, ]
plot(streets_int$geometry)
plot(cycle_hire$geometry, col = "red", pch = 16, add = TRUE)
# randomly select 25 points
points = cycle_hire[sample(1:nrow(cycle_hire), 25), ]

library(rgrass7)
library(maptools)
library(link2GI)
linkGRASS7(streets_int, c("C:/OSGeo4W64", "grass-7.2.1", "osgeo4W"))

# library(spgrass6)
# g <- initGRASS(gisBase = "C:/Program Files (x86)/GRASS GIS 6.4.4", )
# execGRASS("g.mapset", mapset = "PERMANENT")
# execGRASS("g.proj", flag = "c", 
#           proj4 = "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0")
writeVECT(as(streets_int, "Spatial"), "streets", v.in.ogr_flags = "o")
writeVECT(as(points, "Spatial"), "points")

execGRASS(cmd = "v.clean", input = "streets", output = "streets_clean",
          tool = "break")
# prepare network by connecting points to roads / changes roads
execGRASS(cmd = "v.net", input = "streets_clean", output = "streets_points_con", 
          points = "points", operation = "connect", threshold = 0.001,
          flags = c("overwrite", "c"))
# run tsp
execGRASS(cmd = "v.net.salesman", input = "streets_points_con",
          output = "shortest_route", center_cats = paste0("1-", nrow(points)),
          flags = c("overwrite"))

route <- readVECT("shortest_route")
mapview::mapview(route) +
  mapview::mapview(points)
```


## When to use what?
<!-- more or less brainstorming -->
**RQGIS** unified interface to GRASS and SAGA (and GDAL/OGR, Orfeo Toolbox, TauDEM, Lastools) + QGIS geoalgorithms. 
So far QGIS supports a subset of the available geoalgorithms. Depending on your version this is around 235 for SAGA and 160 for GRASS. 
So you should stick to **RSAGA** and **rgrass7** if you are also using less common GIS functionality.
In case you aim for topological correct data, we recommend the usage of GRASS. 
Many of the unsupported GRASS functions in QGIS relate to geodatabase-management.
Hence, if you strive for the advantages of a geodatabase, we would again  recommend to use GRASS.
Also if you would like to run simulations using GRASS, use **rgrass7** directly (-> include an example from Roger) since **RQGIS** always starts a new GRASS session with each function call.
**RSAGA** provides numerous user-friendly functions (i.e. wrappers).
Secondly, **RSAGA** also offers geocomputing functions written in R (e.g., `multi.local.function`, `pick.from.grid`, etc.).
To sum it up, we like all three packages.
To recommend a single one is hard since the usage depends on personal preferences, the tasks at hand and your familiarity with different GIS.
The latter means if you have already prefer the GUI of a certain GIS, you are quite likely to use the corresponding interface.
Certainly, RQGIS is an appropriate choice for most use cases. Additionally:
- unified access to several GIS, and therefore provision of >1000 geoalgorithms (of course, this includes duplicated functionality, e.g., you can perform overlay-operations using QGIS-, SAGA- or GRASS-geoalgorithms) 
- QGIS automatically takes care of data format conversions
- **RQGIS** can also handle spatial objects residing in R as input for geoalgorithms, and loads QGIS output automatically back into R if desired
- convenience functions to support the access of the online help and R named arguments, automatic default value retrieval though the latter two features are also available in **rgrass7** or to be honest, **rgrass7** features have inspired those now found in RQGIS.
- Currently (but this might change), **RQGIS** supports newer SAGA (2.3.1) versions than **RSAGA** (2.2.3)


```{r gis_comp, echo = FALSE, message = FALSE}
library("tidyverse")
d = tibble("GIS" = c("GRASS", "QGIS", "SAGA"),
            "first release" = c("1984", "2002", "2004"),
            "no. functions" = c(">500", ">1000", ">600"),
            "support" = c("hybrid", "hybrid", "hybrid"))
knitr::kable(x = d, caption = "Comparison between three open-source GIS. Hybrid refers to the support of vector and raster operations.") %>%
  kableExtra::add_footnote(label = "Comparing downloads of different providers is rather difficult (see [http://spatialgalaxy.net/2011/12/19/qgis-users-around-the-world/](http://spatialgalaxy.net/2011/12/19/qgis-users-around-the-world/)), and here also useless since every Windows QGIS download automatically also downloads SAGA and GRASS.", notation = "alphabet")
```

