# Conclusion {#conclusion}

## Prerequisites {-}

Like the introduction, this concluding chapter contains few code chunks.
But its prerequisites are demanding.
It assumes that you have:

- read through and attempted the exercises in all the chapters of Part I (Foundations);
- considered how you can use geocomputation\index{geocomputation} to solve real-world problems, at work and beyond, after engaging with Part III (Applications).

## Introduction

The aim of this chapter is to synthesize the contents, with reference to recurring themes/concepts, and to inspire future directions of application and development.
Section \@ref(package-choice) discusses the wide range of options for handling geographic data in R.
Choice is a key feature of open source software; the section provides guidance on choosing between the various options.
Section \@ref(gaps) describes gaps in the book's contents and explains why some areas of research were deliberately omitted, while others were emphasized.
This discussion leads to the question (which is answered in Section \@ref(next)): having read this book, where to go next?
Section \@ref(benefit) returns to the wider issues raised in Chapter \@ref(intro).
In it we consider geocomputation as part of a wider 'open source approach' that ensures methods are publicly accessible, reproducible\index{reproducibility} and supported by collaborative communities.
This final section of the book also provides some pointers on how to get involved.

## Package choice

A characteristic of R\index{R} is that there are often multiple ways to achieve the same result.
The code chunk below illustrates this by using three functions, covered in Chapters \@ref(attr) and \@ref(geometry-operations), to combine the 16 regions of New Zealand into a single geometry:

```{r 15-synthesis-1}
library(spData)
nz_u1 = sf::st_union(nz)
nz_u2 = aggregate(nz["Population"], list(rep(1, nrow(nz))), sum)
nz_u3 = dplyr::summarise(nz, t = sum(Population))
identical(nz_u1, nz_u2$geometry)
identical(nz_u1, nz_u3$geom)
```

Although the classes, attributes and column names of the resulting objects `nz_u1` to `nz_u3` differ, their geometries are identical.
This is verified using the base R function `identical()`.^[
The first operation, undertaken by the function `st_union()`\index{vector!union}, creates an object of class `sfc` (a simple feature column).
The latter two operations create `sf` objects, each of which *contains* a simple feature column.
Therefore, it is the geometries contained in simple feature columns, not the objects themselves, that are identical.
]
Which to use?
It depends: the former only processes the geometry data contained in `nz` so is faster, while the other options performed attribute operations, which may be useful for subsequent steps.

The wider point is that there are often multiple options to choose from when working with geographic data in R, even within a single package.
The range of options grows further when more R packages are considered: you could achieve the same result using the older **sp** package, for example.
We recommend using **sf** and the other packages showcased in this book, for reasons outlined in Chapter \@ref(spatial-class), but it's worth being aware of alternatives and being able to justify your choice of software.

A common (and sometimes controversial) choice is between **tidyverse**\index{tidyverse (package)} and base R approaches.
We cover both and encourage you to try both before deciding which is more appropriate for different tasks.
The following code chunk, described in Chapter \@ref(attr), shows how attribute data subsetting works in each approach, using the base R operator `[` and the `select()` function from the **tidyverse** package **dplyr**.
The syntax differs but the results are (in essence) the same:

```{r 15-synthesis-2, message=FALSE}
library(dplyr)                          # attach tidyverse package
nz_name1 = nz["Name"]                   # base R approach
nz_name2 = nz |> select(Name)          # tidyverse approach
identical(nz_name1$Name, nz_name2$Name) # check results
```

Again the question arises: which to use?
Again the answer is: it depends.
Each approach has advantages: the pipe syntax is popular and appealing to some, while base R\index{R!base} is more stable, and is well known to others.
Choosing between them is therefore largely a matter of preference.
However, if you do choose to use **tidyverse**\index{tidyverse (package)} functions to handle geographic data, beware of a number of pitfalls (see the supplementary article [`tidyverse-pitfalls`](https://geocompr.github.io/geocompkg/articles/tidyverse-pitfalls.html) on the website that supports this book).

While commonly needed operators/functions were covered in depth --- such as the base R `[` subsetting operator and the **dplyr** function `filter()` --- there are many other functions for working with geographic data, from other packages, that have not been mentioned.
Chapter \@ref(intro) mentions 20+ influential packages for working with geographic data, and only a handful of these are demonstrated in subsequent chapters.
There are hundreds more.
As of mid-2022, there are about 200 packages mentioned in the Spatial [Task View](https://cran.r-project.org/web/views/);
more packages and countless functions for geographic data are developed each year, making it impractical to cover them all in a single book.

```{r 15-synthesis-3, eval=FALSE, echo=FALSE}
# aim: find number of packages in the spatial task view
# how? see:
# vignette("selectorgadget")
stv_pkgs = xml2::read_html("https://cran.r-project.org/web/views/Spatial.html")
pkgs = rvest::html_nodes(stv_pkgs, "ul:nth-child(5) a")
pkgs_char = rvest::html_text(pkgs)
length(pkgs_char)
```

The rate of evolution in R's spatial ecosystem may seem overwhelming, but there are strategies to deal with the wide range of options.
Our advice is to start by learning one approach *in depth* but to have a general understand of the *breadth* of options available.
This advice applies equally to solving geographic problems in R (Section \@ref(next) covers developments in other languages) as it does to other fields of knowledge and application.

Of course, some packages perform much better than others, making package selection an important decision.
From this diversity, we have focused on packages that are future-proof (they will work long into the future), high performance (relative to other R packages) and complementary.
But there is still overlap in the packages we have used, as illustrated by the diversity of packages for making maps, for example (see Chapter \@ref(adv-map)).

Package overlap is not necessarily a bad thing.
It can increase resilience, performance (partly driven by friendly competition and mutual learning between developers) and choice, a key feature of open source software.
In this context the decision to use a particular approach, such as the **sf**/**tidyverse**/**raster** ecosystem advocated in this book should be made with knowledge of alternatives.
The **sp**/**rgdal**/**rgeos** ecosystem that **sf**\index{sf} is designed to supersede, for example, can do many of the things covered in this book and, due to its age, is built on by many other packages.^[
At the time of writing 452 package `Depend` or `Import`  **sp**, showing that its data structures are widely used and have been extended in many directions.
The equivalent number for **sf** was 69 in October 2018; with the growing popularity of **sf**, this is set to grow.
]
Although best known for point pattern analysis, the **spatstat** package also supports raster\index{raster} and other vector geometries [@baddeley_spatstat_2005].
At the time of writing (October 2018) 69 packages depend on it, making it more than a package: **spatstat** is an alternative R-spatial ecosystem.

It is also being aware of promising alternatives that are under development.
The package **stars**, for example, provides a new class system for working with spatiotemporal data.
If you are interested in this topic, you can check for updates on the package's [source code](https://github.com/r-spatial/stars) and the broader [SpatioTemporal Task View](https://cran.r-project.org/web/views/SpatioTemporal.html).
The same principle applies to other domains: it is important to justify software choices and review software decisions based on up-to-date information. 

```{r 15-synthesis-4, echo=FALSE, eval=FALSE}
revdeps_sp = devtools::revdep(pkg = "sp", dependencies = c("Depends", "Imports"))
revdeps_sf = devtools::revdep(pkg = "sf", dependencies = c("Depends", "Imports"))
revdeps_spatstat = devtools::revdep(pkg = "spatstat", dependencies = c("Depends", "Imports"))
```

## Gaps and overlaps {#gaps}

There are a number of gaps in, and some overlaps between, the topics covered in this book.
We have been selective, emphasizing some topics while omitting others.
We have tried to emphasize topics that are most commonly needed in real-world applications such as geographic data operations, projections, data read/write and visualization.
These topics appear repeatedly in the chapters, a substantial area of overlap designed to consolidate these essential skills for geocomputation\index{geocomputation}.

On the other hand, we have omitted topics that are less commonly used, or which are covered in-depth elsewhere.
Statistical topics including point pattern analysis, spatial interpolation (kriging) and spatial epidemiology, for example, are only mentioned with reference to other topics such as the machine learning\index{machine learning} techniques covered in Chapter \@ref(spatial-cv) (if at all).
There is already excellent material on these methods, including statistically orientated chapters in @bivand_applied_2013 and a book on point pattern analysis by @baddeley_spatial_2015.
Other topics which received limited attention were remote sensing and using R alongside (rather than as a bridge to) dedicated GIS\index{GIS} software.
There are many resources on these topics, including @wegmann_remote_2016 and the GIS-related teaching materials available from [Marburg University](https://moc.online.uni-marburg.de/doku.php).

Instead of covering spatial statistical modeling and inference techniques, we focussed on machine learning\index{machine learning} (see Chapters \@ref(spatial-cv) and \@ref(eco)).
Again, the reason was that there are already excellent resources on these topics, especially with ecological use cases, including @zuur_mixed_2009, @zuur_beginners_2017 and freely available teaching material and code on *Geostatistics & Open-source Statistical Computing* by David Rossiter, hosted at [css.cornell.edu/faculty/dgr2](http://www.css.cornell.edu/faculty/dgr2/teach/) and the [*granolarr*](https://sdesabbata.github.io/granolarr/) project by [Stefano De Sabbata](https://stefanodesabbata.com/) [at the University of Leicester](https://www2.le.ac.uk/departments/geography/people/stefano-de-sabbata) for an introduction to R\index{R} for geographic data science\index{data science}.
There are also excellent resources on spatial statistics\index{spatial!statistics} using Bayesian modeling, a powerful framework for modeling and uncertainty estimation [@blangiardo_spatial_2015;@krainski_advanced_2018].

Finally, we have largely omitted big data\index{big data} analytics.
This might seem surprising since especially geographic data can become big really fast. 
But the prerequisite for doing big data analytics is to know how to solve a problem on a small dataset.
Once you have learned that, you can apply the exact same techniques on big data questions, though of course you need to expand your toolbox. 
The first thing to learn is to handle geographic data queries.
This is because big data\index{big data} analytics often boil down to extracting a small amount of data from a database for a specific statistical analysis.
For this, we have provided an introduction to spatial databases\index{spatial database} and how to use a GIS\index{GIS} from within R in Chapter \@ref(gis).
If you really have to do the analysis on a big or even the complete dataset, hopefully, the problem you are trying to solve is embarrassingly parallel.
For this, you need to learn a system that is able to do this parallelization efficiently such as [Apache Sedona](https://sedona.apache.org/), as mentioned in Section \@ref(postgis).
Regardless of dataset size, the techniques and concepts you have used on small datasets will be useful\index{big data} question, the only difference being the extra considterations when working in a big data setting.
<!-- TODO: add reference on big data?-->

## Getting help? {#questions}
<!-- Now wondering if this should be an appendix, or even a new chapter?? -->

<!-- Chapter \@ref(intro) states that the approach advocated in this book "can help remove constraints on your creativity imposed by software". -->
<!-- We have covered many techniques that should enable you to put many of your ideas into reproducible and scalable code for research and applied geocomputation. -->
<!-- However, creativity involves thinking coming up with *new* ideas that have not yet been implemented, raising the question: what happens when software *does* impose a constraint because you are not sure how to implement your creative ideas? -->

<!-- In Chapter \@ref(intro) we set out our aim of providing strong foundations on which a wide range of data analysis, research and methodological and software development projects can build. -->
<!-- Geocomputation is about not only using existing techniques but developing new tools which, by definition, involves generating new knowledge. -->

Geocomputation is a large field and it is highly likely that you will encounter challenges preventing you from achieving an outcome that you are aiming towards.
In many cases you may just 'get stuck' at a particular point in data analysis workflows, with a cryptic error message or an unexpected result providing little clues as to what is going on.
This section provides pointers to help you overcome such problems, by clearly defining the problem, searching for existing knowledge on solutions and, if those approaches to not solve the problem, through the art of asking good questions.
<!-- generating new open knowledge by engaging with the community. -->

When you get stuck at a particular point, it is worth first taking a step back and working out which approach is most likely to solve the issue.
Trying each of the following steps, in order (or skipping steps if you have already tried them), provides a structured approach to problem solving:

1. Define exactly what you are trying to achieve, starting from first principles (and often a sketch, as outlined below)
2. Diagnose exactly where in your code the unexpected results arise, by running and exploring the outputs of individual lines of code and their individual components (you can can run individual parts of a complex command by selecting them with a cursor and pressing Ctrl+Enter in RStudio, for example)
3. Read the documentation of the function that has been diagnosed as the 'point of failure' in the previous step. Simply understanding the required inputs to functions, and running the examples that are often provided at the bottom of help pages, can help solve a surprisingly large proportion of issues (run the command `?terra::rast` and scroll down to the examples that are worth reproducing when getting started with the function, for example)
4. If reading R's inbuilt documentation, as outlined in the previous step, does not help solve the problem, it is probably time to do a broader search online to see if others have written about the issue you're seeing. See a list of places to search for help below for places to search
5. If all the previous steps above fail, and you cannot find a solution from your online searches, it may be time to compose a question with a reproducible example and post it in an appropriate place

Steps 1 to 3 outlined above are fairly self-explanatory but, due to the vastness of the internet and multitude of search options, it is worth considering effective search strategies before deciding to compose a question.

### Searching for solutions online

A logical place to start for many issues is search engines.
'Googling it' can in some cases result in the discovery of blog posts, forum messages and other online content about the precise issue you're having.
Simply typing in a clear description of the problem/question is a valid approach here but it is important to be specific (e.g. with reference to function and package names and input dataset sources if the problem is dataset specific).
You can also make online searches more effective by including additional detail:
<!-- To provide a concrete example, imagine you want to know how to use custom symbols in an interactive map. -->

- Use quote marks to maximise the chances that 'hits' relate to the exact issue you're having by reducing the number of results returned
<!-- todo: add example -->
- Set [time restraints](https://uk.pcmag.com/software-services/138320/21-google-search-tips-youll-want-to-learn), for example only returning content created within the last year can be useful when searching for help on an evolving package.
- Make use of additional [search engine features](https://www.makeuseof.com/tag/6-ways-to-search-by-date-on-google/), for example restricting searches to content hosted on CRAN with site:r-project.org

### Places to search for (and ask) for help {#help}

<!-- toDo:rl-->
<!-- Todo: provide pros and cons and maybe how to search each:  -->
- R's Special Interest Group on Geographic data email list ([R-SIG-GEO](https://stat.ethz.ch/mailman/listinfo/r-sig-geo))
- The GIS Stackexchange website at [gis.stackexchange.com](https://gis.stackexchange.com/)
- The large and general purposes programming Q&A site [stackoverflow.com](https://stackoverflow.com/)
- Online forums associated with a particular entity, such as the [RStudio Community](https://community.rstudio.com/), the [rOpenSci Discuss](https://discuss.ropensci.org/) web forum and forums associated with particular software tools such as the [Stan](https://discourse.mc-stan.org/) forum
- Software development platforms such as GitHub, which hosts issue trackers for the majority of R-spatial packages and also, increasingly, inbuilt discussion pages such as that created to encourage discussion (not just bug reporting) around the **sfnetworks** package (see [luukvdmeer/sfnetworks/discussions](https://github.com/luukvdmeer/sfnetworks/discussions/))

### How to ask a good question with a reproducible example {#reprex}

In terms asking a good question, a clearly stated questions supported by an accessible and fully reproducible example is key.
It is also helpful, after showing the code that 'did not work' from the user's perspective, to explain what you would like to see.
A very useful tool for creating reproducible examples is the **reprex** package.
<!-- Todo: show how reprex works. -->



<!-- A strength of open source and collaborative approaches to geocomputation is that they generate a vast and ever evolving body on knowledge, of which this book is a part. -->
<!-- Thousands of exchanges have taken place in publicly accessible fora, demonstrating the importance of knowing how to search for answers and, perhaps more importantly, show how beginners can support open source software communities by asking good questions. -->
<!-- This section covers these interrelated topics, with a focus on common places to search for answers and ask questions and how to ask good questions. -->
<!-- Should we divide these topics in 2? RL 2022-02 -->

<!-- Key fora for discussing methods and code for working with geographic data in R include: -->
<!-- I was thinking of saying "in descening order of ease of use" or something but not sure that's a good idea (RL 2022-02) -->

### Defining and sketching the problem

The best starting point when developing a new geocomputational methodology or approach is often a pen and paper (or equivalent digital sketching tools such as [Excalidraw](https://excalidraw.com/) and [tldraw](https://www.tldraw.com/) which allow collaborative sketching and rapid sharing of ideas): during the most creative early stages of methodological development work software *of any kind* can slow down your thoughts and direct your thinking away from important abstract thoughts.
Framing the question with mathematics is also highly recommended, with reference to a minimal example that you can sketch 'before and after' versions of numerically.
If you have the skills and if the problem warrants it, describing the approach algebraically can in some cases help develop effective implementations.

## Where to go next? {#next}

As indicated in Section \@ref(gaps), the book has covered only a fraction of the R's geographic ecosystem, and there is much more to discover.
We have progressed quickly, from geographic data models in Chapter \@ref(spatial-class), to advanced applications in Chapter \@ref(eco).
Consolidation of skills learned, discovery of new packages and approaches for handling geographic data, and application of the methods to new datasets and domains are suggested future directions.
This section expands on this general advice by suggesting specific 'next steps', highlighted in **bold** below.

In addition to learning about further geographic methods and applications with R\index{R}, for example with reference to the work cited in the previous section, deepening your understanding of **R itself** is a logical next step.
R's fundamental classes such as `data.frame` and `matrix` are the foundation of `sf` and `raster` classes, so studying them will improve your understanding of geographic data.
This can be done with reference to documents that are part of R, and which can be found with the command `help.start()` and additional resources on the subject such as those by @wickham_advanced_2019 and @chambers_extending_2016.

Another software-related direction for future learning is **discovering geocomputation with other languages**.
There are good reasons for learning R as a language for geocomputation, as described in Chapter \@ref(intro), but it is not the only option.^[
R's strengths relevant to our definition of geocomputation include its emphasis on scientific reproducibility\index{reproducibility}, widespread use in academic research and unparalleled support for statistical modeling of geographic data.
Furthermore, we advocate learning one language (R) for geocomputation\index{geocomputation} in depth before delving into other languages/frameworks because of the costs associated with context switching.
It is preferable to have expertise in one language than basic knowledge of many.
]
It would be possible to study *Geocomputation with: Python*\index{Python}, *C++*\index{C++}, *JavaScript*\index{JavaScript}, *Scala*\index{Scala} or *Rust*\index{Rust} in equal depth.
Each has evolving geospatial capabilities.
[**rasterio**](https://github.com/mapbox/rasterio), for example, is a Python package
that could supplement/replace the **raster** package used in this book --- see @garrard_geoprocessing_2016 and online tutorials such as [automating-gis-processes](https://automating-gis-processes.github.io/CSC18) for more on the Python\index{Python} ecosystem.
Dozens of geospatial libraries have been developed in C++\index{C++}, including well known libraries such as GDAL\index{GDAL} and GEOS\index{GEOS}, and less well known libraries such as the **[Orfeo Toolbox](https://github.com/orfeotoolbox/OTB)** for processing remote sensing (raster) data.
[**Turf.js**](https://github.com/Turfjs/turf) is an example of the potential for doing geocomputation with JavaScript.
\index{Scala}
\index{JavaScript}
[GeoTrellis](https://geotrellis.io/) provides functions for working with raster and vector data in the Java-based language Scala.
And [WhiteBoxTools](https://github.com/jblindsay/whitebox-tools) provides an example of a rapidly evolving command-line GIS implemented in Rust.
\index{Rust}
\index{WhiteboxTools}
Each of these packages/libraries/languages has advantages for geocomputation and there are many more to discover, as documented in the curated list of open source geospatial resources [Awesome-Geospatial](https://github.com/sacridini/Awesome-Geospatial).

There is more to geocomputation\index{geocomputation} than software, however.
We can recommend **exploring and learning new research topics and methods** from academic and theoretical perspectives.
Many methods that have been written about have yet to be implemented.
Learning about geographic methods and potential applications can therefore be rewarding, before writing any code.
An example of geographic methods that are increasingly implemented in R is sampling strategies for scientific applications.
A next step in this case is to read-up on relevant articles in the area such as @brus_sampling_2018, which is accompanied by reproducible code and tutorial content hosted at [github.com/DickBrus/TutorialSampling4DSM](https://github.com/DickBrus/TutorialSampling4DSM).

## The open source approach {#benefit}

This is a technical book so it makes sense for the next steps, outlined in the previous section, to also be technical.
However, there are wider issues worth considering in this final section, which returns to our definition of geocomputation\index{geocomputation}.
One of the elements of the term introduced in Chapter \@ref(intro) was that geographic methods should have a positive impact.
Of course, how to define and measure 'positive' is a subjective, philosophical question, beyond the scope of this book.
Regardless of your worldview, consideration of the impacts of geocomputational work is a useful exercise:
the potential for positive impacts can provide a powerful motivation for future learning and, conversely, new methods can open-up many possible fields of application.
These considerations lead to the conclusion that geocomputation is part of a wider 'open source approach'.

Section \@ref(what-is-geocomputation) presented other terms that mean roughly the same thing as geocomputation, including geographic data science\index{data science} (GDS) and 'GIScience'.
Both capture the essence of working with geographic data, but geocomputation has advantages: it concisely captures the 'computational' way of working with geographic data advocated in this book --- implemented in code and therefore encouraging reproducibility --- and builds on desirable ingredients of its early definition [@openshaw_geocomputation_2000]:

- The *creative* use of geographic data
- Application to *real-world problems*
- Building 'scientific' tools
- Reproducibility\index{reproducibility}

We added the final ingredient: reproducibility was barely mentioned in early work on geocomputation, yet a strong case can be made for it being a vital component of the first two ingredients.
Reproducibility\index{reproducibility}

- encourages *creativity* by shifting the focus away from the basics (which are readily available through shared code) and towards applications;
- discourages people from 'reinventing the wheel': there is no need to re-do what others have done if their methods can be used by others; and
- makes research more conducive to real world applications, by enabling anyone in any sector to apply your methods in new areas.

If reproducibility is the defining asset of geocomputation (or command-line GIS) it is worth considering what makes it reproducible.
This brings us to the 'open source approach', which has three main components:

- A command-line interface\index{command-line interface} (CLI), encouraging scripts recording geographic work to be shared and reproduced
- Open source software, which can be inspected and potentially improved by anyone in the world
- An active developer community, which collaborates and self-organizes to build complementary and modular tools

Like the term geocomputation\index{geocomputation}, the open source approach is more than a technical entity.
It is a community composed of people interacting daily with shared aims: to produce high performance tools, free from commercial or legal restrictions, that are accessible for anyone to use.
The open source approach to working with geographic data has advantages that transcend the technicalities of how the software works, encouraging learning, collaboration and an efficient division of labor.

There are many ways to engage in this community, especially with the emergence of code hosting sites, such as GitHub, which encourage communication and collaboration.
A good place to start is simply browsing through some of the source code, 'issues' and 'commits' in a geographic package of interest.
A quick glance at the `r-spatial/sf` GitHub repository, which hosts the code underlying the **sf**\index{sf} package, shows that 40+ people have contributed to the codebase and documentation.
Dozens more people have contributed by asking question and by contributing to 'upstream' packages that **sf** uses.
More than 600 issues have been closed on its [issue tracker](https://github.com/r-spatial/sf/issues), representing a huge amount of work to make **sf** faster, more stable and user-friendly.
This example, from just one package out of dozens, shows the scale of the intellectual operation underway to make R a highly effective and continuously evolving language for geocomputation.

It is instructive to watch the incessant development activity happen in public fora such as GitHub, but it is even more rewarding to become an active participant.
This is one of the greatest features of the open source approach: it encourages people to get involved.
This book itself is a result of the open source approach:
it was motivated by the amazing developments in R's geographic capabilities over the last two decades, but made practically possible by dialogue and code sharing on platforms for collaboration.
We hope that in addition to disseminating useful methods for working with geographic data, this book inspires you to take a more open source approach.
Whether it's raising a constructive issue alerting developers to problems in their package; making the work done by you and the organizations you work for open; or simply helping other people by passing on the knowledge you've learned, getting involved can be a rewarding experience.

