
1. Use `summary()` on the geometry column of the `world` data object. What does the output tell us about:
    - Its geometry type?
    - The number of countries?
    - Its coordinate reference system (CRS)?
1. Run the code that 'generated' the map of the world in Figure \@ref(fig:contpop) at the end of Section \@ref(base-args).
Find two similarities and two differences between the image on your computer and that in the book.
    - What does the `cex` argument do (see `?plot`)?
    - Why was `cex` set to the `sqrt(world$pop) / 10000`?
    - Bonus: experiment with different ways to visualize the global population.
1. Use `plot()` to create maps of Nigeria in context (see Section \@ref(base-args)).
    - Adjust the `lwd`, `col` and `expandBB` arguments of `plot()`. 
    - Challenge: read the documentation of `text()` and annotate the map.
1. Create an empty `SpatRaster` object called `my_raster` with 10 columns and 10 rows.
Assign random values between 0 and 10 to the new raster and plot it.

```{r, message = FALSE}
library(terra)
my_raster = rast(ncol = 10, nrow = 10,
                 vals = sample(0:10, size = 10 * 10, replace = TRUE))
```


1. Read-in the `raster/nlcd.tif` file from the **spDataLarge** package. 
What kind of information can you get about the properties of this file?

```{r, message = FALSE}
library(terra)
nlcd = rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
dim(nlcd)
res(nlcd)
ext(nlcd)
crs(nlcd)
```

