
Some of the exercises use a vector (`zion_points`) and raster dataset (`srtm`) from the **spDataLarge** package.
They also use a polygonal 'convex hull' derived from the vector dataset (`ch`) to represent the area of interest:

```{r 05-geometry-operations-68, message=FALSE, include=TRUE}
library(sf)
library(terra)
library(spData)
zion_points = read_sf(system.file("vector/zion_points.gpkg", package = "spDataLarge"))
srtm = rast(system.file("raster/srtm.tif", package = "spDataLarge"))
ch = st_combine(zion_points) %>%
  st_convex_hull() %>% 
  st_as_sf()
```

E1. Generate and plot simplified versions of the `nz` dataset.
Experiment with different values of `keep` (ranging from 0.5 to 0.00005) for `ms_simplify()` and `dTolerance` (from 100 to 100,000) `st_simplify()`.

- At what value does the form of the result start to break down for each method, making New Zealand unrecognizable?
- Advanced: What is different about the geometry type of the results from `st_simplify()` compared with the geometry type of `ms_simplify()`? What problems does this create and how can this be resolved?

```{r 05-geometry-operations-69, eval=FALSE}
plot(rmapshaper::ms_simplify(st_geometry(nz), keep = 0.5))
plot(rmapshaper::ms_simplify(st_geometry(nz), keep = 0.05))
# Starts to breakdown here at 0.5% of the points:
plot(rmapshaper::ms_simplify(st_geometry(nz), keep = 0.005))
# At this point no further simplification changes the result
plot(rmapshaper::ms_simplify(st_geometry(nz), keep = 0.0005))
plot(rmapshaper::ms_simplify(st_geometry(nz), keep = 0.00005))
plot(st_simplify(st_geometry(nz), dTolerance = 100))
plot(st_simplify(st_geometry(nz), dTolerance = 1000))
# Starts to breakdown at 10 km:
plot(st_simplify(st_geometry(nz), dTolerance = 10000))
plot(st_simplify(st_geometry(nz), dTolerance = 100000))
plot(st_simplify(st_geometry(nz), dTolerance = 100000, preserveTopology = TRUE))

# Problem: st_simplify returns POLYGON and MULTIPOLYGON results, affecting plotting
# Cast into a single geometry type to resolve this
nz_simple_poly = st_simplify(st_geometry(nz), dTolerance = 10000) %>% 
  st_sfc() %>% 
  st_cast("POLYGON")
nz_simple_multipoly = st_simplify(st_geometry(nz), dTolerance = 10000) %>% 
  st_sfc() %>% 
  st_cast("MULTIPOLYGON")
plot(nz_simple_poly)
length(nz_simple_poly)
nrow(nz)
```

E2. In the first exercise in Chapter Spatial data operations it was established that Canterbury region had 70 of the 101 highest points in New Zealand. 
Using `st_buffer()`, how many points in `nz_height` are within 100 km of Canterbury?

```{r 05-geometry-operations-70, eval=FALSE}
canterbury = nz[nz$Name == "Canterbury", ]
cant_buff = st_buffer(canterbury, 100)
nz_height_near_cant = nz_height[cant_buff, ]
nrow(nz_height_near_cant) # 75 - 5 more
```

E3. Find the geographic centroid of New Zealand. 
How far is it from the geographic centroid of Canterbury?

```{r 05-geometry-operations-71, eval=FALSE}
cant_cent = st_centroid(canterbury)
nz_centre = st_centroid(st_union(nz))
st_distance(cant_cent, nz_centre) # 234 km
```

E4. Most world maps have a north-up orientation.
A world map with a south-up orientation could be created by a reflection (one of the affine transformations not mentioned in this chapter) of the `world` object's geometry.
Write code to do so.
Hint: you need to use a two-element vector for this transformation.
 Bonus: create an upside-down map of your country.
 
```{r 05-geometry-operations-72, eval=FALSE}
world_sfc = st_geometry(world)
world_sfc_mirror = world_sfc * c(1, -1)
plot(world_sfc)
plot(world_sfc_mirror)

us_states_sfc = st_geometry(us_states)
us_states_sfc_mirror = us_states_sfc * c(1, -1)
plot(us_states_sfc)
plot(us_states_sfc_mirror)
## nicer plot
library(ggrepel)
us_states_sfc_mirror_labels = st_centroid(us_states_sfc_mirror) %>% 
  st_coordinates() %>%
  as_data_frame() %>% 
  mutate(name = us_states$NAME)
us_states_sfc_mirror_sf = st_set_geometry(us_states, us_states_sfc_mirror)
ggplot(data = us_states_sfc_mirror_sf) +
  geom_sf(color = "white") +
  geom_text_repel(data = us_states_sfc_mirror_labels, mapping = aes(X, Y, label = name), size = 3, min.segment.length = 0) +
  theme_void() 
```

E5. Subset the point in `p` that is contained within `x` *and* `y`.

- Using base subsetting operators.
- Using an intermediary object created with `st_intersection()`\index{vector!intersection}.

```{r 05-geometry-operations-73, eval=FALSE}
p_in_y = p[y]
p_in_xy = p_in_y[x]
x_and_y = st_intersection(x, y)
p[x_and_y]
```

E6. Calculate the length of the boundary lines of US states in meters.
Which state has the longest border and which has the shortest?
Hint: The `st_length` function computes the length of a `LINESTRING` or `MULTILINESTRING` geometry.

```{r 05-geometry-operations-74, eval=FALSE}
us_states2163 = st_transform(us_states, "EPSG:2163")
us_states_bor = st_cast(us_states2163, "MULTILINESTRING")
us_states_bor$borders = st_length(us_states_bor)
arrange(us_states_bor, borders)
arrange(us_states_bor, -borders)
```

E7. Crop the `srtm` raster using (1) the `zion_points` dataset and (2) the `ch` dataset.
Are there any differences in the output maps?
Next, mask `srtm` using these two datasets.
Can you see any difference now?
How can you explain that?

```{r 05-geometry-operations-75}
plot(srtm)
plot(st_geometry(zion_points), add = TRUE)
plot(ch, add = TRUE)

srtm_crop1 = crop(srtm, vect(zion_points))
srtm_crop2 = crop(srtm, vect(ch))
plot(srtm_crop1)
plot(srtm_crop2)

srtm_mask1 = mask(srtm, vect(zion_points))
srtm_mask2 = mask(srtm, vect(ch))
plot(srtm_mask1)
plot(srtm_mask2)
```

E8. Firstly, extract values from `srtm` at the points represented in `zion_points`.
Next, extract average values of `srtm` using a 90 buffer around each point from `zion_points` and compare these two sets of values. 
When would extracting values by buffers be more suitable than by points alone?

```{r 05-geometry-operations-76}
zion_points_buf = st_buffer(zion_points, dist = 90)
plot(srtm)
plot(st_geometry(zion_points_buf), add = TRUE)
plot(ch, add = TRUE)

zion_points_points = extract(srtm, vect(zion_points))
zion_points_buf = extract(srtm, vect(zion_points_buf))
plot(zion_points_points$srtm, zion_points_buf$srtm2)
```

E9. Subset points higher than 3100 meters in New Zealand (the `nz_height` object) and create a template raster with a resolution of 3 km for the extent of the new point dataset. 
Using these two new objects:

- Count numbers of the highest points in each grid cell.
- Find the maximum elevation in each grid cell.

```{r 05-geometry-operations-77}
nz_height3100 = dplyr::filter(nz_height, elevation > 3100)
new_graticule = st_graticule(nz_height3100, datum = "EPSG:2193")
plot(st_geometry(nz_height3100), graticule = new_graticule, axes = TRUE)

nz_template = rast(ext(nz_height3100), resolution = 3000, crs = crs(nz_height3100))

nz_raster = rasterize(vect(nz_height3100), nz_template, 
                       field = "elevation", fun = "length")
plot(nz_raster)
plot(st_geometry(nz_height3100), add = TRUE)

nz_raster2 = rasterize(vect(nz_height3100), nz_template, 
                       field = "elevation", fun = max)
plot(nz_raster2)
plot(st_geometry(nz_height3100), add = TRUE)
```

E10. Aggregate the raster counting high points in New Zealand (created in the previous exercise), reduce its geographic resolution by half (so cells are 6 by 6 km) and plot the result.

- Resample the lower resolution raster back to the original resolution of 3 km. How have the results changed?
- Name two advantages and disadvantages of reducing raster resolution.

```{r 05-geometry-operations-78}
nz_raster_low = raster::aggregate(nz_raster, fact = 2, fun = sum, na.rm = TRUE)
res(nz_raster_low)

nz_resample = resample(nz_raster_low, nz_raster)
plot(nz_raster_low)
plot(nz_resample) # the results are spread over a greater area and there are border issues
plot(nz_raster)
```

```{asis}
Advantages:

- lower memory use
- faster processing
- good for viz in some cases

Disadvantages:

- removes geographic detail
- adds another processing step
```

E11. Polygonize the `grain` dataset and filter all squares representing clay.

```{r}
grain = rast(system.file("raster/grain.tif", package = "spData"))
```

- Name two advantages and disadvantages of vector data over raster data.
- When would it be useful to convert rasters to vectors in your work?

```{r 05-geometry-operations-79}
grain_poly = as.polygons(grain) %>% 
  st_as_sf()
levels(grain)
clay = dplyr::filter(grain_poly, grain == "clay")
plot(clay)
```

```{asis}
Advantages: 

- can be used to subset other vector objects
- can do affine transformations and use sf/dplyr verbs

Disadvantages: 

- better consistency
- fast processing on some operations
- functions developed for some domains
```

