```{r 08-ex-e0, message=FALSE}
library(sf)
library(terra)
```

E1. List and describe three types of vector, raster, and geodatabase formats.

```{asis 08-ex-e0-asis}
Vector formats: Shapefile (old format supported by many programs), GeoPackage (more recent format with better support of attribute data) and GeoJSON (common format for web mapping).

Raster formats: GeoTiff, Arc ASCII, R-raster (see book for descriptions).

Database formats: PostGIS, SQLite, FileGDB (see book for details).
```


E2. Name at least two differences between the **sf** functions `read_sf()` and `st_read()`.

```{asis 08-ex-e2-asis}
`st_read()` prints outputs and keeps strings as text strings (`st_read()` creates factors). This can be seen from the source code of `read_sf()`, which show's it wraps `st_read()`:
```

```{r 08-ex-e2}
read_sf
```

E3. Read the `cycle_hire_xy.csv` file from the **spData** package as a spatial object (Hint: it is located in the `misc` folder).
What is a geometry type of the loaded object? 

```{r 08-ex-e3}
c_h = read.csv(system.file("misc/cycle_hire_xy.csv", package = "spData")) |> 
  st_as_sf(coords = c("X", "Y"))
c_h
```

E4. Download the borders of Germany using **rnaturalearth**, and create a new object called `germany_borders`.
Write this new object to a file of the GeoPackage format.

```{r 08-ex-e4}
library(rnaturalearth)
germany_borders = ne_countries(country = "Germany", returnclass = "sf")
plot(germany_borders)
st_write(germany_borders, "germany_borders.gpkg")
```

E5. Download the global monthly minimum temperature with a spatial resolution of five minutes using the **geodata** package.
Extract the June values, and save them to a file named `tmin_june.tif` file (hint: use `terra::subset()`).

```{r 08-ex-e5}
library(geodata)
gmmt = worldclim_global(var = "tmin", res = 5, path = tempdir())
names(gmmt)
plot(gmmt)

gmmt_june = terra::subset(gmmt, "wc2.1_5m_tmin_06")
plot(gmmt_june)
writeRaster(gmmt_june, "tmin_june.tif")
```

E6. Create a static map of Germany's borders, and save it to a PNG file.

```{r 08-ex-e6}
png(filename = "germany.png", width = 350, height = 500)
plot(st_geometry(germany_borders), axes = TRUE, graticule = TRUE)
dev.off()
```

E7. Create an interactive map using data from the `cycle_hire_xy.csv` file. 
Export this map to a file called `cycle_hire.html`.

```{r 08-ex-e7, eval=FALSE}
library(mapview)
mapview_obj = mapview(c_h, zcol = "nbikes", legend = TRUE)
mapshot(mapview_obj, file = "cycle_hire.html")
```
