These exercises rely on a new object, `africa`.
Create it using the `world` and `worldbank_df` datasets from the **spData** package as follows (see Chapter \@ref(attr)):

```{r 08-mapping-41, warning=FALSE}
africa = world |> 
  filter(continent == "Africa", !is.na(iso_a2)) |> 
  left_join(worldbank_df, by = "iso_a2") |> 
  dplyr::select(name, subregion, gdpPercap, HDI, pop_growth) |> 
  st_transform("+proj=aea +lat_1=20 +lat_2=-23 +lat_0=0 +lon_0=25")
```

We will also use `zion` and `nlcd` datasets from **spDataLarge**:

```{r 08-mapping-42, results='hide'}
zion = st_read((system.file("vector/zion.gpkg", package = "spDataLarge")))
data(nlcd, package = "spDataLarge")
```

1. Create a map showing the geographic distribution of the Human Development Index (`HDI`) across Africa with base **graphics** (hint: use `plot()`) and **tmap** packages (hint: use `tm_shape(africa) + ...`).
    - Name two advantages of each based on the experience.
    - Name three other mapping packages and an advantage of each.
    - Bonus: create three more maps of Africa using these three packages.
1. Extend the **tmap** created for the previous exercise so the legend has three bins: "High" (`HDI` above 0.7), "Medium" (`HDI` between 0.55 and 0.7) and "Low" (`HDI` below 0.55).
    - Bonus: improve the map aesthetics, for example by changing the legend title, class labels and color palette.
1. Represent `africa`'s subregions on the map. 
Change the default color palette and legend title.
Next, combine this map and the map created in the previous exercise into a single plot.
1. Create a land cover map of the Zion National Park.
    - Change the default colors to match your perception of the land cover categories
    - Add a scale bar and north arrow and change the position of both to improve the map's aesthetic appeal
    - Bonus: Add an inset map of Zion National Park's location in the context of the Utah state. (Hint: an object representing Utah can be subset from the `us_states` dataset.) 
1. Create facet maps of countries in Eastern Africa:
    - With one facet showing HDI and the other representing population growth (hint: using variables `HDI` and `pop_growth`, respectively)
    - With a 'small multiple' per country
<!--comment to the next q:  worldbank_df along the iso_a2 variable is pegging Kosovo and South Cyprus to Somaliland -->
1. Building on the previous facet map examples, create animated maps of East Africa:
    - Showing first the spatial distribution of HDI scores then population growth
    - Showing each country in order
1. Create an interactive map of Africa:
    - With **tmap**
    - With **mapview**
    - With **leaflet**
    - Bonus: For each approach, add a legend (if not automatically provided) and a scale bar
1. Sketch on paper ideas for a web mapping app that could be used to make transport or land-use policies more evidence based:
    - In the city you live, for a couple of users per day
    - In the country you live, for dozens of users per day
    - Worldwide for hundreds of users per day and large data serving requirements
1. Update the code in `coffeeApp/app.R` so that instead of centering on Brazil the user can select which country to focus on:
    - Using `textInput()`
    - Using `selectInput()`
1. Reproduce Figure \@ref(fig:tmshape) and the 1st and 6th panel of Figure \@ref(fig:break-styles) as closely as possible using the **ggplot2** package.
1. Join `us_states` and `us_states_df` together and calculate a poverty rate for each state using the new dataset.
Next, construct a continuous area cartogram based on total population. 
Finally, create and compare two maps of the poverty rate: (1) a standard choropleth map and (2) a map using the created cartogram boundaries.
What is the information provided by the first and the second map?
How do they differ from each other?
1. Visualize population growth in Africa. 
Next, compare it with the maps of a hexagonal and regular grid created using the **geogrid** package.
