```{r 09-ex-e0, message=FALSE}
library(sf)
library(terra)
library(dplyr)
library(spData)
```

These exercises rely on a new object, `africa`.
Create it using the `world` and `worldbank_df` datasets from the **spData** package as follows:

```{r 08-mapping-41, warning=FALSE, include=TRUE}
library(spData)
africa = world |> 
  filter(continent == "Africa", !is.na(iso_a2)) |> 
  left_join(worldbank_df, by = "iso_a2") |> 
  select(name, subregion, gdpPercap, HDI, pop_growth) |> 
  st_transform("ESRI:102022") |> 
  st_make_valid() |> 
  st_collection_extract("POLYGON")
```

We will also use `zion` and `nlcd` datasets from **spDataLarge**:

```{r 08-mapping-42, results='hide', include=TRUE}
zion = read_sf((system.file("vector/zion.gpkg", package = "spDataLarge")))
nlcd = rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
```

E1. Create a map showing the geographic distribution of the Human Development Index (`HDI`) across Africa with base **graphics** (hint: use `plot()`) and **tmap** packages (hint: use `tm_shape(africa) + ...`).

- Name two advantages of each based on the experience.
- Name three other mapping packages and an advantage of each.
- Bonus: create three more maps of Africa using these three other packages.

```{r}
# graphics
plot(africa["HDI"])
# # tmap
# remotes::install_github("r-tmap/tmap")
library(tmap)
tm_shape(africa) + 
  tm_polygons("HDI")
# ggplot
library(ggplot2)
ggplot() +
  geom_sf(data = africa, aes(fill = HDI))
# ggplotly
library(plotly)
g = ggplot() +
  geom_sf(data = africa, aes(fill = HDI))
ggplotly(g)
# mapsf
library(mapsf)
mf_map(x = africa, var = "HDI", type = "choro")
```

E2. Extend the **tmap** created for the previous exercise so the legend has three bins: "High" (`HDI` above 0.7), "Medium" (`HDI` between 0.55 and 0.7) and "Low" (`HDI` below 0.55). 
Bonus: improve the map aesthetics, for example by changing the legend title, class labels and color palette.

```{r}
library(tmap)
tm_shape(africa) + 
  tm_polygons("HDI",
              fill.scale = tm_scale_intervals(breaks = c(0, 0.55, 0.7, 1),
                                              labels = c("Low", "Medium", "High"),
                                              values = "-viridis"),
              fill.legend = tm_legend(title = "Human Development Index")) 
```

E3. Represent `africa`'s subregions on the map. 
Change the default color palette and legend title.
Next, combine this map and the map created in the previous exercise into a single plot.

```{r}
asubregions = tm_shape(africa) +
  tm_polygons("subregion",
              fill.scale = tm_scale_categorical(values = "Set3"),
              fill.legend = tm_legend(title = "Subregion:"))
ahdi = tm_shape(africa) + 
  tm_polygons("HDI",
              fill.scale = tm_scale_intervals(breaks = c(0, 0.55, 0.7, 1),
                                              labels = c("Low", "Medium", "High"),
                                              values = "-viridis"),
              fill.legend = tm_legend(title = "Human Development Index:")) 
tmap_arrange(ahdi, asubregions)
```

E4. Create a land cover map of Zion National Park.

- Change the default colors to match your perception of the land cover categories
- Add a scale bar and north arrow and change the position of both to improve the map's aesthetic appeal
- Bonus: Add an inset map of Zion National Park's location in the context of the state of Utah. (Hint: an object representing Utah can be subset from the `us_states` dataset.) 

```{r}
tm_shape(nlcd) +
  tm_raster(col.scale = tm_scale_categorical(values = c("#495EA1", "#AF5F63", "#EDE9E4",
                                                        "#487F3F", "#EECFA8", "#A4D378",
                                                        "#FFDB5C", "#72D593"), levels.drop = TRUE)) +
  tm_scalebar(bg.color = "white", position = c("left", "bottom")) +
  tm_compass(bg.color = "white", position = c("right", "top")) +
  tm_layout(legend.position = c("left", "top"), legend.bg.color = "white")
```

```{r}
# Bonus
utah = subset(us_states, NAME == "Utah")
utah = st_transform(utah, st_crs(zion))

zion_region = st_bbox(zion) |> 
  st_as_sfc()

main = tm_shape(nlcd) +
  tm_raster(col.scale = tm_scale_categorical(values = c("#495EA1", "#AF5F63", "#EDE9E4",
                                                        "#487F3F", "#EECFA8", "#A4D378",
                                                        "#FFDB5C", "#72D593"), levels.drop = TRUE)) +
  tm_scalebar(bg.color = "white", position = c("left", "bottom")) +
  tm_compass(bg.color = "white", position = c("right", "top")) +
  tm_layout(legend.position = c("left", "top"), legend.bg.color = "white")

inset = tm_shape(utah) +
  tm_polygons() +
  tm_text("UTAH", size = 3) +
  #tm_shape(zion) +
  #tm_polygons(col = "red") +
  tm_shape(zion_region) +
  tm_borders(col = "red") +
  tm_layout(frame = FALSE)

library(grid)
norm_dim = function(obj){
    bbox = st_bbox(obj)
    width = bbox[["xmax"]] - bbox[["xmin"]]
    height = bbox[["ymax"]] - bbox[["ymin"]]
    w = width / max(width, height)
    h = height / max(width, height)
    return(unit(c(w, h), "snpc"))
}
main_dim = norm_dim(zion)
ins_dim = norm_dim(utah)

main_vp = viewport(width = main_dim[1], height = main_dim[2])
ins_vp = viewport(width = ins_dim[1] * 0.4, height = ins_dim[2] * 0.4,
                  x = unit(1, "npc") - unit(0.5, "cm"), y = unit(0.5, "cm"),
                  just = c("right", "bottom"))

grid.newpage()
print(main, vp = main_vp)
pushViewport(main_vp)
print(inset, vp = ins_vp)
```

E5. Create facet maps of countries in Eastern Africa:

- With one facet showing HDI and the other representing population growth (hint: using variables `HDI` and `pop_growth`, respectively)
- With a 'small multiple' per country

```{r}
ea = subset(africa, subregion == "Eastern Africa")
#1
tm_shape(ea) +
  tm_polygons(c("HDI", "pop_growth"))
#2
tm_shape(ea) +
  tm_polygons() +
  tm_facets_wrap("name")
```

E6. Building on the previous facet map examples, create animated maps of East Africa:

- Showing each country in order
- Showing each country in order with a legend showing the HDI

```{r, eval=FALSE}
tma1 = tm_shape(ea) +
  tm_polygons() +
  tm_facets(by = "name", nrow = 1, ncol = 1)
tmap_animation(tma1, filename = "tma2.gif", width = 1000, height = 1000)
browseURL("tma1.gif")

tma2 = tm_shape(africa) +
  tm_polygons(fill = "lightgrey") +
  tm_shape(ea) +
  tm_polygons(fill = "darkgrey") +
  tm_shape(ea) +
  tm_polygons(fill = "HDI") +
  tm_facets(by = "name", nrow = 1, ncol = 1)
tmap_animation(tma2, filename = "tma2.gif", width = 1000, height = 1000)
browseURL("tma2.gif")
```

E7. Create an interactive map of HDI in Africa:

- With **tmap**
- With **mapview**
- With **leaflet**
- Bonus: For each approach, add a legend (if not automatically provided) and a scale bar

```{r, eval=FALSE}
# tmap
tmap_mode("view")
tm_shape(africa) + tm_polygons("HDI") + tm_scalebar()
# mapview
mapview::mapview(africa["HDI"])
# leaflet
africa4326 = st_transform(africa, "EPSG:4326")
library(leaflet)
pal = colorNumeric(palette = "YlGnBu", domain = africa4326$HDI)
leaflet(africa4326) |> 
  addTiles() |> 
  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1, color = ~pal(HDI)) |> 
  addLegend("bottomright", pal = pal, values = ~HDI, opacity = 1) |> 
  addScaleBar()
```

E8. Sketch on paper ideas for a web-mapping application that could be used to make transport or land-use policies more evidence-based:

- In the city you live, for a couple of users per day
- In the country you live, for dozens of users per day
- Worldwide for hundreds of users per day and large data serving requirements

```{asis}
Ideas could include identification of routes where many people currently drive short distances, ways to encourage access to parks, or prioritization of new developments to reduce long-distance travel.

At the city level a web map would be sufficient.

A the national level a mapping application, e.g., with shiny, would probably be needed.

Worldwide, a database to serve the data would likely be needed. Then various front-ends could plug in to this.
```

E9. Update the code in `coffeeApp/app.R` so that instead of centering on Brazil the user can select which country to focus on:

- Using `textInput()`
- Using `selectInput()`

```{asis}
The answer can be found in the `shinymod` branch of the geocompr repo: https://github.com/Robinlovelace/geocompr/pull/318/files
You create the new widget and then use it to set the center.
Note: the input data must be fed into the map earlier to prevent the polygons disappearing when you change the center this way.
```

E10. Reproduce Figure 9.1 and Figure 9.7 as closely as possible using the **ggplot2** package.

```{r}
library(ggplot2)
ggplot() + 
  geom_sf(data = nz, color = NA) +
  coord_sf(crs = st_crs(nz), datum = NA) +
  theme_void()
ggplot() +
  geom_sf(data = nz, fill = NA) +
  coord_sf(crs = st_crs(nz), datum = NA) +
  theme_void()
ggplot() +
  geom_sf(data = nz) + 
  coord_sf(crs = st_crs(nz), datum = NA) +
  theme_void()
# fig 9.7
ggplot() +
  geom_sf(data = nz, aes(fill = Median_income)) +
  coord_sf(crs = st_crs(nz), datum = NA) +
  scale_fill_distiller(palette = "Blues", direction = 1) + 
  theme_void()
ggplot() + 
  geom_sf(data = nz, aes(fill = Island)) + 
  coord_sf(crs = st_crs(nz), datum = NA) +
  scale_fill_manual(values = c("#CC6677", "#332288")) + 
  theme_void()
```

E11. Join `us_states` and `us_states_df` together and calculate a poverty rate for each state using the new dataset.
Next, construct a continuous area cartogram based on total population. 
Finally, create and compare two maps of the poverty rate: (1) a standard choropleth map and (2) a map using the created cartogram boundaries.
What is the information provided by the first and the second map?
How do they differ from each other?

```{r}
tmap_mode("plot")
library(cartogram)
# prepare the data
us = st_transform(us_states, "EPSG:9311")
us = left_join(us, us_states_df, by = c("NAME" = "state"))
# calculate a poverty rate
us$poverty_rate = us$poverty_level_15 / us$total_pop_15
# create a regular map
ecm1 = tm_shape(us) +
  tm_polygons("poverty_rate", fill.legend = tm_legend(title = "Poverty rate"))
# create a cartogram
us_carto = cartogram_cont(us, "total_pop_15")
ecm2 = tm_shape(us_carto) + 
  tm_polygons("poverty_rate", fill.legend = tm_legend(title = "Poverty rate"))
# combine two maps
tmap_arrange(ecm1, ecm2)
```

E12. Visualize population growth in Africa. 
Next, compare it with the maps of a hexagonal and regular grid created using the **geogrid** package.

```{r}
library(geogrid)

hex_cells = calculate_grid(africa, grid_type = "hexagonal", seed = 25, learning_rate = 0.03)
africa_hex = assign_polygons(africa, hex_cells)

reg_cells = calculate_grid(africa, grid_type = "regular", seed = 25, learning_rate = 0.03)
africa_reg = assign_polygons(africa, reg_cells)

tgg1 = tm_shape(africa) +
  tm_polygons("pop_growth", fill.legend = tm_legend(title = "Population's growth (annual %)"))
tgg2 = tm_shape(africa_hex) + 
  tm_polygons("pop_growth", fill.legend = tm_legend(title = "Population's growth (annual %)"))
tgg3 = tm_shape(africa_reg) + 
  tm_polygons("pop_growth", fill.legend = tm_legend(title = "Population's growth (annual %)"))

tmap_arrange(tgg1, tgg2, tgg3)
```
