```{r 10-ex-e0, message=FALSE}
library(sf)
library(terra)
```

<!-- qgisprocess 1-3 -->
<!-- sagagis 1 -->
<!-- sagagis supercells -->
<!-- explain/mention other segmentation techniques -->
<!-- mention supercells -- exercises?? -->
<!-- https://github.com/joaofgoncalves/SegOptim ?? -->
<!-- rgrass 1 -->
<!-- gdal 1-2 -->
1. Use `gdalinfo` via a system call for a raster\index{raster} file stored on disk of your choice.
What kind of information you can find there?

```{r}
link2GI::linkGDAL()
our_filepath = system.file("raster/elev.tif", package = "spData")
cmd = paste("gdalinfo", our_filepath)
system(cmd)
# Driver, file path, dimensions, CRS, resolution, bounding box, summary statistics
```

1. Use `gdalwarp` to decrease the resolution of your raster file (for example, if the resolution is 0.5, change it into 1). Note: `-tr` and `-r` flags will be used in this exercise.

```{r}
our_filepath = system.file("raster/elev.tif", package = "spData")
cmd2 = paste("gdalwarp", our_filepath, "new_elev.tif", "-tr 1 1", "-r bilinear")
system(cmd2)
```

<!-- postgis 1? -->
1. Query all Californian highways from the PostgreSQL/PostGIS\index{PostGIS} database living in the QGIS\index{QGIS} Cloud introduced in this chapter.

```{r}
library(RPostgreSQL)
conn = dbConnect(drv = PostgreSQL(), 
                 dbname = "rtafdf_zljbqm", host = "db.qgiscloud.com",
                 port = "5432", user = "rtafdf_zljbqm", password = "d3290ead")
query = paste(
  "SELECT *",
  "FROM highways",
  "WHERE state = 'CA';")
ca_highways = read_sf(conn, query = query, geom = "wkb_geometry")
plot(st_geometry(ca_highways))
```

<!-- stac+gdalcubes 1 -->
1. The `ndvi.tif` raster (`system.file("raster/ndvi.tif", package = "spDataLarge")`) contains NDVI calculated for the Mongón study area based on Landsat data from September 22nd, 2000.
Use **rstac**, **gdalcubes**, and **terra** to download Sentinel-2 images for the same area from 
2020-08-01 to 2020-10-31, calculate its NDVI, and then compare it with the results from `ndvi.tif`.

```{r}
library(rstac)
library(gdalcubes)
?spDataLarge::ndvi.tif
ndvi1 = rast(system.file("raster/ndvi.tif", package = "spDataLarge"))
bbox1 = as.numeric(st_bbox(project(ndvi1, "EPSG:4326")))

# get data
s = stac("https://earth-search.aws.element84.com/v0")
items = s |>
  stac_search(collections = "sentinel-s2-l2a-cogs",
              bbox = bbox1, 
              datetime = "2020-08-01/2020-10-31") |>
  post_request() |> items_fetch()
collection = stac_image_collection(items$features, 
                  property_filter = function(x) {x[["eo:cloud_cover"]] < 10})
v = cube_view(srs = "EPSG:32717", extent = collection,
              dx = xres(ndvi1), dy = yres(ndvi1),
              dt = "P1D")

# calculate ndvi
ndvi2 = raster_cube(collection, v) |>
  select_bands(c("B04", "B08")) |>
  apply_pixel("(B08-B04)/(B08+B04)", "NDVI")

# write results to file
gdalcubes_options(parallel = 2)
gdalcubes::write_tif(ndvi2, dir = ".", prefix = "ndvi2")

# unify two datasets
ndvi2 = rast("ndvi22020-10-10.tif")
plot(ndvi2)
ndvi2 = resample(ndvi2, ndvi1, method = "bilinear")
plot(ndvi2)

# vizualize the final results
ndvi_all = c(ndvi1, ndvi2)
names(ndvi_all) = c("y2000", "y2020")
library(tmap)
tm_shape(ndvi_all) +
  tm_raster(style = "cont")
```


<!-- 1. Create two overlapping polygons (`poly_1` and `poly_2`) with the help of the **sf**-package (see Chapter \@ref(spatial-class)).  -->

<!-- 1. Union `poly_1` and `poly_2` using `st_union()` and `qgis:union`. -->
<!-- What is the difference between the two union operations\index{vector!union}?  -->
<!-- How can we use the **sf**\index{sf} package to obtain the same result as QGIS\index{QGIS}? -->

<!-- 1. Calculate the intersection\index{vector!intersection} of `poly_1` and `poly_2` using: -->

<!--     - **RQGIS**, **RSAGA** and **rgrass** -->
<!--     - **sf** -->

<!-- 1. Attach `data(dem, package = "spDataLarge")` and `data(random_points, package = "spDataLarge")`. -->
<!-- Select randomly a point from `random_points` and find all `dem` pixels that can be seen from this point (hint: viewshed\index{viewshed}). -->
<!-- Visualize your result. -->
<!-- For example, plot a hillshade\index{hillshade}, and on top of it the digital elevation model\index{digital elevation model}, your viewshed\index{viewshed} output and the point. -->
<!-- Additionally, give `mapview` a try. -->

<!-- 1. Compute catchment area\index{catchment area} and catchment slope of `data("dem", package = "spDataLarge")` using **RSAGA** (see Section \@ref(saga)). -->


