```{r 10-ex-e0, message=FALSE}
library(sf)
library(terra)
```

<!-- qgisprocess 1-3 -->
<!-- sagagis 1 -->
1. Compute catchment area\index{catchment area} and catchment slope of `system.file("raster/dem.tif", package = "spDataLarge")` using **Rsagacmd**.

```{r}
library(Rsagacmd)
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
saga = saga_gis(raster_backend = "terra", vector_backend = "sf")
swi = saga$ta_hydrology$saga_wetness_index
tidy(swi)
swi_results = swi(dem, area_type = 0, slope_type = 1)
swi_results_all = rast(swi_results)
plot(swi_results_all[["area"]])
plot(swi_results_all[["slope"]])
```

1. Continue working on the `ndvi_segments` object created in the SAGA GIS section.
Extract average NDVI values from the `ndvi` raster and group them into six clusters using `kmeans()`. 
Visualize the results.

```{r}
library(Rsagacmd)
saga = saga_gis(raster_backend = "terra", vector_backend = "sf")
ndvi = rast(system.file("raster/ndvi.tif", package = "spDataLarge"))
sg = saga$imagery_segmentation$seed_generation

ndvi_seeds = sg(ndvi, band_width = 2)
plot(ndvi_seeds$seed_grid)

srg = saga$imagery_segmentation$seeded_region_growing
ndvi_srg = srg(ndvi_seeds$seed_grid, ndvi, method = 1)
plot(ndvi_srg$segments)

ndvi_segments = as.polygons(ndvi_srg$segments) |> 
  st_as_sf()

# extract values
ndvi_segments_vals = extract(ndvi, ndvi_segments, fun = "mean")
ndvi_segments = cbind(ndvi_segments, ndvi_segments_vals)

# k-means
ks = kmeans(ndvi_segments[["ndvi"]], centers = 6)
ndvi_segments$k = ks$cluster

# merge polygons
library(dplyr)
ndvi_segments2 = ndvi_segments |> 
  group_by(k) |> 
  summarise()

# visualize results
library(tmap)
tm1 = tm_shape(ndvi) +
  tm_raster(style = "cont", palette = "PRGn", title = "NDVI", n = 7) + 
  tm_shape(ndvi_segments2) +
  tm_borders(col = "red") +
  tm_layout(legend.outside = TRUE)

tm2 = tm_shape(ndvi_segments2) +
  tm_polygons(col = "k", style = "cat", palette = "Set1") +
  tm_layout(legend.outside = TRUE)

tmap_arrange(tm1, tm2)
```

<!-- rgrass 1 -->
1. Attach `data(random_points, package = "spDataLarge")` and read `system.file("raster/dem.tif", package = "spDataLarge")` into R.
Select a point randomly from `random_points` and find all `dem` pixels that can be seen from this point (hint: viewshed\index{viewshed} can be calculated using GRASS GIS).
Visualize your result.
For example, plot a hillshade\index{hillshade}, the digital elevation model\index{digital elevation model}, your viewshed\index{viewshed} output, and the point.
Additionally, give `mapview` a try.

```{r}
library(rgrass)
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
data(random_points, package = "spDataLarge")
random_point = random_points[sample(1:nrow(random_points), 1), ]

link2GI::linkGRASS(dem)
write_RAST(dem, vname = "dem")

execGRASS("r.viewshed",
          input = "dem", 
          coordinates = sf::st_coordinates(random_point),
          output = "view",
          flags = "overwrite")
out = read_RAST("view")

# simple viz
plot(out)

# hillshade viz
hs = shade(slope = terrain(dem, "slope", unit = "radians"), 
           aspect = terrain(dem, "aspect", unit = "radians"))

library(tmap)
tm_shape(hs) +
	tm_raster(palette = gray(0:100 / 100), n = 100, legend.show = FALSE) +
	tm_shape(dem) +
	tm_raster(alpha = 0.6, palette = hcl.colors(25, "Geyser"), legend.show = FALSE) +
  tm_shape(out) +
  tm_raster(style = "cont", legend.show = FALSE) +
	tm_shape(random_point) +
	tm_symbols(col = "black") +
	tm_layout(frame = FALSE)

# mapview viz
library(mapview)
mapview(out, col = "white", map.type = "Esri.WorldImagery") +
  mapview(point)
```


<!-- gdal 1-2 -->
1. Use `gdalinfo` via a system call for a raster\index{raster} file stored on disk of your choice.
What kind of information you can find there?

```{r}
link2GI::linkGDAL()
our_filepath = system.file("raster/elev.tif", package = "spData")
cmd = paste("gdalinfo", our_filepath)
system(cmd)
# Driver, file path, dimensions, CRS, resolution, bounding box, summary statistics
```

1. Use `gdalwarp` to decrease the resolution of your raster file (for example, if the resolution is 0.5, change it into 1). Note: `-tr` and `-r` flags will be used in this exercise.

```{r}
our_filepath = system.file("raster/elev.tif", package = "spData")
cmd2 = paste("gdalwarp", our_filepath, "new_elev.tif", "-tr 1 1", "-r bilinear")
system(cmd2)
```

<!-- postgis 1? -->
1. Query all Californian highways from the PostgreSQL/PostGIS\index{PostGIS} database living in the QGIS\index{QGIS} Cloud introduced in this chapter.

```{r}
library(RPostgreSQL)
conn = dbConnect(drv = PostgreSQL(), 
                 dbname = "rtafdf_zljbqm", host = "db.qgiscloud.com",
                 port = "5432", user = "rtafdf_zljbqm", password = "d3290ead")
query = paste(
  "SELECT *",
  "FROM highways",
  "WHERE state = 'CA';")
ca_highways = read_sf(conn, query = query, geom = "wkb_geometry")
plot(st_geometry(ca_highways))
```

<!-- stac+gdalcubes 1 -->
1. The `ndvi.tif` raster (`system.file("raster/ndvi.tif", package = "spDataLarge")`) contains NDVI calculated for the Mongón study area based on Landsat data from September 22nd, 2000.
Use **rstac**, **gdalcubes**, and **terra** to download Sentinel-2 images for the same area from 
2020-08-01 to 2020-10-31, calculate its NDVI, and then compare it with the results from `ndvi.tif`.

```{r}
library(rstac)
library(gdalcubes)
?spDataLarge::ndvi.tif
ndvi1 = rast(system.file("raster/ndvi.tif", package = "spDataLarge"))
bbox1 = as.numeric(st_bbox(project(ndvi1, "EPSG:4326")))

# get data
s = stac("https://earth-search.aws.element84.com/v0")
items = s |>
  stac_search(collections = "sentinel-s2-l2a-cogs",
              bbox = bbox1, 
              datetime = "2020-08-01/2020-10-31") |>
  post_request() |> items_fetch()
collection = stac_image_collection(items$features, 
                  property_filter = function(x) {x[["eo:cloud_cover"]] < 10})
v = cube_view(srs = "EPSG:32717", extent = collection,
              dx = xres(ndvi1), dy = yres(ndvi1),
              dt = "P1D")

# calculate ndvi
ndvi2 = raster_cube(collection, v) |>
  select_bands(c("B04", "B08")) |>
  apply_pixel("(B08-B04)/(B08+B04)", "NDVI")

# write results to file
gdalcubes_options(parallel = 2)
gdalcubes::write_tif(ndvi2, dir = ".", prefix = "ndvi2")

# unify two datasets
ndvi2 = rast("ndvi22020-10-10.tif")
plot(ndvi2)
ndvi2 = resample(ndvi2, ndvi1, method = "bilinear")
plot(ndvi2)

# vizualize the final results
ndvi_all = c(ndvi1, ndvi2)
names(ndvi_all) = c("y2000", "y2020")
library(tmap)
tm_shape(ndvi_all) +
  tm_raster(style = "cont")
```

