The solutions assume the following packages are attached (other packages will be attached when needed):

```{r 12-ex-e0, message=FALSE, warning=FALSE}
library(dplyr)
# library(kernlab)
library(mlr3)
library(mlr3learners)
library(mlr3extralearners)
library(mlr3spatiotempcv)
library(mlr3tuning)
library(qgisprocess)
library(raster)
# library(rlang)
library(sf)
library(tmap)
```

E1. Compute the following terrain attributes from the `ta$elev` datasets loaded with `data("ta", package = "spDataLarge")` with the help of R-GIS bridges (see this [Chapter](https://geocompr.robinlovelace.net/gis.html#gis)):
    - Slope
    - Plan curvature
    - Profile curvature
    - Catchment area
    
```{r, eval=FALSE}
# attach data
data("ta", package = "spDataLarge")
dem = ta$elev

algs = qgisprocess::qgis_algorithms()
dplyr::filter(algs, grepl("curvature", algorithm))
alg = "saga:slopeaspectcurvature"
qgisprocess::qgis_show_help(alg)
qgisprocess::qgis_arguments(alg)
# terrain attributes (ta)
out_nms = paste0(tempdir(), "/", c("aspect", "slope", "cplan", "cprof"),
                 ".sdat")
args = rlang::set_names(out_nms, c("ASPECT", "SLOPE", "C_PLAN", "C_PROF"))
out = qgis_run_algorithm(alg, ELEVATION = dem, METHOD = 6, 
                         UNIT_SLOPE = "[1] degree",
                         UNIT_ASPECT = "[1] degree",
                         !!!args,
                         .quiet = TRUE
                         )
# use brick because then the layers will be in memory and not on disk
ta = raster::brick(lapply(out[names(args)], \(x) x[1]))
# hillshade (needs radians)
# hs = hillShade(ta$SLOPE * pi / 180, ta$ASPECT * pi / 180, 40, 270)
# plot(hs, col = gray(seq(0, 1, length.out = 100)), legend = FALSE)
# plot(dem, add = TRUE, alpha = 0.6)
# dplyr::filter(lsl, lslpts == TRUE) |> 
#   sf::st_as_sf(coords = c("x", "y"), crs = 32717) |>
#   plot(add = TRUE, pch = 16)

ta = raster::dropLayer(ta, "ASPECT")
names(ta) = c("slope", "cplan", "cprof")
# catchment area
dplyr::filter(algs, grepl("[Cc]atchment", algorithm))
alg = "saga:catchmentarea"
qgis_show_help(alg)
qgis_arguments(alg)
carea = qgis_run_algorithm(alg,
                           ELEVATION = dem, 
                           METHOD = 4, 
                           FLOW = file.path(tempdir(), "carea.sdat"))
# transform carea
carea = raster(carea$FLOW[1])
log10_carea = log10(carea)
names(log10_carea) = "log10_carea"
# add log_carea and dem to the terrain attributes
ta = addLayer(x = ta, dem, log10_carea)
```

E2. Extract the values from the corresponding output rasters to the `lsl` data frame (`data("lsl", package = "spDataLarge"`) by adding new variables called `slope`, `cplan`, `cprof`, `elev` and `log_carea` (see this [section](https://geocompr.robinlovelace.net/spatial-cv.html#case-landslide) for details).

```{r, eval=FALSE}
# attach terrain attribute raster stack (in case you have skipped the previous
# exercise)
data("ta", "lsl", package = "spDataLarge")
lsl = dplyr::select(lsl, x, y, lslpts)
# extract values to points, i.e., create predictors
lsl[, names(ta)] = raster::extract(ta, lsl[, c("x", "y")])
```

E3. Use the derived terrain attribute rasters in combination with a GLM to make a spatial prediction map similar to that shown in this [Figure](https://geocompr.robinlovelace.net/spatial-cv.html#fig:lsl-susc).
Running `data("study_mask", package = "spDataLarge")` attaches a mask of the study area.

```{r, eval=FALSE}
# attach data (in case you have skipped exercises 1) and 2)
# landslide points with terrain attributes and terrain attribute raster stack
data("lsl", "ta", "study_mask", package = "spDataLarge")

# fit the model
fit = glm(lslpts ~ slope + cplan + cprof + elev + log10_carea, 
          data = lsl, family = binomial())

# make the prediction
pred = raster::predict(object = ta, model = fit, type = "response")

# make the map
lsl_sf = sf::st_as_sf(lsl, coords = c("x", "y"), crs = 32717)
hs = hillShade(ta$slope * pi / 180, terrain(ta$elev, opt = "aspect"))
rect = tmaptools::bb_poly(hs)
bbx = tmaptools::bb(hs, xlim = c(-0.02, 1), ylim = c(-0.02, 1), relative = TRUE)
tm_shape(hs, bbox = bbx) +
  tm_grid(col = "black", n.x = 1, n.y = 1, labels.inside.frame = FALSE,
          labels.rot = c(0, 90)) +
  tm_raster(palette = "white", legend.show = FALSE) +
  # hillshade
  tm_shape(mask(hs, study_area), bbox = bbx) +
	tm_raster(palette = gray(0:100 / 100), n = 100, legend.show = FALSE) +
	# prediction raster
  tm_shape(mask(pred, study_area)) +
	tm_raster(alpha = 0.5, palette = RColorBrewer::brewer.pal(name = "Reds", 6),
	          auto.palette.mapping = FALSE, legend.show = TRUE,
	          title = "Susceptibility\nprobability") +
	# rectangle and outer margins
  qtm(rect, fill = NULL) +
	tm_layout(outer.margins = c(0.04, 0.04, 0.02, 0.02), frame = FALSE,
	          legend.position = c("left", "bottom"),
	          legend.title.size = 0.9)
```

E4. Compute a 100-repeated 5-fold non-spatial cross-validation and spatial CV based on the GLM learner and compare the AUROC values from both resampling strategies with the help of boxplots (see this [Figure](https://geocompr.robinlovelace.net/spatial-cv.html#fig:boxplot-cv).
Hint: You need to specify a non-spatial resampling strategy.

```{r, eval=FALSE}
# attach data (in case you have skipped exercises 1) and 2)
data("lsl", package = "spDataLarge")  # landslide points with terrain attributes

# create task
task = TaskClassifST$new(
  id = "lsl_ecuador",
  backend = mlr3::as_data_backend(lsl), target = "lslpts", positive = "TRUE",
  extra_args = list(
    coordinate_names = c("x", "y"),
    coords_as_features = FALSE,
    crs = 32717)
)

# construct learners (for all subsequent exercises)
# GLM
lrn_glm = lrn("classif.log_reg", predict_type = "prob")

# SVM
# construct SVM learner (using ksvm function from the kernlab package)
lrn_ksvm = lrn("classif.ksvm", predict_type = "prob", kernel = "rbfdot",
               type = "C-svc")
# specify nested resampling and adjust learner accordingly
# five spatially disjoint partitions
tune_level = rsmp("spcv_coords", folds = 5)
# use 50 randomly selected hyperparameters
terminator = trm("evals", n_evals = 50)
tuner = tnr("random_search")
# define the outer limits of the randomly selected hyperparameters
ps = ps(
  C = p_dbl(lower = -12, upper = 15, trafo = function(x) 2^x),
  sigma = p_dbl(lower = -15, upper = 6, trafo = function(x) 2^x)
)
at_ksvm = AutoTuner$new(
  learner = lrn_ksvm,
  resampling = tune_level,
  measure = msr("classif.auc"),
  search_space = ps,
  terminator = terminator,
  tuner = tuner
)

# QDA
lrn_qda = lrn("classif.qda", predict_type = "prob")

# SVM without tuning hyperparameters
vals = lrn_ksvm$param_set$values
lrn_ksvm_notune = lrn_ksvm$clone()
lrn_ksvm_notune$param_set$values = c(vals, C = 1, sigma = 1)

# define resampling strategies
# specify the reampling method, i.e. spatial CV with 100 repetitions and 5 folds
# -> in each repetition dataset will be splitted into five folds
# method: repeated_spcv_coords -> spatial partioning
rsmp_sp = rsmp("repeated_spcv_coords", folds = 5, repeats = 100)
# method: repeated_cv -> non-spatial partitioning
rsmp_nsp = rsmp("repeated_cv", folds = 5, repeats = 100)

# (spatial) cross-validataion
#****************************
# create your design
grid = benchmark_grid(tasks = task, 
                      learners = list(lrn_glm, at_ksvm, lrn_qda, 
                                      lrn_ksvm_notune),
                      resamplings = list(rsmp_sp, rsmp_nsp))
# execute the cross-validation
library(future)
# execute the outer loop sequentially and parallelize the inner loop
future::plan(list("sequential", "multisession"), 
             workers = floor(availableCores() / 2))
set.seed(021522)
bmr = benchmark(grid, store_backends = FALSE)
# stop parallelization
future:::ClusterRegistry("stop")
# save your result, e.g. to 
# saveRDS(bmr, file = "extdata/12-bmr.rds")

# plot your result
autoplot(bmr, measure = msr("classif.auc"))
```

E5. Model landslide susceptibility using a quadratic discriminant analysis (QDA).
Assess the predictive performance of the QDA. 
What is the a difference between the spatially cross-validated mean AUROC value of the QDA and the GLM?

```{r, eval=FALSE}
# attach data (in case you have skipped exercise 4)
bmr = readRDS("extdata/12-bmr.rds")

# plot your result
autoplot(bmr, measure = msr("classif.auc"))
# QDA has higher AUROC values on average than GLM which indicates moderately
# non-linear boundaries
```

E6. Run the SVM without tuning the hyperparameters.
Use the `rbfdot` kernel with $\sigma$ = 1 and *C* = 1. 
Leaving the hyperparameters unspecified in **kernlab**'s `ksvm()` would otherwise initialize an automatic non-spatial hyperparameter tuning.

```{r, eval=FALSE}
# attach data (in case you have skipped exercise 4)
bmr = readRDS("extdata/12-bmr.rds")
# plot your result
autoplot(bmr, measure = msr("classif.auc"))
```
