# Location analysis

## Prerequisites

```{r} 
library(sp)
library(raster)
library(osmdata)
library(tidyverse)
```

## Introduction and use case
An ecological amplitude specifies a species' needs and optimum along different gradients.
For example, an icebear prefers northern latitudinal regions where the ice is persisent throughout the summer, where there is enough food (seals and sea lions) and where it can find shelter for its offspring.
The farther away an icebear is forced to live from these favorable conditions the more unlikely is its survival.
Similarily, a market product can have an economic niche.
There are locations of low and high turnovers.
And it is one task of geomarketing to identify regions where business is profitable.
Typical geomarketing questions include:
- Where are my clients, my target groups, my competitors?
- Can many customers easily reach and find my stores?
- Do my stores over-/underexploit the market potential?
- How big is my market share and turnovers, and are there spatial differences?

Suppose we would like to set up bike shops in all metropolitan areas of Germany.
We know that the majority of our clients are between 20-40 years old. 
Besides, our clientele is predominantly male, and lives alone or with just one person (single households, not families).

<!-- just internal structure and brain storming -->
Procedure:
- download German census data and convert it into a raster (resolution: 1 km^2^)
- find metropolitan areas, i.e. aggregate to a resolution of 10 sqkm (define a threshold for metropolitan areas, and point out that there are still more advanced ways to do so)
- download OSM, find bike shops + other interesting POIs.
Create a POI raster (= attraction raster)
- restrict to metropolitan areas
- weights:
  - reclassify age raster: 16-40 weight = 2; all other weight = 1
  - reclassify gender raster:
  - reclassify household raster
	- POIs: reclassify into 5 classes, the more POIs, the better
	- finally map algebra: age + gender + POI raster will show the most favorable locations
- Exercise: download the 100 m census inhabitant data and convert it into a raster (will be used in the chapter continuing location analysis)
  
In an advanced chapter we could continue the analysis as follows:
 - restrict our analysis to Nuremberg (but point out that we could do the whole analysis simultaneously for all metropolitan areas in Germany)
 - choose optimal location based on number of inhabitants -> you should reach as much people within 15 minutes as possible (routing, catchment area)
 - the farther away we get from a shop, the more unlikely people will go to our shop (-> decay distance)
 - however, we have to take into account that there are already competitors (huff model, attraction) 

## Create input rasters

The German government provides a csv file containing the census data for Germany.
You can either download a 1 km or a 100 m resolution.
Here, we download the 1 km data, unzip it and read it into R.

```{r, eval = FALSE}
url = paste0("https://www.zensus2011.de/SharedDocs/Downloads/DE/", 
             "Pressemitteilung/DemografischeGrunddaten/csv_Zensusatlas_", 
             "klassierte_Werte_1km_Gitter.zip?__blob=publicationFile&v=8")
download.file(url = url, destfile = file.path(tempdir(), "census.zip"),
              method = "auto", mode = "wb")
# list the file names
nms = unzip(file.path(tempdir(), "census.zip"), list = TRUE)
# unzip only the csv file
base_name = grep(".csv$", nms$Name, value = TRUE)
unzip(file.path(tempdir(), "census.zip"), files = base_name, exdir = tempdir())
# read in the csv file
input = readr::read_csv2(file.path(tempdir(), base_name))
```

Next we select the variables we need: x and y coordinates, number of inhabitants, mean age, portion of women and the household size. 
Additionally, we set the values -1 and -9 to `NA` since these values are unknown.

```{r}
# only select the variables we need
# inh = inhabitants, hh_size = household size
input = dplyr::select(input, x = x_mp_1km, y = y_mp_1km, inh = Einwohner,
                      mean_age = Alter_D, women = Frauen_A, 
                      hh_size = HHGroesse_D)
# set -1 and -9 to NA
input = mutate_all(input, funs(ifelse(. %in% c(-1, -9), NA, .)))
```

After the data preprocessing, we convert the table into a raster stack making use of the x- and y-coordinates.

```{r}
# convert table into a raster (x and y are cell midpoints)
coordinates(input) =~ x + y
# use the correct projection
proj4string(input) = CRS("+init=epsg:3035")
gridded(input) = TRUE
# convert into a raster stack
input = stack(input)
# plot(input)
```


## Define metroplitan areas
Reclassify inhabitant data.

```{r}
rcl = matrix(c(1, 1, 125, 2, 2, 375, 3, 3, 1250, 4, 4, 3000, 5, 5, 6000,
               6, 6, 8000), ncol = 3, byrow = TRUE)
inh = reclassify(input$inh, rcl = rcl, right = NA)
```

Aggregate it to a resolution of 20000 sqkm.

```{r}
inh_agg = aggregate(inh, fact = 20000 / res(inh)[1], fun = sum)
plot(inh_agg > 500000)
polys = rasterToPolygons(inh_agg[inh_agg > 500000, drop = FALSE])
polys = st_as_sf(polys)
```

Plotting our polygons reveals eight metropolitan regions.
Each region consists of one ore more polygons (pixels).
It would be nice if we could join all polygons belonging to one region.
One approach is to dissolve all polygons.

```{r}
polys = summarize(polys, pop = sum(layer))
```

This returns one multipolygon feature with its elements corresponding to the metropolitan regions. 
To extract these polygons from the multipolygon, we can use `st_cast()`.

```{r}
metros = st_cast(polys, "POLYGON")
``` 

The warning message tells us that the attributes are repeated for all sub-geometries.
In our case, this means that each region receives the total population of all metropolitan areas combined.
Clearly this is wrong but here we can safely ignore this since we are merely interested in the geometry.

<!-- maybe a good if advanced exercise
This requires finding the nearest neighbors (`st_intersects()`), and some additional processing.
Do not worry too much about the following code.
There is probably a better way to do it. 
Nevertheless, it finds all pixels belonging to one region in a generic way.
We use this information to assign each polygon (pixel) to a region.
Subsequently, we can use the region information to dissolve the pixels into region polygons.

```{r}
# dissolve on spatial neighborhood
nbs = st_intersects(polys, polys)
# nbs = over(polys, polys, returnList = TRUE)

fun = function(x, y) {
  tmp = lapply(y, function(i) {
  if (any(x %in% i)) {
   union(x, i)
  } else {
   x
    }
  })
  Reduce(union, tmp)
}
# call function recursively
fun_2 = function(x, y) {
  out = fun(x, y)
  while (length(out) < length(fun(out, y))) {
    out = fun(out, y)
  }
  out
}

cluster = map(nbs, ~ fun_2(., nbs) %>% sort)
# just keep unique clusters
cluster = cluster[!duplicated(cluster)]
# assign the cluster classes to each pixel
for (i in seq_along(cluster)) {
  polys[cluster[[i]], "region_id"] = i
}
# dissolve pixels based on the the region id
polys = group_by(polys, region_id) %>%
  summarize(pop = sum(layer, na.rm = TRUE))
# polys_2 = aggregate(polys, list(polys$region_id), sum)
plot(polys[, "region_id"])

# Another approach, can be also be part of an excercise

coords = st_coordinates(polys_3) %>% 
  as.data.frame
ls = split(coords, f = coords$L2)
ls = lapply(ls, function(x) {
  dplyr::select(x, X, Y) %>%
    as.matrix %>%
    list %>%
    st_polygon
})
metros = do.call(st_sfc, ls)
metros = st_set_crs(metros, 3035)
metros = st_sf(data.frame(region_id = 1:9), geometry = metros)
st_intersects(metros, metros)
plot(metros[-5,])
st_centroid(metros) %>%
  st_coordinates
```
-->

This leaves us with a tiny problem.
Visual inspection reveals eight metropolitan areas whereas the dissolving comes up with nine.
This is because one polygon just touches the corner of another polygon (western Germany, Cologne/Düsseldorf area).
One could assign it to the neighboring region using another dissolving procedure, however, we leave this as an exercise to the reader, and here simply delete the offending polygon.

```{r}
# find out about the offending polygon
int = st_intersects(metros, metros)
# polygons 5 and 9 share one border, delete polygon number 5
metros = metros[-5, ]
```

Now that we have defined the metropolitan areas where to place our bike shops, we additionally want to give them a name.
A reverse geocoding approach can help with that.
Given a coordinate, reverse gecoding finds the corresponding address.
Consequently, we need to extract the centroid coordinate of each metropolitan area.
The resulting coordinates can then serve as an input for a reverse geocoding API.
Here, we make use of the one provided by Google accessed through the **ggmap** package.
Note that Google lets you use its services on a free basis for a maximum of 2500 queries a day.
Since `ggmap::revgeocode()` only accepts lat/lon coordinates, we need to first reproject the metropolitan polygons.

```{r}
# reverse geocoding to find out about the metropolitan names
metros_wgs = st_transform(metros, 4326)
coords = st_centroid(metros_wgs) %>%
  st_coordinates %>%
  round(., 4)
metro_ls = lapply(1:nrow(coords), function(i) {
  # give the server a bit time
  Sys.sleep(sample(seq(1.2, 3.4, 0.1), 1))
  add = ggmap::revgeocode(coords[i, ], output = "more")
  x = 3
  while (is.na(add$address) & x > 0) {
    add = ggmap::revgeocode(coords[i, ], output = "more")
    # just try three times
    x = x - 1
  }
  add
})
metro_names = bind_rows(metro_ls) %>%
  select(locality, administrative_area_level_2)
# replace Velbert
metro_names = 
  mutate(metro_names, 
         locality = ifelse(locality == "Velbert",
                           administrative_area_level_2, locality)) %>%
  select(locality) %>%
  pull
```

Munich, Stuttgart, Nuremberg, Frankfurt, Cologne/Düsseldorf, Hamburg, Leipzig, Berlin
(missing: Dresden, Bielefeld, Bremen, Hannover, but ok)

## Points of interests

OpenStreetMap is definition (VGI).
In the meantime the Vollständigkeit and quality is often comparable to that of commercial data providers (TomTom, Here).
The **osmdata** package provides a fantastic and easy-to-use interface to download OSM data.
Here, we will download all shops for each metropolitan region.
`opq()` defines the bounding box for the area for which data should be retrieved.
You can provide the bounding box yourself or indicate the name of a location.
We use the name of the metropolitan regions to define the bounding boxes. 
Furthermore, we indicate that we only would like download `shop` features (see for a full list).
`osmdata_sf()` returns a list with several spatial objects (points, lines, polygons, etc.).
Here, we will only keep the point objects.

To learn more about how to use the **osmdata**-package, please refer to [https://ropensci.github.io/osmdata/](https://ropensci.github.io/osmdata/).

```{r, eval = FALSE}
library(osmdata)
shops = lapply(metro_names, function(x) {
  # give the server a bit time
  Sys.sleep(sample(seq(5, 10, 0.1), 1))
  q0 = opq(x)
  q1 = add_osm_feature(q0, key = "shop")
  points = osmdata_sf(q1)
  # request the same data again if nothing has been downloaded
  iter = 3
  while (nrow(points$osm_points) == 0 & iter > 0) {
    points = osmdata_sf(q1)
    iter = iter - 1
  }
  points = points$osm_points
  points = st_set_crs(points, 4326)
  })
```

Make sure that each list elements has at least a few features:

```{r}
if (any(sapply(shops, nrow) == 0)) {
  message("There is still a metropolitan area without any features. Please fix!")
}
```

If so, we select only the shop column, and finally bind all shops into one large sf-object.

```{r}
# select only the shop column and rbind all list elements
shops = lapply(shops, select, shop)
shops = Reduce(rbind, shops)
```



```{r}
shops$count = 1
shops = st_transform(shops, proj4string(input$inh))
shops = as(shops, "Spatial")
# create poi raster
poi = input$inh
poi[] = NA
poi = rasterize(x = shops, y = poi, field = "count", fun = "sum")
plot(poi, xlim = c(4380000, 4420000), ylim = c(2920000, 2940000))
```

## Finding suitable locations



Reference (Huff model): https://journal.r-project.org/archive/2017/RJ-2017-020/RJ-2017-020.pdf

## Exercises

1. In the text we have deleted one polygon of the `metros` object (polygon number 5) since it only touches the border of another polygon.
Recreate the `metros` object and instead of deleting polygon number 5, make it part of the Cologne/Düsseldorf metropolitan region (hint: create a column named region_id, add polygon number 5 to the Cologne/Düsseldorf area and dissolve).
1. Suppose our bike shop predominantly sold E-bikes to older people. 
Change the age raster accordingly, repeat the remaining analyses and compare the changes with our original result.
