# Location analysis

Suppose we would like to set up bike shops in all metropolitan areas of Germany.
We know that the majority of our clients are between 20-40 years old. 
Besides, our clientele is predominantly male, and lives alone or with just one person (single households, not families).

Procedure:
- download German census data and convert it into a raster (resolution: 1 km^2^)
- find metropolitan areas, i.e. aggregate to a resolution of 10 sqkm (define a threshold for metropolitan areas, and point out that there are still more advanced ways to do so)
- download OSM, find bike shops + other interesting POIs.
Create a POI raster (= attraction raster)
- restrict to metropolitan areas
- weights:
  - reclassify age raster: 16-40 weight = 2; all other weight = 1
  - reclassify gender raster:
  - reclassify household raster
	- POIs: reclassify into 5 classes, the more POIs, the better
	- finally map algebra: age + gender + POI raster will show the most favorable locations
- Exercise: download the 100 m census inhabitant data and convert it into a raster (will be used in the chapter continuing location analysis)
  
In an advanced chapter we could continue the analysis as follows:
 - restrict our analysis to Nuremberg (but point out that we could do the whole analysis simultaneously for all metropolitan areas in Germany)
 - choose optimal location based on number of inhabitants -> you should reach as much people within 15 minutes as possible (routing, catchment area)
 - the farther away we get from a shop, the more unlikely people will go to our shop (-> decay distance)
 - however, we have to take into account that there are already competitors (huff model, attraction) 

## Prerequisites

```{r} 
library(sp)
library(raster)
library(osmdata)
library(tidyverse)
```

## Data download

Download the census data, unzip it and read it into R.
```{r}
url = paste0("https://www.zensus2011.de/SharedDocs/Downloads/DE/", 
             "Pressemitteilung/DemografischeGrunddaten/csv_Zensusatlas_", 
             "klassierte_Werte_1km_Gitter.zip?__blob=publicationFile&v=8")
download.file(url = url, destfile = file.path(tempdir(), "class.zip"),
              method = "auto", mode = "wb")
# list the file names
nms = unzip(file.path(tempdir(), "class.zip"), exdir = tempdir(), list = TRUE)
# unzip only the csv file
unzip(file.path(tempdir(), "class.zip"), files = nms$Name[2], exdir = tempdir())
# read in the csv file
input = readr::read_csv2(file.path(tempdir(), nms$Name[2]))
```

Next we would like to only select the variables we need. 
Additionally, we would like to set the values -1 and -9 to `NA` since these values are either unknown or secret.

```{r}
# only select the variables we need
# inh = inhabitants, hh_size = household size
input = dplyr::select(input, x = x_mp_1km, y = y_mp_1km, inh = Einwohner,
                      mean_age = Alter_D, women = Frauen_A, 
                      hh_size = HHGroesse_D)
# set -1 and -9 to NA since this corresponds to unknown or secret values
input = mutate_all(input, funs(ifelse(. %in% c(-1, -9), NA, .)))
```

After the data preprocessing, we can convert the table into a raster stack making use of the x- and y-coordinates.

```{r}
# convert table into a raster (x and y are cell midpoints)
coordinates(input) =~ x + y
# use the correct projection
proj4string(input) = CRS("+init=epsg:3035")
gridded(input) = TRUE
# convert into a raster stack
input = stack(input)
# plot(input)
```

Reclassify inhabitant data.

```{r}
rcl = matrix(c(1, 1, 125, 2, 2, 375, 3, 3, 1250, 4, 4, 3000, 5, 5, 6000,
               6, 6, 8000), ncol = 3, byrow = TRUE)
inh = reclassify(input$inh, rcl = rcl, right = NA)
```

Aggregate it to a resolution of 20000 sqkm.

```{r}
inh_agg = aggregate(inh, fact = 20000 / res(inh)[1], fun = sum)
plot(inh_agg > 500000)
```

Munich, Stuttgart, Nuremberg, Frankfurt, Cologne/Düsseldorf, Hamburg, Leipzig, Berlin
(missing: Dresden, Bielefeld, Bremen, Hannover, but ok)


Reference (Huff model): https://journal.r-project.org/archive/2017/RJ-2017-020/RJ-2017-020.pdf