---
title: "sfheaders"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sfheaers-setup}

## Detatch {sf} to remove 'print' methods
## because I want to show the underlying structure 
##
## library(sf) will be called later
pkgload::unload("sf")
```


The design philosophy of `{sfheaders}` is to

1. **build** `{sf}` objects from vectors, matrices and data.frames
  - without depending on the `{sf}` library
2. **expose** all C++ code through header files (hence the name, `sfheaders`)

It's intended that every operation you can do in R, you can also do in C++. I anticipate 99% of users will only need the R functions. But the C++ header files are there if you are building your own package with C++ code and would like to build `{sf}` objects directly in C++.

This package is not directly affiliated with the `{sf}` library, but every effort is made to keep it aligned and create valid `sf` objects.

## Building sf objects in R

The simplest use-case for `{sfheaders}` is best demonstrated with examples of building  `sfg`, `sfc` and `sf` objects.

### Simple Feature Geometry - sfg

Here are examples showing

- a vector converted to `sfg_POINT`
- a matrix converted to `sfg_LINESTRING`
- a data.frame converted to `sfg_POLYGON`

```{r sfheaders-sfg_point} 
v = c(1,1)
sfheaders::sfg_point(obj = v)
```


```{r sfheaders-sfg_linestring}
n = 8
m = matrix(1:n, ncol = 2)

sfheaders::sfg_linestring(obj = m)

```


```{r sfheaders-sfg_polygon}

n = 4

df = data.frame(
  x = 1:n
  , y = n:1
)

sfheaders::sfg_polygon(obj = df)
```

In these examples I have deliberately not used the print methods that come with the `{sf}` library, to show you the underlying structures of the objects. If I now load `{sf}` 

```{r sfheaders-library-sf}
library(sf)
```

and create the `sfg_POLYGON` again, you'll see it's printed to the console as a regular `{sf}` object


```{r sfheaders-sfg_polygon_print}
sfheaders::sfg_polygon(obj = df)
```

### Simple Feature Collections - sfc

Reusing the objects `v`, `m`, and `df` we can also build `sfc` objects

```{r sfheaders-sfc_point}
sfheaders::sfc_point(obj = v)
```

```{r sfheaders-sfc_linestring}
sfheaders::sfc_linestring(obj = m)
```

```{r sfheaders-sfc_polygon}
sfheaders::sfc_polygon(obj = df)
```

### Simple Features - sf

And similarly, `sf` objects

```{r sfheaders-sfc_point}
sfheaders::sf_point(obj = v)
```

```{r sfheaders-sfc_linestring}
sfheaders::sf_linestring(obj = m)
```

```{r sfheaders-sfc_polygon}
sfheaders::sf_polygon(obj = df)
```

In each of these examples you'll notices the CRS (coordinate reference system) is not defined. If you plan on doing any calculations or geometric operations using `{sf}` functions you'll need to set this manually, for example

```{r sfheaders-crs}
sf <- sfheaders::sfg_polygon(obj = df)
sf::st_crs(sf) <- 4326  ## Web Mecarter (long/lat)
```


## Building sf objects in C++

To use the C++ API you need to link to both the `{sfheaders}` and `{geometries}` libraries.

If you're building a package you define these links in the `LinkingTo` section of the `DESCRIPTION` file. For example, this is the `LinkingTo` section of the [`{geojsonsf}`](https://github.com/SymbolixAU/geojsonsf/blob/master/DESCRIPTION) library.

```
LinkingTo: 
    geometries,
    jsonify (>= 1.1.1),
    rapidjsonr (>= 1.2.0),
    Rcpp,
    sfheaders (>= 0.2.2)

```

Now all the C++ headers in `{sfheaders}` will be available in your package, just define them as `#include` statements at the top of your code and away you go.

Here are the same examples as before, showing

- a vector converted to `sfg_POINT`
- a matrix converted to `sfg_LINESTRING`
- a data.frame converted to `sfg_POLYGON`

but written as C++ (through Rcpp)

```c++
#include "sfheaders/sfg/point/sfg_point.hpp"
#include "sfheaders/sfg/linestring/sfg_linestring.hpp"
#include "sfheaders/sfg/polygon/sfg_polygon.hpp"

//' Create Point
//'
//' builds an sfg_point object from a valid SEXP type
//'
//' [[Rcpp::export]]
SEXP create_point(SEXP x) {
  return sfheaders::sfg::sfg_point( x );
}

//' Create Linestring
//' 
//' builds an sfg_linestring object from a vaild SEXP type
//' 
//' [[Rcpp::export]]
SEXP create_linestring(SEXP x) {
  return sfheaders::sfg::sfg_linestring( x );
}

//' Create Polygon
//' 
//' builds an sfg_polygon object from a vaild SEXP type
//' 
//' [[Rcpp::export]]
SEXP create_polygon(SEXP x) {
  return sfheaders::sfg::sfg_polygon( x );
}

```









