---
title: "Geocomputation for active transport planning: a case study of cycling networks"
subtitle: "🌐<br/>  "
author: "Robin Lovelace"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    # libs: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
# download.file("https://raw.githubusercontent.com/Robinlovelace/geocompr/master/refs.bib", "refs.bib")
# download.file("https://raw.githubusercontent.com/Robinlovelace/geocompr/master/packages.bib", "packages.bib")
```

```{r, echo=FALSE}
library(methods)
```

Although it has academic origins, R is now widely used for many real-world applications including finance, epidemiology, store location analysis and conservation.
Indications suggest that the language continues to gain adoption in other areas due to its advantages of cost, reproducibility and flexibility over proprietary products.
This talk will explore the potential uptake of R for transport planning, a domain in which powerful yet accessible scriptable languages have much to offer.
Transport planning relies on data analysis, a range of spatial and temporal data forms, and visualisation, areas that R excels in, especially with add-on packages such as **sf**.
To illustrate the point the talk will describe work commissioned by the UK's Department for Transport to develop a Propensity to Cycle Tool (PCT).
The PCT is now being used around England to help design strategic cycle networks and improve the effectiveness of precious public investment in transport infrastructure.
Based on the experience developing the PCT we will discuss the limitations of R as a tool for transport planning, and the potential of recent developments in packages such as **stplanr** and **dodgr** packages to address them.
The talk will conclude by outlining reasons for transport planning authorities to demand the use of open source software and reproducibility to ensure democratic accountability and the cost-effective use of public funds.
