# Advanced spatial subsetting

This vignette demonstrates some subtleties of spatial subsetting of `sf` objects using base R and **tidyvers** methods.

The starting point is to load necessary packages:

```{r}
library(sf)
library(spData)
library(tidyverse)
```

And subsetting canterbury and `nz_height` features in the region (as per Chapter 4 of the book):

```{r}
canterbury = nz %>% filter(Name == "Canterbury")
canterbury_height = nz_height[canterbury, ]
```

We can see how to get a list representation for subsetting:

```{r}
sel_sgbp = st_intersects(x = nz_height, y = canterbury)
class(sel_sgbp)
sel_logical = lengths(sel_sgbp) > 0
canterbury_height2 = nz_height[sel_logical, ]
```

And use the result with the tidyverse:

````{r}
canterbury_height3 = nz_height %>% filter(sel_logical)
```

We can test whether the subsetted objects are identical as follows:

```{r}
identical(x = canterbury_height, y = canterbury_height2)
identical(x = canterbury_height, y = canterbury_height3)
```

What is different about `canterbury_height3`?
The only difference is that `filter()` changed the row names:

```{r}
row.names(canterbury_height)[1:3]
row.names(canterbury_height3)[1:3]
```

If the row names are re-set, the objects become identical:

```{r}
attr(canterbury_height3, "row.names") = attr(x = canterbury_height, "row.names")
identical(canterbury_height, canterbury_height3)
```

```{block type='rmdnote'}
This discarding of row names is not something that is specific to spatial
data, as illustrated in the code chunk below.
**dplyr** discards row names by design.
For further discussion of this decision, and some controversy, see the (closed) issue [#366](https://github.com/tidyverse/dplyr/issues/366) in the package's issue tracker.
```

In the code below a dense matrix is created.
`sel` is not, as one might imagine, a `logical` `vector` (although it behaves as one as it only has one column) but a `logical` `matrix`:

```{r}
sel = st_intersects(nz_height, canterbury, sparse = FALSE)
class(sel)
typeof(sel)
dim(sel)
```

The dimensions of `sel` (returned by the base R command `dim()`) show one row per feature in the target object (`nz_height`) and a column per feature in the subsetting object (`canterbury`).
The general pattern here is that `sel[i, j]` is `TRUE` if the i^th^ feature in the target object intersects with the j^th^ feature in the subsetting object.
If there is more than one feature in `y` the resulting selection `matrix` must be converted into a `vector` before it is used for subsetting, e.g. with `rowSums(sel_matrix) > 0`.
Another solution is to convert the default sparse matrix (`list`) output from `st_intersects()` to a `logical` vector using the function `lengths()`.
This approach to spatial subsetting, used internally by **sf** (see the source code of `` sf:::`[.sf` ``), is illustrated in the code chunk below:

```{r}
co = filter(nz, grepl("Canter|Otag", Name))
sel_sparse = st_intersects(nz_height, co)
sel_vector = lengths(sel_sparse) > 0
heights_co = nz_height[sel_vector, ]
```

The above code chunk results in an object, `heights_co`, that represents the high points that intersect with either Canterbury *or* Otago region (hence the object name `co`).
It did this in four stages:

1. Subset the regions of `nz` containing "Canter" or "Otago" in their names. This was done using the pattern matching function `grepl()` in combination with the `|` character, which means 'or', resulting in the subsetting object `co`.
2. Create a sparse geometry binary predicate `sgbp` object, a list representing which features of `nz_height` intersect with the regions in `co`.
3. Convert the selection list into a `logical` 'selection vector'. `lengths()`  finds the features in `nz_height` matching *any* features in `co`.
4. Use the result to subset `nz_heights`, creating a new object `heights_co`. 