
```{r}
t_area = function(x) {
  abs(
    x[1, 1] * (x[2, 2] - x[3, 2]) +
    x[2, 1] * (x[3, 2] - x[1, 2]) +
    x[3, 1] * (x[1, 2] - x[2, 2])
  ) / 2
}
```

The function `t_area` generalizes the solution by taking any object `x`, assumed to have the same dimensions as the triangle represented in `T1`.
We can verify it works on the triangle matrix `T1` as follows:

```{r}
t_area(T1)
```

Likewise we can create a function that finds the triangle's centroid:

```{r}
t_centroid = function(x) {
  (x[1, ] + x[2, ] + x[3, ]) / 3
}
t_centroid(T1)
```

<!-- Commented-out because it makes more sense to do it in base R. Introduce decido later (Robin) -->
<!-- With these functions created and tested on the first triangle of the polygon, we can we can apply the solution to many triangles, which will be created with the **decido** package: -->

```{r, echo=FALSE, eval=FALSE}
ind = decido::earcut(poly_mat)
decido::plot_ears(poly_mat, idx = ind)
i = seq(1, length(ind), by = 3)
i_list = purrr::map(i, ~c(.:(.+2), .))
T_all = purrr::map(i_list, ~poly_mat[ind[.], ])
```

The next stage is to create the second triangle and calculate its area and centroid.
This is done in the code chunk below, which binds the 3^rd^ and 4^th^ coordinates onto the 1^st^, and uses the same functions we created above to calculate its area and width:

```{r}
T2 = rbind(O, poly_mat[3:4, ], O)
A2 = t_area(T2)
C2 = t_centroid(T2)
```

From this point it is not a big leap to see how to create the 3^rd^ and final triangle that constitutes the polygon represented by `poly_mat` (see exercises below).
However, an aim of algorithms is often to *generalise* and *automate* the solution.