---
title: "Geocomputation with R - Benchmark"
author: "Robin Lovelace, Jakub Nowosad, Jannes Muenchow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Reading vector data

```{r, message=FALSE}
library(sf)
vector_filepath = system.file("shapes/world.gpkg", package = "spData")
world = st_read(vector_filepath)
```

To demonstrate this, we will use a function to compare `st_read()` with its **sp** equivalent, `rgdal::readOGR()`:

```{r}
bench_read = function(file, n) {
  m = microbenchmark(times = n,
                     rgdal::readOGR(vector_filepath),
                     st_read(vector_filepath)
  )
  mean(m$time[1:n]) / mean(m$time[(n + 1):(n * 2)])
}
```

This function takes as arguments an input file (`file`) and a number of times to run each command (`n`) and returns how many times faster `st_read()` is than `readOGR()`.
Let's run the benchmark for the `world.gpkg` file represented by the object `vector_filepath`:

```{r, results='hide'}
library(microbenchmark)
read_world_gpkg = bench_read(file = vector_filepath, n = 5)
```

```{r}
read_world_gpkg
```

The results demonstrate that **sf** was around `r round(read_world_gpkg)` times faster than **rgdal** at reading-in the world countries vector.
The relative performance of `st_read()` compared with other functions will vary depending on file format and the nature of the data.
To illustrate this point, we performed the same operation on a geojson file and found a greater speed saving:

```{r, results='hide'}
vector_filepath_gj = system.file("shapes/cycle_hire_osm.geojson", package = "spData")
read_lnd_geojson = bench_read(file = vector_filepath_gj, n = 5)
```

```{r}
read_lnd_geojson
```

In this case **sf** was around `r round(read_lnd_geojson)` times faster than **rgdal**.

# Writing vector data

```{r, message=FALSE}
library(rgdal)
world_sp = as(world, "Spatial")
```


```{r, warning=FALSE}
system.time(writeOGR(world_sp, dsn = "world_sp.geojson", layer = "world_sp.geojson", driver = "GeoJSON"))
system.time(writeOGR(world_sp, dsn = ".", layer = "world_sp", driver = "ESRI Shapefile")) 
system.time(writeOGR(world_sp, dsn = "world_sp.gpkg", layer = "world_sp.gpkg", driver = "GPKG"))
system.time(st_write(world, "world_sf.geojson", quiet = TRUE))
system.time(st_write(world, "world_sf.shp", quiet = TRUE)) 
system.time(st_write(world, "world_sf.gpkg", quiet = TRUE))
```

```{r, echo=FALSE, results='hide'}
world_files = list.files(pattern = "world*")
file.remove(world_files)
```