```{r}
library(sf)
library(spData)
library(tidyverse)
```


We will use an `sf` object `north_america` with country codes (`iso_a2`), names and geometries, as well as a `data.frame` object `wb_north_america` containing information about urban population and unemployment for three countries.
Note that `north_america` contains data about Canada, Greenland and the United States but the World Bank dataset (`wb_north_america`) contains information about Canada, Mexico and the United States:

```{r}
north_america = world %>%
  filter(subregion == "Northern America") %>%
  dplyr::select(iso_a2, name_long)
north_america$name_long
```

```{r}
wb_north_america = worldbank_df %>% 
  filter(name %in% c("Canada", "Mexico", "United States")) %>%
  dplyr::select(name, iso_a2, urban_pop, unemploy = unemployment)
```

We will use a left join to combine the two datasets.
Left joins are the most commonly used operation for adding attributes to spatial data, as they return all observations from the left object (`north_america`) and the matched observations from the right object (`wb_north_america`) in new columns.
Rows in the left object without matches in the right (`Greenland` in this case) result in `NA` values.

To join two objects we need to specify a key.
This is a variable (or a set of variables) that uniquely identifies each observation (row). 
The `by` argument of **dplyr**'s join functions lets you identify the key variable. 
In simple cases, a single, unique variable exist in both objects like the `iso_a2` column in our example (you may need to rename columns with identifying information for this to work):

```{r}
left_join1 = north_america %>% 
  left_join(wb_north_america, by = "iso_a2")
```

This has created a spatial dataset with the new variables added.
The utility of this is shown in Figure \@ref(fig:unemploy), which shows the unemployment rate (a World Bank variable) across the countries of North America.

```{r unemploy, echo=FALSE, fig.cap="The unemployment rate (taken from World Bank statistics) in Canada and the United States to illustrate the utility of joining attribute data on to spatial datasets."}
# tmap::qtm(left_join1, "unemploy", fill.breaks = c(6, 6.5, 7), fill.title="Unemployment rate: ",
#           projection = "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
library(tmap)
tm_shape(left_join1) + 
  tm_polygons("unemploy", breaks = c(6, 6.5, 7), title = "Unemployment rate: ") +
  tm_layout(legend.position = c("right", "bottom"))
```

It is also possible to join objects by different variables.
Both of the datasets have variables with names of countries, but they are named differently.
The `north_america` has a `name_long` column and the `wb_north_america` has a `name` column.
In these cases a named vector, such as `c("name_long" = "name")`, can specify the connection:

```{r}
left_join2 = north_america %>% 
  left_join(wb_north_america, by = c("name_long" = "name"))
names(left_join2)
```

Note that the result contains two duplicated variables - `iso_a2.x` and `iso_a2.y` because both `x` and `y` objects have the column `iso_a2`.
This can be solved by specifying all the keys:

```{r}
left_join3 = north_america %>% 
  left_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
```

Joins also work when a data frame is the first argument.
This keeps the geometry column but drops the `sf` class, returning a `data.frame` object.

```{r}
# keeps the geom column, but drops the sf class
left_join4 = wb_north_america %>%
  left_join(north_america, by = c("iso_a2"))
class(left_join4)
```

<!-- On the other hand, it is also possible to remove the geometry column of `left_join4` using base R functions or `dplyr`. -->
<!-- Here, this is this simple because the geometry column is just another `data.frame` column and no longer the sticky geometry column of an `sf` object (see also Chapter \@ref(spatial-class)): -->

<!-- ```{r} -->
<!-- # base R -->
<!-- left_join4_df = subset(left_join4, select = -geom) -->
<!-- # or dplyr -->
<!-- left_join4_df = left_join4 %>% dplyr::select(-geom) -->
<!-- left_join4_df -->
<!-- class(left_join4_df) -->
<!-- ``` -->

In contrast to `left_join()`, `inner_join()` keeps only observations from the left object (`north_america`) where there are matching observations in the right object (`wb_north_america`). 
All columns from the left and right object are still kept:

```{r}
inner_join1 = north_america %>% 
  inner_join(wb_north_america, by = c("iso_a2", "name_long" = "name"))
inner_join1$name_long
```