---
title: "Geocomputation with R - Exercises and Solutions"
author: "Robin Lovelace, Jakub Nowosad, Jannes Muenchow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r packages, message=FALSE, warning=FALSE}
library(sf)
library(tidyverse)
library(spData)
data("world")
data("us_states")
data("us_states_df")
```

# Chapter 1

1. Think about the terms 'GIS', 'GDS' and 'Geocomputation' described above. Which is your favorite and and why?

1. Provide 3 reasons for using a scriptable language such R for geocomputation instead of established GIS programs such as QGIS.

- Reproducibility: a sequence of operations is much easier to save and share when using a scripting language.
- Scalability: scripting languages make it easy to break-up the work and parallelise, making it more scalable.
- Flexibility: the building blocks of the language allows it to interface efficiently with external software, enabling R to perform a wide range of tasks (e.g. animations, online interactive maps).
- Efficiency: the GUI of GIS programs can be slow to use and install.

1. Name two advantages and two disadvantages of using the older **sp** package compared with the new **sf** package.

- `+` Number of packages supported.
- `+` Stability
- `-` Slow performance
- `-` Non standard data format

# Chapter 2

1. What does the summary of the `geometry` column tell us about the `world` dataset, in terms of:

- The geometry type?
- How many countries there are?
- The coordinate reference system (CRS)?

2. Using **sf**'s `plot()` command, create a map of Nigeria in context, building on the code that creates and plots Asia above (see Figure \@ref(fig:nigeria) for an example of what this could look like). 

- Hint: this used the `lwd`, `main` and `col` arguments of `plot()`. 
- Bonus: make the country boundaries a dotted grey line.
- Hint: `border` is an additional argument of `plot()` for **sf** objects.

```{r nigeria, warning=FALSE, echo=FALSE, fig.cap="Map of Nigeria in context illustrating sf's plotting capabilities"}
nigeria = world[world$name_long == "Nigeria", ]
africa = world[world$continent == "Africa", ]
bb_africa = st_bbox(africa)
plot(africa[2], col = "white", lwd = 3, main = "Nigeria in context", border = "lightgrey")
# plot(world, lty = 3, add = TRUE, border = "grey")
plot(world, add = TRUE, border = "grey")
plot(nigeria, col = "yellow", add = TRUE, border = "darkgrey")
ncentre = st_centroid(nigeria)
ncentre_num = st_coordinates(ncentre)
text(x = ncentre_num[1], y = ncentre_num[2], labels = "Nigeria")
```

3. Exercise 3 <!--missing title-->

- What does the `lwd` argument do in the `plot()` code that generates Figure \@ref(fig:africa). 
- Perform the same operations and map making for another continent of your choice.
- Bonus: Download some global geographic data and add attribute variables assigning them to the continents of the world.

# Chapter 3

<!-- Attribute subsetting -->
1. Select only the `NAME` column in `us_states` and create a new object called `us_states_name`. 
What is the class of the new object? <!--why there is a "sf" part? -->
```{r}
us_states_name = us_states %>% select(NAME)
class(us_states_name)
```
2. Select columns which contain information about a total population.
Think about as many ways as possible to do it. 
Hint: try to use helper functions, such as `contains` or `starts_with`.
```{r}
us_states %>% select(total_pop_10, total_pop_15)
```
```{r}
us_states %>% select(starts_with("total_pop"))
```
```{r}
us_states %>% select(contains("total_pop"))
```
3. Find all states that:
- Belongs to the Midwest region
```{r}
us_states %>% filter(total_pop_15 < 750000)
```
- Had a total population in 2015 lower than 750,000 residents
```{r}
us_states %>% filter(total_pop_15 < 750000)
```
- Belongs to the West region, had an area smaller than 250,000 km^2^ and total population in 2015 larger than 5,000,000 residents
```{r}
us_states %>% filter(REGION == "West", AREA < units::set_units(250000, km^2),total_pop_15 > 5000000)
# or
us_states %>% filter(REGION == "West", as.numeric(AREA) < 250000,total_pop_15 > 5000000)
```
- Belongs to the South region, had an area larger than 150,000 km^2^ or total population in 2015 larger than 7,000,000 residents
```{r}
us_states %>% filter(REGION == "South", AREA > units::set_units(150000, km^2), total_pop_15 > 7000000)
# or
us_states %>% filter(REGION == "South", as.numeric(AREA) > 150000, total_pop_15 > 7000000)
```
<!-- 4. Exercises about additional select and filter helper functions (one_of; between, is.na, near, etc.) -->
<!-- Attribute aggregation -->
4. What was the total population in 2015 in the `us_states` database? What was the minimum and maximum total population in 2015?
```{r}
us_states %>% summarize(total_pop = sum(total_pop_15),
                        min_pop = min(total_pop_15),
                        max_pop = max(total_pop_15))
```
5. How many states are in each region?
```{r}
us_states %>%
  group_by(REGION) %>%
  summarize(nr_of_states = n())
```
6. What was the minimum and maximum total population in 2015 in each region? What was the total population in 2015 in each region?
```{r}
us_states %>%
  group_by(REGION) %>%
  summarize(min_pop = min(total_pop_15),
            max_pop = max(total_pop_15),
            tot_pop = sum(total_pop_15))
```
<!-- Attribute joining -->
7. Add variables from `us_states_df` to `us_states` and create a new object called `us_states_stats`. What is the best function to do it? Which variable is the key in the both datasets? What is the class of a new object?
```{r}
us_states_stats = us_states %>%
  left_join(us_states_df, by = c("NAME" = "state"))
class(us_states_stats)
```
8. `us_states_df` has two more variables than `us_states`. How you can find them?
```{r}
us_states_df %>%
  anti_join(us_states, by = c("state" = "NAME"))
```
<!-- Attribute creation -->
9. What was the population density in 2015 in each state? What was the population density in 2010 in each state?
```{r}
us_states2 = us_states %>%
  mutate(pop_dens_15 = total_pop_15/AREA,
         pop_dens_10 = total_pop_10/AREA)
```
10. How much the population density changed between 2010 and 2015 in each state? Calculate the change in percentages.
```{r}
us_states2 %>%
  mutate(pop_dens_diff_10_15 = pop_dens_15 - pop_dens_10,
         pop_dens_diff_10_15p = (pop_dens_diff_10_15/pop_dens_15) * 100)
```
11. Change the columns names in `us_states` to lowercase. Try to use two helper functions - `tolower()` and `colnames()`.
```{r}
us_states %>%
  set_names(tolower(colnames(.)))
```
<!-- Mixed exercises -->
<!-- combination of use of select, mutate, group_by, summarize, etc  -->
12. Using `us_states` and `us_states_df` create a new object called `us_states_sel`. The new object should have only two variables - `median_income_15` and `geometry`. Change the name of the `median_income_15` column to `Income`.
```{r}
us_states %>%
  left_join(us_states_df, by = c("NAME" = "state")) %>%
  select(Income = median_income_15)
```
13. Calculate the change in median income between 2010 and 2015 for each state. What was the minimum, average and maximum median income in 2015 for each region? What is the region with the largest increase of the median income?
```{r}
us_states %>%
  left_join(us_states_df, by = c("NAME" = "state")) %>%
  mutate(income_change = median_income_15 - median_income_10) %>%
  group_by(REGION) %>%
  summarize(min_income_change = min(income_change),
            mean_income_change = mean(income_change),
            max_income_change = max(income_change)) %>%
  filter(mean_income_change == max(mean_income_change)) %>%
  pull(REGION) %>%
  as.character()
```

# Chapter 4

# Chapter 5
