# Further questions about world populations.

```{r}
library(sf)
library(tidyverse)
```


1. Write code that uses the functions `aggregate()` and `st_buffer()` to answers the following question: what proportion of the world's population lives in countries that intersect a circle with a 10 degree radius of the intersection between the equator and the [9^th^ meridian](https://en.wikipedia.org/wiki/9th_meridian_east)?
```{r, echo=FALSE, eval=FALSE}
center9 = st_sf(st_sfc(st_point(c(-9, 0)), crs = 4326))
buff9 = st_buffer(center9, dist = 10)
agg9 = aggregate(world["pop"], buff9, FUN = sum)
agg9$pop / sum(world$pop, na.rm = TRUE) # 1%
```
1. Assuming that people are evenly distributed across countries, estimate the population living *within* the circle created to answer the previous question.
```{r, echo=FALSE, eval=FALSE}
interp9 = st_interpolate_aw(x = world["pop"], to = buff9, extensive = TRUE)
```
1. Warning messages should have been produced during the working to find the answer to the previous questions. What do these warnings mean and how could they be stopped? 
    - Bonus: rewrite code that generated the answer to the previous question using a projected CRS (suggestion: UTM).
1. Advanced challenge: find the point with the highest number of people within a 10 degree radius.